### 2020-03-29 index.tex
Initial commit

\documentclass[12pt,a4paper]{article}
\usepackage{ccicons}
\usepackage{hyperref}

\newcommand{\version}{v0.1}

\begin{document}
\begin{centering}
	\Huge{The Title}
	\par
	\vspace{2ex}

	\normalsize{
		Manuel Alabor\\
		Supervised by The Supervisor\\
		\par
		\vspace{2ex}
		HSR University of Applied Sciences Rapperswil\\
		\par
		\vspace{2ex}
		\today{} (\version)
	}
	\par
	\vspace{2ex}

	\begin{quotation}
		\small{
			\noindent\textbf{Abstract}---\input{./abstract.tex}
		}
		\par
		\vspace{2ex}

		\small{
			\noindent\textbf{Keywords}: tbd.
		}
		\par
		\vspace{4ex}
	\end{quotation}
\end{centering}

\section{Intro}
Some Content \cite{Author:Year:Abbr:1111111}
Test

\section{Intro}
Duis autem vel eum iriure dolor in hendrerit in vulputate velit esse molestie consequat, vel illum dolore eu feugiat nulla facilisis at vero eros et accumsan et iusto odio dignissim qui blandit praesent luptatum zzril delenit augue duis dolore te feugait nulla facilisi. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.

\bibliographystyle{splncs04}
\bibliography{index}

\section*{License}
\ccby\thinspace\thinspace This work is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International License}.
\end{document}

### 2020-04-02 paper.tex
Outline

\documentclass[12pt,a4paper]{article}
\usepackage{ccicons}
\usepackage{hyperref}

\newcommand{\version}{v0.1 Outline}

\begin{document}
\begin{centering}

	\Large{Workshop Paper}
	\par
	\Huge{tbd. ``Reactive Debugging for modern Frontend Web Applications''}
	\par
	\vspace{2ex}

	\normalsize{
		Manuel Alabor\\
		Supervised by Prof. Dr. Markus Stolze\\
		\par
		\vspace{2ex}
		HSR University of Applied Sciences Rapperswil\\
		\par
		\vspace{2ex}
		\today{} (\version)
	}
	\par
	\vspace{2ex}

	\begin{quotation}
		\small{
			\noindent\textbf{Abstract}---\input{./abstract.tex}
		}
		\par
		\vspace{2ex}

		\small{
			\noindent\textbf{Keywords}: tbd.
		}
		\par
		\vspace{4ex}
	\end{quotation}
\end{centering}

\section{Overview}
\begin{itemize}
	\item What is RP?
	\item What is debugging?
	\item Debug concepts in RP
	\item Research so far
	\item Tooling development so far
	\item Research questions:
	\begin{itemize}
		\item RQ1 How do web frontend engineers debug RP code?
		\item RQ2 How can we improve the debugging experience for FE?
		\item RQ3 Does a specific RP debugger for FE improve the debugging experience?
	\end{itemize}
	\item \bf{DEBUGGING EXPERIENCE: How to measure? Use ``time'' instead?}
\end{itemize}

\section{``Current State''}
\begin{itemize}
	\item Closer look to past Research
	\item \dots and the tools which were built based on it
	\item Pick up gaps/deficiencies from past research/tools

\end{itemize}

\section{Interviews}
\begin{itemize}
	\item Answer RQ1
	\item Reference previous research
	\item Present results of own informal interviews
	\begin{itemize}
		\item Validate/invalidate results of past research based on interviews
	\end{itemize}
	\item Lead over to frontend engineering
	\item Compose past research with interview results
\end{itemize}

\section{Solution Design}
\begin{itemize}
	\item Work out concepts for FE debugger tooling to answer RQ2
	\item What do we want to improve compared to existing/previous approaches?
	\item Lay out user stories
\end{itemize}

\section{Study Design}
\begin{itemize}
	\item Propose ways how to validate the proposed solution in order to answer RQ3
\end{itemize}

\section{Threats to Validity}
\begin{itemize}
	\item Informal interviews only
	\begin{itemize}
		\item Describe background of interview partners
		\item Frontend engineering heavy
	\end{itemize}
	\item Real study to validate/reassure?
\end{itemize}

\section{Conclusion}
\begin{itemize}
	\item Recap Past Developments in RP Debugging
	\item Recap proposed solution
	\item Whats next in PA2?
	\item Whats the vision for the thesis?
\end{itemize}

\bibliographystyle{splncs04}
\bibliography{index.bib}

\section*{License}
\ccby\thinspace\thinspace This work is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International License}.
\end{document}

### 2020-04-16 paper.tex
Refine Outline

@@
 
-\newcommand{\version}{v0.1 Outline}
+\newcommand{\version}{v0.2 Draft}
 
@@
 	\par
-	\Huge{tbd. ``Reactive Debugging for modern Frontend Web Applications''}
+	\Huge{Debugging of Modern RxJS-based Frontend Web Applications}
 	\par
@@
 
-\section{Overview}
+\section{Introduction}
+
+\emph{Brief summary over this paper}
+
+\emph{What is RP?}
+
+\emph{What is debugging?}
+
+\emph{Debug concepts in RP}
+
+\emph{Research so far}
+
+\emph{Tooling development so far}
+
+\emph{Conclude with Research Questions}
+
 \begin{itemize}
-	\item What is RP?
-	\item What is debugging?
-	\item Debug concepts in RP
-	\item Research so far
-	\item Tooling development so far
-	\item Research questions:
-	\begin{itemize}
-		\item RQ1 How do web frontend engineers debug RP code?
-		\item RQ2 How can we improve the debugging experience for FE?
-		\item RQ3 Does a specific RP debugger for FE improve the debugging experience?
-	\end{itemize}
-	\item \bf{DEBUGGING EXPERIENCE: How to measure? Use ``time'' instead?}
+	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS based code?}
+	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS based code be improved?}
+	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a frontend engineer?}
 \end{itemize}
 
-\section{``Current State''}
+\subsection{Reactive Programming with RxJS}
+
+\subsection{Debugging Concepts}
+\begin{itemize}
+	\item Imperative
+	\item Declarative
+	\item Omniscient
+\end{itemize}
+
+\subsection{Specialized Reactive Programming Tooling}
+
+\begin{itemize}
+	\item Brief overview on specific tooling
+	\item For frontend, but also other tiers
+	\item Rxviz \footnote{\url{https://rxviz.com/}} see also \url{https://rxviz.com/feedback} for input?
+	\item rxjs-spy \footnote{\url{https://github.com/cartant/rxjs-spy}}
+	\item rxfiddle \cite{10.1145/3180155.3180156}
+	\item reactive-inspector \cite{10.1145/2884781.2884815}
+	\item TOD \cite{5287015}
+\end{itemize}
+
+\section{Interviews and War Stories}
+\begin{itemize}
+	\item We answer RQ1 here
+	\item Present results of own informal interviews
+	\item Combine with war stories
+	\item Extract 3 main challenges
+\end{itemize}
+
+\section{Discussion}
 \begin{itemize}
@@
 	\item \dots and the tools which were built based on it
-	\item Pick up gaps/deficiencies from past research/tools
-
-\end{itemize}
-
-\section{Interviews}
-\begin{itemize}
-	\item Answer RQ1
-	\item Reference previous research
-	\item Present results of own informal interviews
-	\begin{itemize}
-		\item Validate/invalidate results of past research based on interviews
-	\end{itemize}
-	\item Lead over to frontend engineering
-	\item Compose past research with interview results
-\end{itemize}
-
-\section{Solution Design}
-\begin{itemize}
-	\item Work out concepts for FE debugger tooling to answer RQ2
-	\item What do we want to improve compared to existing/previous approaches?
-	\item Lay out user stories
+	\item Set up war story extract to be replicated with existing tooling
 \end{itemize}
@@
 \begin{itemize}
-	\item Propose ways how to validate the proposed solution in order to answer RQ3
+	\item Use prev section to build experiments using existing tooling
+	\item Validate stories and interviews
+\end{itemize}
+
+\section{Future Work}
+\begin{itemize}
+	\item Pick up results from war stories and experiment
+	\item Propose one solution per challenges
+	\item Sketch how this could look like
+	\item This answers RQ2
 \end{itemize}
@@
 \begin{itemize}
-	\item Informal interviews only
+	\item Informal interviews/war stories only
 	\begin{itemize}
@@
 	\end{itemize}
-	\item Real study to validate/reassure?
 \end{itemize}
@@
 \bibliographystyle{splncs04}
-\bibliography{index.bib}
+\bibliography{bibliography.bib}
 

### 2020-04-19 paper.tex
Introduction

@@
 \usepackage{hyperref}
+\usepackage{minted}
+
+\setminted{autogobble=true, tabsize=2, linenos=true, frame=single, breaklines=true}
 
@@
 		\par
-		\vspace{2ex}
-
-		\small{
-			\noindent\textbf{Keywords}: tbd.
-		}
-		\par
 		\vspace{4ex}
@@
 
-\emph{Brief summary over this paper}
+The debugger is an essential tool to get a deeper understanding of the runtime behavior of a software system.
+Software engineers use it to understand why particular circumstances lead to an unexpected failure during execution or to step through unknown program code in order to get themselves accustomed with it.
+
+IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities \footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime.
+With an imperative programming model in mind, they let the user jump from program statement to statement, stop execution upon hitting a breakpoint while having a closer look at the values currently assigned to variables.
+By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before \footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.
+
+Modern web frontend applications use the browser for rendering sophisticated, reactive user interfaces.
+Reactive applications contrast more traditional computer programs: Where i.e.
+the conversion from one image format to another is a one-off process with clearly defined start and end point, a reactive application handles a continuos flow of stimuli: While moving the cursor over UI elements, the application provides steady feedback to the user on the display.
+
+Todays state of the art debugger tools can be used to work with reactive applications as well.
+(Conditional) breakpoints are essential to interrupt the code execution at the right moment to inspect the inner workings of the reactive program.
+As we will see later though, software engineers tend to fall back to a log-based debugging approach when the complexity of a system increases.
+
+% Previous research \cite{10.1145/3180155.3180156} \cite{10.1145/2884781.2884815} in the field of debugging for reactive applications as well as our own experience, particular with RxJS\footnote{\url{https://rxjs.dev/}}-based web applications, show that another review on this area may yield new insights.
+
+Within this paper, we will work on the following two research questions:
+
+\begin{itemize}
+	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS based code?}
+	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS-based code be improved?}
+\end{itemize}
+
+Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give us insight in the challenges they face in their day-to-day jobs.
+We will evaluate and validate three scenarios extracted from their statements in order to answer \emph{RQ1}.
+
+Later, we propose for \emph{RQ2} three solution concepts to resolve previously identified problems which ultimately should improve the debugging experience for RxJS-based code.
+
+The implementation and validation of these proposals is then part of a third research question \emph{RQ3} and will be part of a future iteration on the topic:
+
+\begin{itemize}
+	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a frontend engineer?}
+\end{itemize}
+
+This introductory section will conclude with the clarification of important terms.
+Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes three recurrent challenges.
+With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of reactive applications as well as specific tooling.
+We are going to validate presented challenges against currently available tools within Section~\ref{sec:study} and give an outlook on future iterations on this papers topic right after.
+
+\subsection{Reactive Applications}
+
+\subsection{Reactive Programming with RxJS}
 
@@
 
-\emph{What is debugging?}
-
-\emph{Debug concepts in RP}
-
-\emph{Research so far}
-
-\emph{Tooling development so far}
-
-\emph{Conclude with Research Questions}
-
-\begin{itemize}
-	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS based code?}
-	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS based code be improved?}
-	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a frontend engineer?}
-\end{itemize}
-
-\subsection{Reactive Programming with RxJS}
+\begin{listing}[H]
+	\begin{minted}{TypeScript}
+		// Add Example
+		const x: string = 'Hello World';
+	\end{minted}
+	\caption{Reactive Programming with RxJS}
+	\label{lst:rp-with-rxjs}
+\end{listing}
 
@@
 \section{Interviews and War Stories}
+\label{sec:interviews}
+
 \begin{itemize}
@@
 \section{Discussion}
+\label{sec:discussion}
+
 \begin{itemize}
@@
 \section{Study Design}
+\label{sec:study}
+
 \begin{itemize}
@@
 \section{Future Work}
+\label{sec:future}
+
 \begin{itemize}
@@
 \section{Threats to Validity}
+\label{sec:threats}
+
 \begin{itemize}
@@
 \section{Conclusion}
+\label{sec:conclusion}
+
 \begin{itemize}

### 2020-04-23 paper.tex
RP

@@
 
-\subsection{Reactive Programming with RxJS}
+Salvaneschi et al.
+\cite{comprehendingReactiveApplications} define the ``Reactive Application'' \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
+It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
 
-\emph{What is RP?}
+\begin{figure}[h]
+	\centering
+	\emph{TODO}
+
+	\caption{Traditional vs. Reactive Application}
+	\label{fig:reactive-application}
+\end{figure}
+
+Further Salvaneschi et al.
+provide a set of examples of what a reactive application might achieve with their three-parted classification schema: A \emph{synthetic application} transforms and processes streams of information.
+hence multiple image conversions one after another or in parallel could be implemented as a synthetic, reactive application.
+Beside the modelling of pure data flows, reactive applications are prevalent in the area of user interfaces \emph{(UI)}: They can be identified as \emph{interactive applications} and \emph{graphical animations} here.
+
+Various ways of implementing reactive applications emerged over time.
+The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations providing reactivity \cite{ownPaper}.
+Functional Programming \emph{(FP)} experiments with reactive language extensions like the on animations specialized library Fran (Haskell) \cite{SurveyOnReactiveProgramming} or more recently the reactive runtime and DSL REScala (Scala) \cite{10.1145/2577080.2577083}.
+The declarative nature of FP lead finally to what we understand as ``Reactive Programming'' today.
+
+\subsection{Reactive Programming}
+
+Reactive programming \emph{(RP)} is a programming paradigm.
+
+\subsection{Debugging Concepts}
+
+\begin{itemize}
+	\item Imperative
+	\item Declarative
+	\item Omniscient
+\end{itemize}
+
+\subsection{RxJS}
 
@@
 \end{listing}
-
-\subsection{Debugging Concepts}
-\begin{itemize}
-	\item Imperative
-	\item Declarative
-	\item Omniscient
-\end{itemize}
-
-\subsection{Specialized Reactive Programming Tooling}
-
-\begin{itemize}
-	\item Brief overview on specific tooling
-	\item For frontend, but also other tiers
-	\item Rxviz \footnote{\url{https://rxviz.com/}} see also \url{https://rxviz.com/feedback} for input?
-	\item rxjs-spy \footnote{\url{https://github.com/cartant/rxjs-spy}}
-	\item rxfiddle \cite{10.1145/3180155.3180156}
-	\item reactive-inspector \cite{10.1145/2884781.2884815}
-	\item TOD \cite{5287015}
-\end{itemize}
 
@@
 	\item Set up war story extract to be replicated with existing tooling
+\end{itemize}
+
+\begin{itemize}
+	\item Brief overview on specific tooling
+	\item For frontend, but also other tiers
+	\item Rxviz \footnote{\url{https://rxviz.com/}} see also \url{https://rxviz.com/feedback} for input?
+	\item rxjs-spy \footnote{\url{https://github.com/cartant/rxjs-spy}}
+	\item rxfiddle \cite{10.1145/3180155.3180156}
+	\item reactive-inspector \cite{10.1145/2884781.2884815}
+	\item TOD \cite{5287015}
 \end{itemize}

### 2020-04-23 paper.tex
RP

@@
 
-Reactive programming \emph{(RP)} is a programming paradigm.
+Reactive programming \emph{(RP)} is a declarative programming paradigm.
+Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved.
+The languages runtime system then figures out a way to satisfy that given description.
+RP specifically allows the description of a dependency graph of values and transformations over time.
+
+\begin{figure}[h]
+	\centering
+	\emph{TODO}
+
+	\caption{Reactive Programming Graph}
+	\label{fig:reactive-programming}
+\end{figure}
+
+An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{REScala} \cite{otherRPDSL}.
+At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.
 

### 2020-04-23 paper.tex
Debugging Concepts

@@
 
-\begin{itemize}
-	\item Imperative
-	\item Declarative
-	\item Omniscient
-\end{itemize}
+Debugging concepts can be divided roughly into three categories: Traditional \bf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once the code execution is paused (i.e.
+using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
+Manual control of the programs execution allows to inspect the programs behavior step by step as well as assigning new values to variables ``on-the-fly''.
+
+Where individual transformations of an RP based program can be inspected using a traditional debugger, they fall short in understanding the greater model of the dependency graph.
+In fact, RP provides its own challenges to debuggers.
+So are call stacks spoiled with the invocation of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than any given point in the dependency graph.
+Various \bf{reactive} debuggers and visualization tools for different RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
+They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
+
+In matters of time, traditional as well as reactive debuggers work with the very present state of the programs execution only.
+They do not have any knowledge about what happened before or what is going to happen in the future.
+This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
+An \bf{omniscient} debugger \cite{omniscientDebugger} does not interact with the executed program directly.
+Instead, it records runtime telemetry and provides an interface for later inspection.
+This way, engineers can ``time travel'' back and fore through program execution without the hassle of reconstructing a given failure situation.
+This concept comes with its own set of challenges, i.e.
+demands on storage capacity or user experience \emph{(UX)} design.
+\cite{omniscientDebuggerProblems}
 

### 2020-04-23 paper.tex
Fix Styling

@@
 
-Debugging concepts can be divided roughly into three categories: Traditional \bf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once the code execution is paused (i.e.
+Debugging concepts can be divided roughly into three categories: Traditional \textbf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once the code execution is paused (i.e.
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
@@
 So are call stacks spoiled with the invocation of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than any given point in the dependency graph.
-Various \bf{reactive} debuggers and visualization tools for different RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
+Various \textbf{reactive} debuggers and visualization tools for different RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
 They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
@@
 This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
-An \bf{omniscient} debugger \cite{omniscientDebugger} does not interact with the executed program directly.
+An \textbf{omniscient} debugger \cite{omniscientDebugger} does not interact with the executed program directly.
 Instead, it records runtime telemetry and provides an interface for later inspection.

### 2020-04-23 paper.tex
Rephrase

@@
 Various ways of implementing reactive applications emerged over time.
-The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations providing reactivity \cite{ownPaper}.
-Functional Programming \emph{(FP)} experiments with reactive language extensions like the on animations specialized library Fran (Haskell) \cite{SurveyOnReactiveProgramming} or more recently the reactive runtime and DSL REScala (Scala) \cite{10.1145/2577080.2577083}.
-The declarative nature of FP lead finally to what we understand as ``Reactive Programming'' today.
+The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{ownPaper}.
+Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{SurveyOnReactiveProgramming} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
+The declarativeness of FP finally lead to what we understand under ``Reactive Programming'' today.
 

### 2020-04-24 paper.tex
Edit Debugger Concepts

@@
 
-Debugging concepts can be divided roughly into three categories: Traditional \textbf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once the code execution is paused (i.e.
+We divide debugging concepts roughly into three categories: Traditional \textbf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
@@
 
-Where individual transformations of an RP based program can be inspected using a traditional debugger, they fall short in understanding the greater model of the dependency graph.
-In fact, RP provides its own challenges to debuggers.
-So are call stacks spoiled with the invocation of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than any given point in the dependency graph.
-Various \textbf{reactive} debuggers and visualization tools for different RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
+Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the greater model of the dependency graph.
+In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph.
+Various \textbf{reactive} debuggers and visualization tools for specific RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
 They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
 
-In matters of time, traditional as well as reactive debuggers work with the very present state of the programs execution only.
-They do not have any knowledge about what happened before or what is going to happen in the future.
+In matters of time, traditional as well as reactive debuggers work with the present state of the programs execution only.
+They lack knowledge about what happened before or what is going to happen in the future.
 This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
-An \textbf{omniscient} debugger \cite{omniscientDebugger} does not interact with the executed program directly.
+An \textbf{omniscient} debugger \cite{omniscientDebuggerPaper} \cite{mozillaRR} does not interact with the executed program directly.
 Instead, it records runtime telemetry and provides an interface for later inspection.
-This way, engineers can ``time travel'' back and fore through program execution without the hassle of reconstructing a given failure situation.
+Engineers can ``time travel'' back and fore through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
 This concept comes with its own set of challenges, i.e.
-demands on storage capacity or user experience \emph{(UX)} design.
+demands on storage capacity or efficient user experience \emph{(UX)} design.
 \cite{omniscientDebuggerProblems}

### 2020-04-24 paper.tex
RxJS

@@
 
-\subsection{RxJS}
+\subsection{ReactiveX and RxJS}
+Reactive Extensions is an open source project providing an API description for RP concepts.
+It describes itself as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming'' \cite{reactivex}.
+The API specification introduces the concept of the  \mintedinline{Observable} at its core.
+An \mintedinline{Observable} has the ability to be composed with other \mintedinline{Observable}s and eventually describe the dependency graph known in RP.
+A multitude of operators allow the transformation of values and \mintedinline{Observable}s within the graph.
+
+The ReactiveX community works on a variety of reference implementations\footnote{\url{http://reactivex.io/languages.html}}: Libraries for Java, C#, JavaScript and many more allow to work with similar API semantics on different software architecture levels.
+
+RxJS is the reference implementation for JavaScript.
+Its current major version 6 is completely done in TypeScript\footnote{\url{https://www.typescriptlang.org/}} and provides the core elements of the ReactiveX API specification.
+Large projects like Angular embed and use RxJS\cite{angularrxjs}.
+The Listing \ref{lst:rp-with-rxjs} shows a short example of RP using RxJS.
 

### 2020-04-24 paper.tex
Fix Build & Minor Edits

--- 2020-04-24 paper.tex
@@
 
-Within this paper, we will work on the following two research questions:
+Within this paper, we will answer the following two research questions:
 
@@
 
-The implementation and validation of these proposals is then part of a third research question \emph{RQ3} and will be part of a future iteration on the topic:
+The implementation and validation of these proposals is then part of a third research question \emph{RQ3} and will be part of a future iteration on the topic and not part of this paper:
 
@@
 
-This introductory section will conclude with the clarification of important terms.
+We will conclude this introductory section with the clarification of important terms.
 Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes three recurrent challenges.
 With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of reactive applications as well as specific tooling.
-We are going to validate presented challenges against currently available tools within Section~\ref{sec:study} and give an outlook on future iterations on this papers topic right after.
+We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to the validity of our work with Section~\ref{sec:threats} and finally end with an outlook on future iterations.
 
@@
 Salvaneschi et al.
-\cite{comprehendingReactiveApplications} define the ``Reactive Application'' \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
+\cite{comprehendingReactiveApplications} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
 It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
@@
 
-Further Salvaneschi et al.
-provide a set of examples of what a reactive application might achieve with their three-parted classification schema: A \emph{synthetic application} transforms and processes streams of information.
-hence multiple image conversions one after another or in parallel could be implemented as a synthetic, reactive application.
-Beside the modelling of pure data flows, reactive applications are prevalent in the area of user interfaces \emph{(UI)}: They can be identified as \emph{interactive applications} and \emph{graphical animations} here.
+Further, Salvaneschi et al.
+provide a set of examples of what a RA might achieve using a three-parted classification schema: A \emph{synthetic application} transforms and processes streams of information.
+hence multiple image conversions one after another or in parallel could be implemented as a synthetic, RA.
+Beside the modelling of pure data flows, RAs are prevalent in the area of user interfaces \emph{(UI)}: They can be identified as \emph{interactive applications} and \emph{graphical animations} here.
 
-Various ways of implementing reactive applications emerged over time.
+Various ways of implementing RAs emerged over time.
 The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{ownPaper}.
@@
 
-We divide debugging concepts roughly into three categories: Traditional \textbf{imperative-focused} debuggers provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
+We divide debugging concepts roughly into three categories: Traditional \textbf{imperative-focused debuggers} provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
@@
 In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph.
-Various \textbf{reactive} debuggers and visualization tools for specific RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
+Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
 They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
@@
 This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
-An \textbf{omniscient} debugger \cite{omniscientDebuggerPaper} \cite{mozillaRR} does not interact with the executed program directly.
+An \textbf{omniscient debugger} \cite{omniscientDebuggerPaper} \cite{mozillaRR} does not interact with the executed program directly.
 Instead, it records runtime telemetry and provides an interface for later inspection.
@@
 
-\subsection{RxJS}
+\subsection{ReactiveX and RxJS}
+Reactive Extensions is an open source project providing an API description for RP concepts.
+It describes itself as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming'' \cite{reactivex}.
+The API specification introduces the concept of the  \mintinline{TypeScript}{Observable} at its core.
+An \mintinline{TypeScript}{Observable} has the ability to be composed with other \mintinline{TypeScript}{Observable}s and eventually describe the dependency graph known in RP.
+A multitude of operators allow the transformation of values and \mintinline{TypeScript}{Observable}s within the graph.
+
+The ReactiveX community works on a variety of reference implementations\footnote{\url{http://reactivex.io/languages.html}}: Libraries for Java, C\#, JavaScript and many more allow to work with similar API semantics on different software architecture levels.
+
+RxJS is the reference implementation for JavaScript.
+Its current major version 6 is completely done in TypeScript\footnote{\url{https://www.typescriptlang.org/}} and provides the core elements of the ReactiveX API specification.
+Large projects like Angular embed and use RxJS\cite{angualrrxjs}.
+The Listing \ref{lst:rp-with-rxjs} shows a short example of RP using RxJS.
 

### 2020-04-24 paper.tex
Align Readme

@@
 Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give us insight in the challenges they face in their day-to-day jobs.
-We will evaluate and validate three scenarios extracted from their statements in order to answer \emph{RQ1}.
+We will evaluate and validate three cases extracted from their statements in order to answer \emph{RQ1}.
 
@@
 With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of reactive applications as well as specific tooling.
-We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to the validity of our work with Section~\ref{sec:threats} and finally end with an outlook on future iterations.
+We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.
 

### 2020-05-07 paper.tex
Interviews

@@
 
+\subsection{Informal Interviews}
+
+In order to answer RQ1 and get an insight in how software engineers apply RP, we started with an early series of informal interviews.
+The only requirement to qualify for the interview was to apply RP in some way, hence no restriction in regards of tech stack was given.
+In the end we interviewed five engineers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
+Two of them stated to have more explicit experience with RxJS further.
+The fifth subject works currently with a combination of akka\footnote{\url{todo}} and akka-streams\footnote{\url{todo}} in Scala to create backend applications.
+The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.
+
+We identified several recurring topics throughout all the interviews.
+All of the subjects noted that they think of RP as an invaluable tool to describe and model dependencies between multiple data sources.
+Most of them appreciate how RP allows them to implement data transformations in a concise and composable way.
+The complexity of concurrent programming can be hidden well behind a simple DSL one subject noted further.
+
+For the implementation, all subjects noted that a language with a strong type system is to be preferred: The compiler gives the guarantee of basic code correctness which is why e.g.
+TypeScript is used by all the subjects working with RxJS and Angular.
+
+None of the subjects remembered debugging RP code to be enjoyable in particular, nonetheless the technology used to do so.
+All of the subjects use traditional debugging tools.
+Redux DevTools \footnote{\url{todo}} is in use by all four subjects working with ngrx.
+There was mention of any more specific debugging tool in regards of akka and akka-streams.
+
+One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented.
+(ii) In order to understand how data traverses through the graph, subjects augmumented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{TODO}} operator to log values.
+A less invasive way would be preferable as stated.
+
+\subsection{War Stories}
+
+\subsection{Main Challenges}
+
 \begin{itemize}
@@
 
+\section{Appendix}
+
+\subsection{Interview Questions}
+\label{sec:interviewquestions}
+
+\begin{itemize}
+	\item When did you last apply RP?
+	\item What technology did you use RP with?
+	\item What was the best experience you had with RP?
+	\item \dots and what was the worst? \dots Can you elaborate on the concrete situation?
+\end{itemize}
+
 \bibliographystyle{splncs04}

### 2020-05-07 paper.tex
Typo

@@
 One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented.
-(ii) In order to understand how data traverses through the graph, subjects augmumented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{TODO}} operator to log values.
+(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{TODO}} operator to log values.
 A less invasive way would be preferable as stated.

### 2020-05-07 paper.tex
War Stories

@@
 
-In order to answer RQ1 and get an insight in how software engineers apply RP, we started with an early series of informal interviews.
-The only requirement to qualify for the interview was to apply RP in some way, hence no restriction in regards of tech stack was given.
-In the end we interviewed five engineers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
+In order to prepare to give an answer on the first research question, we started with a series of informal interviews.
+We talked to professionals about their experience in working with RP in their daily jobs.
+We did not limit the selection of our interview partners by the tech stack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage.
+Eventually, we interviewed five engineers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
 Two of them stated to have more explicit experience with RxJS further.
@@
 
-None of the subjects remembered debugging RP code to be enjoyable in particular, nonetheless the technology used to do so.
+None of the subjects remembered debugging RP code to be a positive experience, independent of the technology used.
 All of the subjects use traditional debugging tools.
@@
 One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented.
-(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{TODO}} operator to log values.
+(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{todo}} operator to log values.
 A less invasive way would be preferable as stated.
@@
 \subsection{War Stories}
+
+Based on the results of the interviews, we compiled a request for more specific factual reports.
+We asked professional software engineers about their most recent debugging experience in regards of RxJS, either good or bad.
+We were able to gather responses from one RxJS core team member, two Google Developer Experts and two engineers building web and mobile applications for a fintech company eventually.
+
+We could verify gathered interview data with the results from the new reports: Subjects tend to add log statements to their code in order to understand its behavior, especially when things get more complex.
+Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{url{todo}}.
+This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe.
+In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjs-lifecycle} \cite{rxjs-spy-events}.
+As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels.
+Without it, they were blind to what actually happens within the system.
+
+In addition to more sophisticated logging, three subjects reported that they built a habit of decomposing complex graphs into smaller, simpler bits and running those in a dedicated environment like Rx Visualizer\footnote{\url{https://rxviz.com/}} or StackBlitz\footnote{\url{https://stackblitz.com/}}.
 

### 2020-05-07 paper.tex
Summarize

@@
 We will conclude this introductory section with the clarification of important terms.
-Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes three recurrent challenges.
+Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes two recurrent challenges.
 With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of reactive applications as well as specific tooling.
@@
 \subsection{ReactiveX and RxJS}
+
 Reactive Extensions is an open source project providing an API description for RP concepts.
@@
 In order to prepare to give an answer on the first research question, we started with a series of informal interviews.
-We talked to professionals about their experience in working with RP in their daily jobs.
-We did not limit the selection of our interview partners by the tech stack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage.
-Eventually, we interviewed five engineers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
+We talked to professionals about their experiestack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage.
+Eventually, we interviewed five engince in working with RP in their daily jobs.
+We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
 Two of them stated to have more explicit experience with RxJS further.
@@
 
-We could verify gathered interview data with the results from the new reports: Subjects tend to add log statements to their code in order to understand its behavior, especially when things get more complex.
+We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex.
 Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{url{todo}}.
@@
 
-\subsection{Main Challenges}
+\subsection{Summary}
 
-\begin{itemize}
-	\item We answer RQ1 here
-	\item Present results of own informal interviews
-	\item Combine with war stories
-	\item Extract 3 main challenges
-\end{itemize}
+The interpretation of the combined interview and war stories data gives us an idea about two main challenges software engineers have to overcome during debugging activities of RxJS-based code:
+
+It is crucial to the engineers to \textbf{understand data flows}  within the \mintinline{TypeScript}{Observable} graph.
+Questions like ``Which source emitted a value at what time?'' or ``What is the result of a specific transformation?'' are common in this context.
+Once a program fails, software engineers need to \textbf{rely on traditional debugger tools}.
+Though providing help to some extent, the tools lack of specific knowledge about the RP runtime environment adds some oddities to the debugging process: The (i) functionalities allowing step-wise navigation through the code does not respect the RP context.
+Assume we have set a breakpoint at \mintinline{TypeScript}{i * 2} on line~2 of Listing~\ref{lst:rxjs-step-controls}.
+Once the debugger hits the breakpoint and we instruct it to ``Step Over'', the debugger will lead us to the internal implementation of the \mintinline{TypeScript}{map} operator instead to line~3.
+(ii) Stack traces include information about the invocation of internal functions which, most of the time, do not provide any valuable information to the engineers further.
+
+\begin{listing}[H]
+	\begin{minted}{TypeScript}
+		timer().pipe(
+		  map(i => i * 2),
+		  map(i => i - 1)
+		).subscribe();
+	\end{minted}
+	\caption{Example RxJS Stream With Breakpoint on Line 2 Within the Arrow Function}
+	\label{lst:rxjs-step-controls}
+\end{listing}
 

### 2020-05-16 paper.tex
Reactive Inspector

@@
 
-\section{Discussion}
+\section{Discussion of Previous Efforts}
 \label{sec:discussion}
+
+Our work is not the first approach on understanding and improving on the struggles software engineers face when working on RP code, no matter of using RxJS or not.
+We are going to present academic as well as non-academic efforts which contributed in the field of RP debugging in the past.
+
+\subsection{Reactive Inspector}
+
+Salvaneschi et al.
+did focus their research heavily on RP during the mid 2010s.
+Beside the creation of REScala in 2014 \cite{REScala}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well.
+In the paper ``Debugging for Reactive Programming'' \cite{positiveEffectsRP}, Salvaneschi et al.
+elaborate on the necessity of RP debugging tools which can understand the specific language abstractions.
+With the plugin \emph{Reactive Inspector} \footnote{\url{http://guidosalva.github.io/reactive-inspector/}} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
+One of its main features is the visualization of the internal dependency graph.
+The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph.
+The execution of the program can be interrupted e.g.
+once a node retains a specific output value or an event of interest takes place in the runtime.
+Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at any time.
+
+\begin{itemize}
+	\item Visualization of the graph
+	\item Omniscient
+
+Herman Benken \cite{rxfiddle}
+
+rxspy-js
 
@@
 		\item Describe background of interview partners
-		\item Frontend engineering heavy
+		\item Only 4 Participants in experiment
+		\item Code in experiment carefully crafted... Though is it representative?
+		\
 	\end{itemize}

### 2020-05-16 paper.tex
Reactive Inspector
RxFiddle

@@
 once a node retains a specific output value or an event of interest takes place in the runtime.
-Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at any time.
+Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.
 
-\begin{itemize}
-	\item Visualization of the graph
-	\item Omniscient
+\subsection{rxfiddle}
 
-Herman Benken \cite{rxfiddle}
+\emph{rxfiddle} proposed by Banken et al.
+is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{rxfiddle}.
+The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g.
+when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected.
+The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{storyflow} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s.
+An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time.
+They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.
+
+The reference implementation supports event collection for RxJS-based code in version 4 and 5.
+It can either be used on a local development machine on existing projects or by copying code snippets to the online available.
 

### 2020-05-16 paper.tex
Citations and Footnotes

@@
 Salvaneschi et al.
-\cite{comprehendingReactiveApplications} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
+\cite{7827078} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
 It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
@@
 Various ways of implementing RAs emerged over time.
-The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{ownPaper}.
-Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{SurveyOnReactiveProgramming} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
+The Observer design pattern \cite{gamma1995design}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{alabor:2019:reactiveappllications}.
+Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
 The declarativeness of FP finally lead to what we understand under ``Reactive Programming'' today.
@@
 
-An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{REScala} \cite{otherRPDSL}.
+An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{10.1145/2577080.2577083}.
 At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.
@@
 In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph.
-Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field.
+Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{10.1145/2884781.2884815} \cite{rxviz} \cite{10.1145/3180155.3180156} exploring this field.
 They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
@@
 This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
-An \textbf{omniscient debugger} \cite{omniscientDebuggerPaper} \cite{mozillaRR} does not interact with the executed program directly.
+An \textbf{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
 Instead, it records runtime telemetry and provides an interface for later inspection.
@@
 Eventually, we interviewed five engince in working with RP in their daily jobs.
-We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications.
+We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{https://angular.io/}} and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend applications.
 Two of them stated to have more explicit experience with RxJS further.
-The fifth subject works currently with a combination of akka\footnote{\url{todo}} and akka-streams\footnote{\url{todo}} in Scala to create backend applications.
+The fifth subject works currently with a combination of akka\footnote{\url{https://akka.io/}} and akka-streams in Scala to create backend applications.
 The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.
@@
 All of the subjects use traditional debugging tools.
-Redux DevTools \footnote{\url{todo}} is in use by all four subjects working with ngrx.
-There was mention of any more specific debugging tool in regards of akka and akka-streams.
+Redux\footnote{\url{https://redux.js.org/}} DevTools \footnote{\url{https://github.com/reduxjs/redux-devtools}} is in use by all four subjects working with ngrx.
+There was no mention of any more specific debugging tool in regards of akka and akka-streams.
 
 One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented.
-(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{todo}} operator to log values.
-A less invasive way would be preferable as stated.
+(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{https://rxjs.dev/api/operators/tap}} side effect operator to log values.
+A less manually invasive way would be preferable as stated.
 
@@
 We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex.
-Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{url{todo}}.
+Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{\url{https://cartant.github.io/rxjs-spy/}}.
 This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe.
-In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjs-lifecycle} \cite{rxjs-spy-events}.
+In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjslifecycle}.
 As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels.
@@
 did focus their research heavily on RP during the mid 2010s.
-Beside the creation of REScala in 2014 \cite{REScala}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well.
-In the paper ``Debugging for Reactive Programming'' \cite{positiveEffectsRP}, Salvaneschi et al.
+Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well.
+In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al.
 elaborate on the necessity of RP debugging tools which can understand the specific language abstractions.
-With the plugin \emph{Reactive Inspector} \footnote{\url{http://guidosalva.github.io/reactive-inspector/}} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
+With the plugin \emph{Reactive Inspector} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
 One of its main features is the visualization of the internal dependency graph.
@@
 \emph{rxfiddle} proposed by Banken et al.
-is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{rxfiddle}.
+is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
 The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g.
 when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected.
-The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{storyflow} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s.
+The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s.
 An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time.
@@
 It can either be used on a local development machine on existing projects or by copying code snippets to the online available.
-
-rxspy-js
-
-\begin{itemize}
-	\item Closer look to past Research
-	\item \dots and the tools which were built based on it
-	\item Set up war story extract to be replicated with existing tooling
-\end{itemize}
-
-\begin{itemize}
-	\item Brief overview on specific tooling
-	\item For frontend, but also other tiers
-	\item Rxviz \footnote{\url{https://rxviz.com/}} see also \url{https://rxviz.com/feedback} for input?
-	\item rxjs-spy \footnote{\url{https://github.com/cartant/rxjs-spy}}
-	\item rxfiddle \cite{10.1145/3180155.3180156}
-	\item reactive-inspector \cite{10.1145/2884781.2884815}
-	\item TOD \cite{5287015}
-\end{itemize}
 

### 2020-05-16 paper.tex
RxJS Example
Clean Up Figures & Examples

@@
 
-\begin{figure}[h]
-	\centering
-	\emph{TODO}
-
-	\caption{Traditional vs. Reactive Application}
-	\label{fig:reactive-application}
-\end{figure}
-
 Further, Salvaneschi et al.
@@
 RP specifically allows the description of a dependency graph of values and transformations over time.
-
-\begin{figure}[h]
-	\centering
-	\emph{TODO}
-
-	\caption{Reactive Programming Graph}
-	\label{fig:reactive-programming}
-\end{figure}
 
@@
 	\begin{minted}{TypeScript}
-		// Add Example
-		const x: string = 'Hello World';
+		of('hello', 'foo', 'world').pipe(
+			filter(s => s !== 'foo'),
+			map(s => s.toUpperCase())
+		).subscribe(console.log)             // HELLO, WORLD
 	\end{minted}
-	\caption{Reactive Programming with RxJS}
+	\caption{RxJS Code Example}
 	\label{lst:rp-with-rxjs}
@@
 	\begin{minted}{TypeScript}
-		timer().pipe(
-		  map(i => i * 2),
+		interval(1000).pipe(      // Increasing numbers with 1s delay
+		  map(i => i * 2),        // Breakpoint on i * 2
 		  map(i => i - 1)
-		).subscribe();
+		).subscribe(console.log); // -1, 1, 3, 5, 7, ...
 	\end{minted}

### 2020-05-17 paper.tex
rxjs-spy

@@
 
-Our work is not the first approach on understanding and improving on the struggles software engineers face when working on RP code, no matter of using RxJS or not.
-We are going to present academic as well as non-academic efforts which contributed in the field of RP debugging in the past.
+Our work is not the first approach on understanding and improving on the struggles software engineers have to overcome when working on RP code, either with or without RxJS.
+We are going to present four of this tools in this section.
+The list is not meant to be exhaustive.
+It contains tools we either got to know during our research, because they were mentioned in one of the interviews or we worked with it personally in the past.
 
@@
 It can either be used on a local development machine on existing projects or by copying code snippets to the online available.
+
+\subsection{rxjs-spy}
+
+Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
+\emph{rxjs-spy} wants to improve on this by introducing the new \mintinline{TypeScript}{tag} operator.
+Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are, hopefully, created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
+
+\begin{listing}[H]
+	\begin{minted}{TypeScript}
+		interval(1000).pipe(
+		  map(i => i * 2),
+		  tag('multiply'),
+		  map(i => i - 1)
+		).subscribe(console.log);
+	\end{minted}
+	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator}
+	\label{lst:rxjs-spy-tag}
+\end{listing}
+
+Later, \emph{rxjs-spy} can be told to log events (subscribe, emit, error, complete and unsubscribe) of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
+
+\subsection{RxViz}
 

### 2020-05-17 paper.tex
rxjs-spy

@@
 We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex.
-Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{\url{https://cartant.github.io/rxjs-spy/}}.
+Three subjects even claimed using specialized logging functionalities provided through rxjs-spy.
 This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe.
@@
 The list is not meant to be exhaustive.
-It contains tools we either got to know during our research, because they were mentioned in one of the interviews or we worked with it personally in the past.
+It contains tools we either got to know during our research, because they were mentioned in one of the interviews or we worked with them personally in the past.
 
@@
 Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
-\emph{rxjs-spy} wants to improve on this by introducing the new \mintinline{TypeScript}{tag} operator.
-Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are, hopefully, created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
+\emph{rxjs-spy}\footnote{\url{https://cartant.github.io/rxjs-spy/}} wants to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
+Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
+A tag as shown in Listing~\ref{lst:rxjs-spy-tag} does not influence runtime behavior at first.
 
@@
 		  map(i => i * 2),
-		  tag('multiply'),
+		  tag('multiply'),        // Create tag "multiply"
 		  map(i => i - 1)
@@
 
-Later, \emph{rxjs-spy} can be told to log events (subscribe, emit, error, complete and unsubscribe) of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
+Later, \emph{rxjs-spy} can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
+It is worth to mention that manual log traces, without additional effort, only show values flowing by.
+The additional events logged by \emph{rxjs-spy} proofed being invaluable in our own experience multiple times.
+
+The library provides further features to interact with tagged \mintinline{TypeScript}{Observable}s through the debuggers console.
+
+Additional features are available through the debuggers console interface.
+A tagged \mintinline{TypeScript}{Observable} can be paused so it collects incoming values.
+They can then be emitted one after another or resumed at once.
 

### 2020-05-17 paper.tex
rxjs-spy & rxviz

@@
 In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph.
-Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{10.1145/2884781.2884815} \cite{rxviz} \cite{10.1145/3180155.3180156} exploring this field.
+Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz} exploring this field.
 They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
@@
 
-In addition to more sophisticated logging, three subjects reported that they built a habit of decomposing complex graphs into smaller, simpler bits and running those in a dedicated environment like Rx Visualizer\footnote{\url{https://rxviz.com/}} or StackBlitz\footnote{\url{https://stackblitz.com/}}.
+In addition to more sophisticated logging, three subjects reported that they built a habit of decomposing complex graphs into smaller, simpler bits and running those in a dedicated environment like Rx Visualizer\cite{rxviz} or StackBlitz\footnote{\url{https://stackblitz.com/}}.
 
@@
 Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
-\emph{rxjs-spy}\footnote{\url{https://cartant.github.io/rxjs-spy/}} wants to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
+\emph{rxjs-spy}\cite{rxjsspy} wants to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
 Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
@@
 
-\subsection{RxViz}
+\subsection{Rx Vizualizer}
+
+The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser.
+It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.
+
+\subsection{Summary}
 

### 2020-05-17 paper.tex
Edit

@@
 
-IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities \footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime.
+IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime.
 With an imperative programming model in mind, they let the user jump from program statement to statement, stop execution upon hitting a breakpoint while having a closer look at the values currently assigned to variables.
-By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before \footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.
+By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before\footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.
 
@@
 We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex.
-Three subjects even claimed using specialized logging functionalities provided through rxjs-spy.
-This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe.
-In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjslifecycle}.
-As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels.
-Without it, they were blind to what actually happens within the system.
+Three subjects even claimed using specialized logging functionalities provided by rxjs-spy\cite{rxjsspy}.
+As the subjects stated, the knowledge about additional events made visible helped in situations where they needed to find problems within a graph composing multiple asynchronous data sources on multiple levels.
+Without it, they would have been blind to what actually happened within the system.
 
@@
 elaborate on the necessity of RP debugging tools which can understand the specific language abstractions.
-With the plugin \emph{Reactive Inspector} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
+With the plugin Reactive Inspector for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
 One of its main features is the visualization of the internal dependency graph.
@@
 once a node retains a specific output value or an event of interest takes place in the runtime.
-Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.
+Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.
 
@@
 
-\emph{rxfiddle} proposed by Banken et al.
+rxfiddle proposed by Banken et al.
 is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
@@
 Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
-\emph{rxjs-spy}\cite{rxjsspy} wants to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
+rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
 Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
@@
 
-Later, \emph{rxjs-spy} can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
+Later, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
 It is worth to mention that manual log traces, without additional effort, only show values flowing by.
-The additional events logged by \emph{rxjs-spy} proofed being invaluable in our own experience multiple times.
-
-The library provides further features to interact with tagged \mintinline{TypeScript}{Observable}s through the debuggers console.
+The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.
 
@@
 
-\subsection{Rx Vizualizer}
+\subsection{Rx Visualizer}
 
@@
 It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.
+Diagrams are generated over a given, fixed time and can be saved as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}}.
+
+The visualizer helps to understand the behavior of a specific part of an \mintinline{TypeScript}{Observable} stream in a sandboxed environment.
 

### 2020-05-21 paper.tex
Edit
Comparison

@@
 As the subjects stated, the knowledge about additional events made visible helped in situations where they needed to find problems within a graph composing multiple asynchronous data sources on multiple levels.
-Without it, they would have been blind to what actually happened within the system.
+Without it, they would have been blind to what actually happend within the system.
 
@@
 		  map(i => i * 2),        // Breakpoint on i * 2
-		  map(i => i - 1)
-		).subscribe(console.log); // -1, 1, 3, 5, 7, ...
+		  map(i => i - 1),
+		  take(2)
+		).subscribe(console.log); // -1, 1
 	\end{minted}
@@
 We are going to present four of this tools in this section.
-The list is not meant to be exhaustive.
-It contains tools we either got to know during our research, because they were mentioned in one of the interviews or we worked with them personally in the past.
+This list is not meant to be exhaustive.
+It contains tools we either got to know during our research because they were mentioned in one of the interviews or we worked with them personally in the past.
 
@@
 
+Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with different RP runtime environment as well.
+
 \subsection{rxfiddle}
@@
 
+The architectural design of \emph{rxfiddle} is similar to the one \emph{Reactive Inspector}: Banken et al.
+provide a system where different RP runtime environments should be able to be attached to the Visualizer easily.
+
 \subsection{rxjs-spy}
@@
 		  map(i => i * 2),
-		  tag('multiply'),        // Create tag "multiply"
-		  map(i => i - 1)
-		).subscribe(console.log);
+		  tag('multiply'),
+		  map(i => i - 1),
+		  take(2)
+		).subscribe();
 	\end{minted}
-	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator}
+	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line 3}
 	\label{lst:rxjs-spy-tag}
@@
 The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.
+
+\begin{listing}[H]
+	\begin{minted}{Typescript}
+		Tag = multiply; notification = subscribe
+		Tag = multiply; notification = next; value = 0
+		Tag = multiply; notification = next; value = 2
+		Tag = multiply; notification = unsubscribe
+	\end{minted}
+	\caption{Trace log generated by \emph{rxjs-spy} \mintinline{TypeScript}{tag} from Listing~\ref{lst:rxjs-spy-tag}}
+	\label{lst:rxjs-spy-log}
+\end{listing}
 
@@
 
-\subsection{Summary}
+\subsection{Comparison}
+
+Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools.
+Most of them provide a way to represent the data flow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy).
+They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in the visual graph representation (Reactive Inspector).
+Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-based tools have a more specific feature set: They usually work outside of an IDE and/or need manual code instrumentation like rxjs-spy.
+
+\textbf{TODO Compare im Table?}
 

### 2020-05-21 paper.tex
Rewrite Intro

@@
 
-The debugger is an essential tool to get a deeper understanding of the runtime behavior of a software system.
-Software engineers use it to understand why particular circumstances lead to an unexpected failure during execution or to step through unknown program code in order to get themselves accustomed with it.
+An application providing a user interface (\emph{UI}) is a reactive application (\emph{RA}): It handles continuos flow of user and system events in order to give the user a representation of the applications state as timely as possible.
+One way of implementing reactive behavior is the Observer design pattern \cite{gamma1995design} where an observer (e.g.
+the UI rendering code) is repeatedly informed about changed application state.
+Different ways to implement a RA were developed since the emergence of the Observer pattern as well.
+Reactive Programming (\emph{RP}) enabled through language extensions, runtime environemnts and domain specific languages (\emph{DSL}) is based on functional programming (\emph{FP}) principles and allows to describe dependencies and transformations within a RA declerativly.
 
-IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime.
-With an imperative programming model in mind, they let the user jump from program statement to statement, stop execution upon hitting a breakpoint while having a closer look at the values currently assigned to variables.
-By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before\footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.
+Debugging tools are essential to software engineers in order to get deeper understanding of the runtime behavior of a software system.
+Engineers use them to understand why particular circumstances lead to unexpected failures during execution or to step through unknown source code getting themselves accustomed with it.
+IDEs as well as the developer tools of modern web browsers\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} provide battle-tested debugger utilities to inspect code.
+They let engineers step from program statement to statement, stop execution upon hitting a breakpoint and give insight on the values assigned to variables.
 
-Modern web frontend applications use the browser for rendering sophisticated, reactive user interfaces.
-Reactive applications contrast more traditional computer programs: Where i.e.
-the conversion from one image format to another is a one-off process with clearly defined start and end point, a reactive application handles a continuos flow of stimuli: While moving the cursor over UI elements, the application provides steady feedback to the user on the display.
+Todays debugging tools can not understand abstractions introduced by RP libraries.
+E.g.
+where the engineer can use the stack trace to track a problem to its root cause in a traditional imperative program, this does not work when an RP environment is used: The debugger does not have any knowledge of the underlying data flow graph which would allow a similar backtracking from graph node to node.
 
-Todays state of the art debugger tools can be used to work with reactive applications as well.
-(Conditional) breakpoints are essential to interrupt the code execution at the right moment to inspect the inner workings of the reactive program.
-As we will see later though, software engineers tend to fall back to a log-based debugging approach when the complexity of a system increases.
-
-% Previous research \cite{10.1145/3180155.3180156} \cite{10.1145/2884781.2884815} in the field of debugging for reactive applications as well as our own experience, particular with RxJS\footnote{\url{https://rxjs.dev/}}-based web applications, show that another review on this area may yield new insights.
-
-Within this paper, we will answer the following two research questions:
+We will elaborate on such challenges in the field of RxJS-based frontend web applications and search answers to the following two research questions:
 
@@
 
-Later, we propose for \emph{RQ2} three solution concepts to resolve previously identified problems which ultimately should improve the debugging experience for RxJS-based code.
+Later, we propose three solution concepts for \emph{RQ2} to resolve previously identified problems which ultimately should improve the debugging experience for RxJS-based code.
 
-The implementation and validation of these proposals is then part of a third research question \emph{RQ3} and will be part of a future iteration on the topic and not part of this paper:
+The future implementation and validation of these proposals leads to a third research question \emph{RQ3} which will be part of future work not part of this paper:
 
@@
 Salvaneschi et al.
-\cite{7827078} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information.
+\cite{7827078} define the RA as a computer program which processes a continuos, indefinite and ever-changing flow of information.
 It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
@@
 
-An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{10.1145/2577080.2577083}.
+An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a DSL\cite{10.1145/2577080.2577083}.
 At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.
@@
 
-In matters of time, traditional as well as reactive debuggers work with the present state of the programs execution only.
+Traditional as well as a reactive debuggers work with the very current state of the programs execution only.
 They lack knowledge about what happened before or what is going to happen in the future.
@@
 Engineers can ``time travel'' back and fore through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
-This concept comes with its own set of challenges, i.e.
-demands on storage capacity or efficient user experience \emph{(UX)} design.
-\cite{omniscientDebuggerProblems}
+This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows.
+As a result, user experience (\emph{UX}) concepts need to be aligned for the engineers being able to interact with the sheer amount of information efficiently.
 
@@
 \begin{itemize}
+	\item Hypothesis
 	\item Use prev section to build experiments using existing tooling

### 2020-05-21 paper.tex
Fix Copy-Paste

@@
 In order to prepare to give an answer on the first research question, we started with a series of informal interviews.
-We talked to professionals about their experiestack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage.
-Eventually, we interviewed five engince in working with RP in their daily jobs.
-We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{https://angular.io/}} and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend applications.
+We talked to professionals about their experience in working with RP in their daily jobs.
+We did not limit the selection of our interview partners by the tech stack of including RxJS since we were interested in as many different perspectives on the topic as possible in this early stage.
+Eventually, we interviewed six engineers: Four work or worked more recently with Angular\footnote{\url{https://angular.io/}} and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend applications.
 Two of them stated to have more explicit experience with RxJS further.
-The fifth subject works currently with a combination of akka\footnote{\url{https://akka.io/}} and akka-streams in Scala to create backend applications.
+One subject works currently with a combination of akka\footnote{\url{https://akka.io/}} and akka-streams in Scala to create backend applications.
+The sixth subject was the creator of rxfiddle.
 The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.
@@
 As the subjects stated, the knowledge about additional events made visible helped in situations where they needed to find problems within a graph composing multiple asynchronous data sources on multiple levels.
-Without it, they would have been blind to what actually happend within the system.
+Without it, they would have been blind to what actually happened within the system.
 

### 2020-05-23 paper.tex
v0.3
Rewrite Intro according Feedback

@@
 
-\newcommand{\version}{v0.2 Draft}
+\newcommand{\version}{v0.4 Draft}
 
@@
 
-An application providing a user interface (\emph{UI}) is a reactive application (\emph{RA}): It handles continuos flow of user and system events in order to give the user a representation of the applications state as timely as possible.
-One way of implementing reactive behavior is the Observer design pattern \cite{gamma1995design} where an observer (e.g.
+An application providing a user interface (\emph{UI}) can be seen as a reactive application (\emph{RA})\cite{7827078}: It handles a continuos flow of user- and system-created events in order to give the user a representation of the applications state in a timely fashion.
+One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design} where an observer (e.g.
 the UI rendering code) is repeatedly informed about changed application state.
-Different ways to implement a RA were developed since the emergence of the Observer pattern as well.
-Reactive Programming (\emph{RP}) enabled through language extensions, runtime environemnts and domain specific languages (\emph{DSL}) is based on functional programming (\emph{FP}) principles and allows to describe dependencies and transformations within a RA declerativly.
+With Reactive Programming (\emph{RP}), another way to implement RAs is trending.
+RP is enabled by language extensions, runtime environments and domain specific languages (\emph{DSL}) and based on functional programming (\emph{FP}) principles.
+With RP, software engineers describe declaratively how data flows through a RA, how that data depends from each other and how it is to be transformed over time.
 
-Debugging tools are essential to software engineers in order to get deeper understanding of the runtime behavior of a software system.
-Engineers use them to understand why particular circumstances lead to unexpected failures during execution or to step through unknown source code getting themselves accustomed with it.
-IDEs as well as the developer tools of modern web browsers\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} provide battle-tested debugger utilities to inspect code.
-They let engineers step from program statement to statement, stop execution upon hitting a breakpoint and give insight on the values assigned to variables.
+Utilities to debug the runtime behavior of a program are essential to software engineers today.
+They help to understand the inner workings of unknown source code or allow to track down a problem which may lead to an unexpected failure in a particular situation.
+Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, software engineers have access to versatile debugging tools at present day.
 
-Todays debugging tools can not understand abstractions introduced by RP libraries.
+Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs.
 E.g.
-where the engineer can use the stack trace to track a problem to its root cause in a traditional imperative program, this does not work when an RP environment is used: The debugger does not have any knowledge of the underlying data flow graph which would allow a similar backtracking from graph node to node.
+when an engineer wants to know which part of the program called a particular function, the stack trace gives a clear answer to this question.
+Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a stack trace: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
 
-We will elaborate on such challenges in the field of RxJS-based frontend web applications and search answers to the following two research questions:
+This lack specific knowledge about RP concepts in traditional debugging utilities does often end in time consuming debugging sessions as we can confirm from our own experience.
+Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
+We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs using RxJS for frontend web applications.
+On the way, we will answer following two research questions:
 
 \begin{itemize}
-	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS based code?}
+	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS-based code?}
 	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS-based code be improved?}
@@
 
-Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give us insight in the challenges they face in their day-to-day jobs.
+Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
 We will evaluate and validate three cases extracted from their statements in order to answer \emph{RQ1}.
 
-Later, we propose three solution concepts for \emph{RQ2} to resolve previously identified problems which ultimately should improve the debugging experience for RxJS-based code.
+In response to this, we will present three solution concepts for \emph{RQ2} to resolve previously identified problems which we expect to improve the debugging experience for RxJS-based code.
 
@@
 Salvaneschi et al.
-\cite{7827078} define the RA as a computer program which processes a continuos, indefinite and ever-changing flow of information.
+\cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of information.
 It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
@@
 Further, Salvaneschi et al.
-provide a set of examples of what a RA might achieve using a three-parted classification schema: A \emph{synthetic application} transforms and processes streams of information.
+provide a set of examples of what a RA might achieve using a tripartite classification schema: A (i) \emph{synthetic application} transforms and processes streams of information.
 hence multiple image conversions one after another or in parallel could be implemented as a synthetic, RA.
-Beside the modelling of pure data flows, RAs are prevalent in the area of user interfaces \emph{(UI)}: They can be identified as \emph{interactive applications} and \emph{graphical animations} here.
+Beside the modelling of pure data flows, RAs are prevalent in the area of UIs: They can be identified as (ii) \emph{interactive applications} and (iii) \emph{graphical animations} here.
 

### 2020-05-24 paper.tex
Rework Intro, again

@@
 Salvaneschi et al.
-\cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of information.
-It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.
+\cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of data.
+Think of a spreadsheet application\cite{10.1145/2501654.2501666} that links the values of cells to other cells using formulas.
+Once a cell is changed, it is the task of the application to update all dependant cells accordingly and cascade the updates throughout the model.
+This data-flow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.
 
-Further, Salvaneschi et al.
-provide a set of examples of what a RA might achieve using a tripartite classification schema: A (i) \emph{synthetic application} transforms and processes streams of information.
-hence multiple image conversions one after another or in parallel could be implemented as a synthetic, RA.
-Beside the modelling of pure data flows, RAs are prevalent in the area of UIs: They can be identified as (ii) \emph{interactive applications} and (iii) \emph{graphical animations} here.
+The classification schema by Salvaneschi et al.
+\cite{7827078} divides RAs into three types: A (i) \emph{synthetic application} transforms and processes streams of information.
+RAs are prevalent in the area of UIs further: They can be identified as (ii) \emph{interactive applications} and help to describe and implement (iii) \emph{graphical animations}.
+
+\subsection{Reactive Programming}
 
 Various ways of implementing RAs emerged over time.
-The Observer design pattern \cite{gamma1995design}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{alabor:2019:reactiveappllications}.
+The Observer design pattern \cite{gamma1995design}, originated in Object Oriented Programming \emph{(OOP)}, is a pattern we can find in many of todays API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior.
 Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
-The declarativeness of FP finally lead to what we understand under ``Reactive Programming'' today.
-
-\subsection{Reactive Programming}
 
@@
 The languages runtime system then figures out a way to satisfy that given description.
-RP specifically allows the description of a dependency graph of values and transformations over time.
+RP environments specifically provide abstractions to describe a data-flow and dependency graph which is used to transform and direct data at runtime.
 
 An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a DSL\cite{10.1145/2577080.2577083}.
-At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.
+At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent application state is maintained.
 
@@
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
-Manual control of the programs execution allows to inspect the programs behavior step by step as well as assigning new values to variables ``on-the-fly''.
+Manual control of the program execution allows to inspect its behavior step by step as well as assigning new values to variables ``on-the-fly''.
 
-Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the greater model of the dependency graph.
-In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph.
-Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz} exploring this field.
-They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.
+Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the data-flow graph.
+In fact, RP provides its own challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g.
+the predecessor transformation according to the data-flow graph.
+Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
+A \textbf{reactive debugger} has knowledge about the internal model of a RP runtime system.
+It leverages on it and provides specialized tools e.g.
+to navigate or visualize the data-flow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
+We will have a closer look to available reactive debugger solutions.
 
-Traditional as well as a reactive debuggers work with the very current state of the programs execution only.
+The traditional as well as the reactive debugger works with the \emph{current} state of a programs execution only.
 They lack knowledge about what happened before or what is going to happen in the future.
@@
 Instead, it records runtime telemetry and provides an interface for later inspection.
-Engineers can ``time travel'' back and fore through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
+Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
 This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows.
-As a result, user experience (\emph{UX}) concepts need to be aligned for the engineers being able to interact with the sheer amount of information efficiently.
+As a result, user experience (\emph{UX}) concepts need to be aligned for the engineers being able to interact with the sheer amount\cite{5287015} of information efficiently.
 
@@
 
-Reactive Extensions is an open source project providing an API description for RP concepts.
-It describes itself as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming'' \cite{reactivex}.
-The API specification introduces the concept of the  \mintinline{TypeScript}{Observable} at its core.
-An \mintinline{TypeScript}{Observable} has the ability to be composed with other \mintinline{TypeScript}{Observable}s and eventually describe the dependency graph known in RP.
-A multitude of operators allow the transformation of values and \mintinline{TypeScript}{Observable}s within the graph.
+``Reactive Extensions'' (\emph{ReactiveX}) is an open source project.
+Its members and contributors defined a generic description of an RP API.
+They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}.
+ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
+The core concept of the API specification is the \mintinline{TypeScript}{Observable}\footnote{There is no known relation between ReactiveX' concept of the \mintinline{TypeScript}{Observable} and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An \mintinline{TypeScript}{Observable} can be composed and linked (``piped'') with other \mintinline{TypeScript}{Observable}s.
+It has the ability to push (``emit'') values to its dependents.
+Multiple depending \mintinline{TypeScript}{Observable}s form and describe a data-flow graph eventually.
+A multitude of operators allow the transformation of values and (higher-order) \mintinline{TypeScript}{Observable}s.
+Values emitted from an \mintinline{TypeScript}{Observable} can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
-The ReactiveX community works on a variety of reference implementations\footnote{\url{http://reactivex.io/languages.html}}: Libraries for Java, C\#, JavaScript and many more allow to work with similar API semantics on different software architecture levels.
-
-RxJS is the reference implementation for JavaScript.
-Its current major version 6 is completely done in TypeScript\footnote{\url{https://www.typescriptlang.org/}} and provides the core elements of the ReactiveX API specification.
-Large projects like Angular embed and use RxJS\cite{angualrrxjs}.
-The Listing \ref{lst:rp-with-rxjs} shows a short example of RP using RxJS.
+RxJS is the reference implementation of the ReactiveX API specification for JavaScript and TypeScript.
+It is used thoroughly by large projects like Angular\cite{angualrrxjs}.
+Listing \ref{lst:rp-with-rxjs} shows a simple example of RP using RxJS in TypeScript.
 
@@
 	\begin{minted}{TypeScript}
-		of('hello', 'foo', 'world').pipe(
-			filter(s => s !== 'foo'),
-			map(s => s.toUpperCase())
-		).subscribe(console.log)             // HELLO, WORLD
+		import { of } from 'rxjs';
+		import { filter, map } from 'rxjs/operators';
+
+		of('hello', 'foo', 'world').pipe(  // Create Observable
+			filter(s => s !== 'foo'),        // filter Operator
+			map(s => s.toUpperCase())        // map Operator
+		).subscribe(console.log)           // Emits: HELLO, WORLD
 	\end{minted}
-	\caption{RxJS Code Example}
+	\caption{Simple RxJS Example}
 	\label{lst:rp-with-rxjs}
@@
 
-\subsection{Interview Questions}
+\subsection{Interview Agenda}
 \label{sec:interviewquestions}

### 2020-05-24 paper.tex
Rework Start of Interviews

@@
 	\par
-	\Huge{Debugging of Modern RxJS-based Frontend Web Applications}
+	\Huge{Debugging of RxJS-based Frontend Web Applications}
 	\par
@@
 
+We have worked together with 15 professional software engineers during the process of answering RQ1.
+On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people which understand themselves as users of RP and related technologies.
+E.g.
+even though Angular makes heavy use of RxJS, we will see that many engineers do not interact with its abstractions directly in order to build simple UIs.
+In the end we were able to gather interesting statements nonetheless which we are going to present in the upcoming section.
+
 \subsection{Informal Interviews}
 
-In order to prepare to give an answer on the first research question, we started with a series of informal interviews.
-We talked to professionals about their experience in working with RP in their daily jobs.
-We did not limit the selection of our interview partners by the tech stack of including RxJS since we were interested in as many different perspectives on the topic as possible in this early stage.
-Eventually, we interviewed six engineers: Four work or worked more recently with Angular\footnote{\url{https://angular.io/}} and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend applications.
-Two of them stated to have more explicit experience with RxJS further.
-One subject works currently with a combination of akka\footnote{\url{https://akka.io/}} and akka-streams in Scala to create backend applications.
-The sixth subject was the creator of rxfiddle.
-The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.
+Informal interviews allowed us to gain first insights in how software engineers work with RP in their daily jobs.
+We talked to six professionals and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
+
+Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
+The fifth partner was the creator of rxfiddle, a visualization tool for RxJS \mintinline{TypeScript}{Observable}s and with this, proficient RxJS user.
+Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}-based, reactive API layer serving a web frontend application.
 
@@
 
-\section{Appendix}
-
-\subsection{Interview Agenda}
-\label{sec:interviewquestions}
-
-\begin{itemize}
-	\item When did you last apply RP?
-	\item What technology did you use RP with?
-	\item What was the best experience you had with RP?
-	\item \dots and what was the worst? \dots Can you elaborate on the concrete situation?
-\end{itemize}
-
 \bibliographystyle{splncs04}

### 2020-05-24 paper.tex
Rewrite Interviews

@@
 
-Informal interviews allowed us to gain first insights in how software engineers work with RP in their daily jobs.
-We talked to six professionals and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
+We organized informal interviews which allowed us to gain first insights in how software engineers work with RP in their daily jobs.
+We talked to six engineers and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
 
 Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
-The fifth partner was the creator of rxfiddle, a visualization tool for RxJS \mintinline{TypeScript}{Observable}s and with this, proficient RxJS user.
-Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}-based, reactive API layer serving a web frontend application.
+By being the creator of a visualizer for RxJS \mintinline{TypeScript}{Observable}s, the fifth interviewee was a proficient RxJS user.
+Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
-We identified several recurring topics throughout all the interviews.
-All of the subjects noted that they think of RP as an invaluable tool to describe and model dependencies between multiple data sources.
-Most of them appreciate how RP allows them to implement data transformations in a concise and composable way.
-The complexity of concurrent programming can be hidden well behind a simple DSL one subject noted further.
+All of the interviewees referred to RP with an overall positive experience.
+We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
+Hence a big strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL.
+The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.
 
-For the implementation, all subjects noted that a language with a strong type system is to be preferred: The compiler gives the guarantee of basic code correctness which is why e.g.
-TypeScript is used by all the subjects working with RxJS and Angular.
+It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
+E.g.
+when using ngrx for state management, ``the framework takes care that you do not have to interact with Observables on your own, but leverages greatly on their potential behind the scenes''.
+As soon as our interviewees had to extend built-in functionalities with their own features, e.g.
+a new effect\footnote{\url{https://ngrx.io/guide/effects}}, all of them valued the possibility to interact with underlying \mintinline{TypeScript}{Observable}s highly.
 
-None of the subjects remembered debugging RP code to be a positive experience, independent of the technology used.
-All of the subjects use traditional debugging tools.
-Redux\footnote{\url{https://redux.js.org/}} DevTools \footnote{\url{https://github.com/reduxjs/redux-devtools}} is in use by all four subjects working with ngrx.
-There was no mention of any more specific debugging tool in regards of akka and akka-streams.
-
-One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented.
-(ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{https://rxjs.dev/api/operators/tap}} side effect operator to log values.
-A less manually invasive way would be preferable as stated.
+When asked explicitly about what they dislike the most about RP, all interview partners with no exception emphasized the debugging process of an RP application as a big nuisance.
+The fact that our interviewees remembered the search for a bug as something negative did not surprise us, hence a bug is usually something negative afflicted.
+What did alarm us though was the fact that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' were prevalent and gave insight in \emph{why} our interview partners dislike RP debugging in particular.
+Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless.
+Further, one single partner mentioned marble diagrams as very valuable in order to understand how an RxJS \mintinline{TypeScript}{Observable} works, no matter if during development or when debugging code.
+Marble diagrams are a way to visualize and describe the value-emitting behavior of an \mintinline{TypeScript}{Observable} over time.
+They are accepted throughout the ReactiveX community for documentation\cite{marblediagrams} and testing reasons\cite{marbletesting}.
 
@@
 The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s.
-An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time.
-They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.
+An additional marble diagram for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.
 

### 2020-05-25 paper.tex
Start Rework War Stories, Add Marble Diagram

@@
 \usepackage{minted}
+\usepackage{graphicx}
 
@@
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
-The core concept of the API specification is the \mintinline{TypeScript}{Observable}\footnote{There is no known relation between ReactiveX' concept of the \mintinline{TypeScript}{Observable} and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An \mintinline{TypeScript}{Observable} can be composed and linked (``piped'') with other \mintinline{TypeScript}{Observable}s.
+The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables.
 It has the ability to push (``emit'') values to its dependents.
-Multiple depending \mintinline{TypeScript}{Observable}s form and describe a data-flow graph eventually.
-A multitude of operators allow the transformation of values and (higher-order) \mintinline{TypeScript}{Observable}s.
-Values emitted from an \mintinline{TypeScript}{Observable} can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+Multiple depending Observables form and describe a data-flow graph eventually.
+A multitude of operators allow the transformation of values and (higher-order) Observables.
+Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
@@
 Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
-By being the creator of a visualizer for RxJS \mintinline{TypeScript}{Observable}s, the fifth interviewee was a proficient RxJS user.
+By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user.
 Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
@@
 As soon as our interviewees had to extend built-in functionalities with their own features, e.g.
-a new effect\footnote{\url{https://ngrx.io/guide/effects}}, all of them valued the possibility to interact with underlying \mintinline{TypeScript}{Observable}s highly.
+a new effect\footnote{\url{https://ngrx.io/guide/effects}}, all of them valued the possibility to interact with underlying Observables highly.
 
@@
 Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless.
-Further, one single partner mentioned marble diagrams as very valuable in order to understand how an RxJS \mintinline{TypeScript}{Observable} works, no matter if during development or when debugging code.
-Marble diagrams are a way to visualize and describe the value-emitting behavior of an \mintinline{TypeScript}{Observable} over time.
+Further, one single partner mentioned marble diagrams as very valuable in order to understand how an RxJS Observable works, no matter if during development or when debugging code.
+Marble diagrams are a way to visualize and describe the value-emitting behavior of an Observable over time.
 They are accepted throughout the ReactiveX community for documentation\cite{marblediagrams} and testing reasons\cite{marbletesting}.
+
+\begin{figure}[H]
+	\centering
+	\input{figures/marble-diagram.pdf_tex}
+	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-step-controls}}
+	\label{fig:marble-diagram}
+\end{figure}
 
@@
 
-Based on the results of the interviews, we compiled a request for more specific factual reports.
-We asked professional software engineers about their most recent debugging experience in regards of RxJS, either good or bad.
-We were able to gather responses from one RxJS core team member, two Google Developer Experts and two engineers building web and mobile applications for a fintech company eventually.
+After we built a first intuition for how software engineers work with RP by evaluating the informal interviews, we were interested in more RxJS-specific, hands-on experiences.
+We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
+After reaching out, we were able to collect responses from one RxJS core team member, two Angular Google Developer Experts and two engineers building web and mobile applications using React and RxJS which includes the author of this paper.
+
+``Mixing productive and test code was something I really did not like.''
+
+``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create  smaller working examples and iterate on them.''
+
+``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all.''
+
+``starting with console.logs and later also using tags from rxjs-spy.''
 
@@
 
-It is crucial to the engineers to \textbf{understand data flows}  within the \mintinline{TypeScript}{Observable} graph.
+It is crucial to the engineers to \textbf{understand data flows}  within the Observable graph.
 Questions like ``Which source emitted a value at what time?'' or ``What is the result of a specific transformation?'' are common in this context.
@@
 		  map(i => i * 2),        // Breakpoint on i * 2
-		  map(i => i - 1),
-		  take(2)
-		).subscribe(console.log); // -1, 1
+		  map(i => i + 1),
+		  take(4)
+		).subscribe(console.log); // 1, 3, 5, 7 & complete
 	\end{minted}
-	\caption{Example RxJS Stream With Breakpoint on Line 2 Within the Arrow Function}
+	\caption{RxJS Observable With Breakpoint Within a \mintinline{TypeScript}{map} Operator}
 	\label{lst:rxjs-step-controls}
@@
 The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g.
-when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected.
-The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s.
-An additional marble diagram for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.
+when an Observable is created or subscribed) which get collected.
+The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables.
+An additional marble diagram for every Observable is used to display the values it emits over time.
 
@@
 rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
-Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined.
+Instead of adding log traces at debug-time, ``tagged'' Observables are created right at the moment the Observable itself is defined.
 A tag as shown in Listing~\ref{lst:rxjs-spy-tag} does not influence runtime behavior at first.
@@
 
-Later, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all.
+Later, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any Observable code at all.
 It is worth to mention that manual log traces, without additional effort, only show values flowing by.
@@
 Additional features are available through the debuggers console interface.
-A tagged \mintinline{TypeScript}{Observable} can be paused so it collects incoming values.
+A tagged Observable can be paused so it collects incoming values.
 They can then be emitted one after another or resumed at once.
@@
 
-The visualizer helps to understand the behavior of a specific part of an \mintinline{TypeScript}{Observable} stream in a sandboxed environment.
+The visualizer helps to understand the behavior of a specific part of an Observable stream in a sandboxed environment.
 

### 2020-05-26 paper.tex
Start with War Story Rewrite

@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect responses from one RxJS core team member, two Angular Google Developer Experts and two engineers building web and mobile applications using React and RxJS which includes the author of this paper.
+After reaching out, we were able to collect five responses: One by a RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.
 
-``Mixing productive and test code was something I really did not like.''
+The reports provided us diverse perspectives on various areas of software development using RxJS.
+One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``\dots in the beginning, it was very hard to write async tests and I did not like it.
+I really disliked that once you started to test async code you were forced to pass a TestScheduler''.
+Allowing to pass the scheduler explicitly as parameter forced them to introduce code which was only necessary for testing reasons they stated further.
+With its current major version, RxJS 6 improved heavily on the  \mintinline{TypeScript}{TestScheduler}.
+The runtime environment itself can now be augmented with the scheduler which results in cleaner code.
+
+Even though the share of non-productive, testing-related code necessary to build a mature RxJS-based application seems to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
+One engineer described a scenario where they suspected a problem in a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
+On top, each data source could re-emit updated versions of previously requested information any time.
+Though tested thoroughly, after a week in production the first of many bugs was reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all'' they told us.
+The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
+``I started to inspect the flow \dots with console.logs and later also using tags from rxjs-spy'', hence manually augmenting the code with trace logs.
+After the time consuming log analysis, they finally were able resolve the bug.
+The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
+(We will elaborate on the functionality of rxjs-spy in the next section more in-depth.)
 
 ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create  smaller working examples and iterate on them.''
-
-``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all.''
-
-``starting with console.logs and later also using tags from rxjs-spy.''
 

### 2020-05-26 paper.tex
Finalize War Story Rewrite

@@
 The reports provided us diverse perspectives on various areas of software development using RxJS.
-One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``\dots in the beginning, it was very hard to write async tests and I did not like it.
+One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it.
 I really disliked that once you started to test async code you were forced to pass a TestScheduler''.
@@
 Even though the share of non-productive, testing-related code necessary to build a mature RxJS-based application seems to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
-One engineer described a scenario where they suspected a problem in a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
+One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
 On top, each data source could re-emit updated versions of previously requested information any time.
@@
 The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
-``I started to inspect the flow \dots with console.logs and later also using tags from rxjs-spy'', hence manually augmenting the code with trace logs.
+``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy'', hence manually augmenting the code with trace logs.
 After the time consuming log analysis, they finally were able resolve the bug.
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
-(We will elaborate on the functionality of rxjs-spy in the next section more in-depth.)
+(We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)
 
-``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create  smaller working examples and iterate on them.''
-
-We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex.
-Three subjects even claimed using specialized logging functionalities provided by rxjs-spy\cite{rxjsspy}.
-As the subjects stated, the knowledge about additional events made visible helped in situations where they needed to find problems within a graph composing multiple asynchronous data sources on multiple levels.
-Without it, they would have been blind to what actually happened within the system.
-
-In addition to more sophisticated logging, three subjects reported that they built a habit of decomposing complex graphs into smaller, simpler bits and running those in a dedicated environment like Rx Visualizer\cite{rxviz} or StackBlitz\footnote{\url{https://stackblitz.com/}}.
+In a related war story, another engineer disclosed similar practices using  external tools: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create smaller working examples and iterate on them.'' Stackblitz\footnote{\url{https://stackblitz.com/}} is a full JavaScript development environment available online.
+They used this sandboxed setting to run, debug and iterate on selected pieces of a larger Observable stream.
+Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
 

### 2020-05-27 paper.tex
Start with Wrap Up

@@
 We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs using RxJS for frontend web applications.
-On the way, we will answer following two research questions:
+
+Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
+We will validate their statement in a controlled experiment and answer our first research question:
 
@@
 	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS-based code?}
+\end{itemize}
+
+In response to this, we will present three potential solutions to resolve previously identified problems and give an answer on the second research question:
+
+\begin{itemize}
 	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS-based code be improved?}
@@
 
-Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
-We will evaluate and validate three cases extracted from their statements in order to answer \emph{RQ1}.
-
-In response to this, we will present three solution concepts for \emph{RQ2} to resolve previously identified problems which we expect to improve the debugging experience for RxJS-based code.
-
-The future implementation and validation of these proposals leads to a third research question \emph{RQ3} which will be part of future work not part of this paper:
+The future implementation and validation of these proposals leads to our third research question which will be part of future work not part of this paper:
 
@@
 It is used thoroughly by large projects like Angular\cite{angualrrxjs}.
-Listing \ref{lst:rp-with-rxjs} shows a simple example of RP using RxJS in TypeScript.
+Listing \ref{lst:rp-with-rxjs} shows an example of RP using RxJS in TypeScript.
 
@@
 	\end{minted}
-	\caption{Simple RxJS Example}
+	\caption{Basic RxJS Example}
 	\label{lst:rp-with-rxjs}
@@
 even though Angular makes heavy use of RxJS, we will see that many engineers do not interact with its abstractions directly in order to build simple UIs.
-In the end we were able to gather interesting statements nonetheless which we are going to present in the upcoming section.
+In the end we were able to gather interesting statements nonetheless which we are going to present in this section.
 
@@
 
-We organized informal interviews which allowed us to gain first insights in how software engineers work with RP in their daily jobs.
+We organized informal interviews which allowed us to gain first insight in how software engineers work with RP in their daily jobs.
 We talked to six engineers and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
@@
 
-Even though the share of non-productive, testing-related code necessary to build a mature RxJS-based application seems to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
+Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
 One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
@@
 
-\subsection{Summary}
+A final story described yet another variation of the application of an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
+Like in the report before, it was necessary to extracted code from the actual applications code base.
+But once done, the diagrams helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.
+
+\subsection{Wrap Up}
+
+The overall sentiment in both the informal interviews as well as the reported war stories in regards of RP with RxJS was positive.
+The software engineers value how they can describe data flows using a DSL, even though the learning curve was perceived as steep.
+We heard some interesting reports how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
+Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.
+
+Most participants considered a statically typed language like TypeScript a fundamental necessity which allowed them to implement data-flow graphs with a minimal, formal correctness.
+We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and lacking ``the right tool'' to work within the context of RP.
+
+% TODO
 

### 2020-05-28 paper.tex
Finish Wrap Up

@@
 E.g.
-when an engineer wants to know which part of the program called a particular function, the stack trace gives a clear answer to this question.
-Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a stack trace: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
+when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
+Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
 
@@
 Multiple depending Observables form and describe a data-flow graph eventually.
-A multitude of operators allow the transformation of values and (higher-order) Observables.
+A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable.
+This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables.
 Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect five responses: One by a RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.
+After reaching out, we were able to collect five responses: One by an RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.
 
-The reports provided us diverse perspectives on various areas of software development using RxJS.
+The reports provided us with diverse perspectives on various areas of software development using RxJS.
 One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it.
@@
 The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
-``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy'', hence manually augmenting the code with trace logs.
+``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs.
 After the time consuming log analysis, they finally were able resolve the bug.
@@
 
-Most participants considered a statically typed language like TypeScript a fundamental necessity which allowed them to implement data-flow graphs with a minimal, formal correctness.
-We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and lacking ``the right tool'' to work within the context of RP.
-
-% TODO
-
-The interpretation of the combined interview and war stories data gives us an idea about two main challenges software engineers have to overcome during debugging activities of RxJS-based code:
-
-It is crucial to the engineers to \textbf{understand data flows}  within the Observable graph.
-Questions like ``Which source emitted a value at what time?'' or ``What is the result of a specific transformation?'' are common in this context.
-Once a program fails, software engineers need to \textbf{rely on traditional debugger tools}.
-Though providing help to some extent, the tools lack of specific knowledge about the RP runtime environment adds some oddities to the debugging process: The (i) functionalities allowing step-wise navigation through the code does not respect the RP context.
-Assume we have set a breakpoint at \mintinline{TypeScript}{i * 2} on line~2 of Listing~\ref{lst:rxjs-step-controls}.
-Once the debugger hits the breakpoint and we instruct it to ``Step Over'', the debugger will lead us to the internal implementation of the \mintinline{TypeScript}{map} operator instead to line~3.
-(ii) Stack traces include information about the invocation of internal functions which, most of the time, do not provide any valuable information to the engineers further.
+Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement data-flow graphs with minimal, formal correctness.
+We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though.
+Yet traditional debugging tools allowed them to inspect various parts of a stream, they were unable to get more information about e.g.
+life cycles of the Observable composition.
+Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem.
+Listing~\ref{lst:rxjs-debugging} exemplifies challenges when debugging a stream of Observables with traditional debugging tools.
 
@@
 		interval(1000).pipe(    // Increasing numbers with 1s delay
-		  map(i => i * 2),        // Breakpoint on i * 2
-		  map(i => i + 1),
-		  take(4)
-		).subscribe(console.log); // 1, 3, 5, 7 & complete
+		  map(i => i * 2),      // Breakpoint possible within i * 2
+		  take(4),              // No breakpoint possible
+		  tap(console.log)      // Manually added log statement
+		).subscribe(showValue); // 0, 2, 4, 6
 	\end{minted}
-	\caption{RxJS Observable With Breakpoint Within a \mintinline{TypeScript}{map} Operator}
-	\label{lst:rxjs-step-controls}
+	\caption{Debugging of an RxJS Observable using breakpoints and log statements. \mintinline{TypeScript}{showValue} renders an emitted value to the UI.}
+	\label{lst:rxjs-debugging}
 \end{listing}
+
+Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3.
+One would need to place the breakpoint within the operators internal implementation instead which can be cumbersome in case the operator is used in a different stream as well.
+Once the breakpoint on Line~2 interrupts the programs execution, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
+
+\begin{listing}[H]
+	\begin{minted}{TypeScript}
+		<anonymous> RxJS
+			rxjs 6.5.2/internal/operators/map.js:49
+			rxjs 6.5.2/internal/Subscriber.js:66
+			rxjs 6.5.2/internal/observable/interval.js:23
+			rxjs 6.5.2/internal/scheduler/AsyncAction.js:71
+			// ...
+	\end{minted}
+	\caption{Call Stack for Arrow Function on Line~2 in Listing~\ref{lst:rxjs-debugging}}
+	\label{lst:rxjs-call-stack}
+\end{listing}
+
+We further learned that log statements as shown on Line~4 in Listing~\ref{lst:rxjs-debugging} indeed help to trace emitted values, but do not expose any information about the life cycle of an Observable.
+The reporter describing their usage of rxjs-spy explained, that the knowledge about when an Observable gets subscribed, unsubscribed or completed helped a lot in order to solve problems related to higher-order Observables.
+
+Finally, we understood that if a problem is hard to replicate within the actual application, the engineers use external sandbox development environments to isolate specific parts of an Observable composition.
+This allows them to iterate on it faster than it would be possible otherwise.
+
+After the evaluation of all reports and interviews it is our speculation that software engineers are truly lacking debugging tools which ``understand'' RP concepts provided by RxJS.
+Though traditional debuggers can help up to some extent, they currently cannot answer all questions an engineer has as we saw repeatedly hence making manual modifications on source code necessary.
+We will validate this assumption after we took a look on previous efforts on the area of reactive debugging.
 

### 2020-05-28 paper.tex
Rework Tools

@@
 \usepackage{graphicx}
+\usepackage{booktabs}
 
@@
 
-After the evaluation of all reports and interviews it is our speculation that software engineers are truly lacking debugging tools which ``understand'' RP concepts provided by RxJS.
-Though traditional debuggers can help up to some extent, they currently cannot answer all questions an engineer has as we saw repeatedly hence making manual modifications on source code necessary.
-We will validate this assumption after we took a look on previous efforts on the area of reactive debugging.
+After the evaluation of all reports and interviews it is our speculation that software engineers are truly lacking debugging tools which ``understand'' RP concepts provided through RxJS.
+Even though traditional debuggers might help up to some extent, they do not provide all the information an engineer is interested in.
+The last resort in this situation is often the manual augmentation, modification and extraction of source code as we saw repeatedly.
+After a look at previous efforts in the area of reactive debugging, we are going to validate our assumption in Section~\ref{sec:study}.
 
@@
 
-Our work is not the first approach on understanding and improving on the struggles software engineers have to overcome when working on RP code, either with or without RxJS.
-We are going to present four of this tools in this section.
-This list is not meant to be exhaustive.
-It contains tools we either got to know during our research because they were mentioned in one of the interviews or we worked with them personally in the past.
+Our work is not the first attempt to understand and improve on the struggles software engineers have to overcome when working on RP code, either with or without RxJS.
+We are going to present a selection of four utilities whereas two of them have an academic background.
+We either learned about them during our research because they were mentioned in one of the interviews and war stories or because we have worked with them personally in the past.
 
@@
 did focus their research heavily on RP during the mid 2010s.
-Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well.
+Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well.
 In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al.
 elaborate on the necessity of RP debugging tools which can understand the specific language abstractions.
-With the plugin Reactive Inspector for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
+With \emph{Reactive Inspector} plugin for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
 One of its main features is the visualization of the internal dependency graph.
@@
 
-rxfiddle proposed by Banken et al.
+\emph{rxfiddle} as proposed by Banken et al.
 is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
@@
 Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
-rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator.
-Instead of adding log traces at debug-time, ``tagged'' Observables are created right at the moment the Observable itself is defined.
-A tag as shown in Listing~\ref{lst:rxjs-spy-tag} does not influence runtime behavior at first.
+rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator along with sophisticated log monitoring facilities.
+Ideally, ``tagged'' Observables are created during development when the data-flows are composed for the first time.
+Like tha \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.
 
@@
 	\end{minted}
-	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line 3}
+	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line~3}
 	\label{lst:rxjs-spy-tag}
@@
 
-Later, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any Observable code at all.
-It is worth to mention that manual log traces, without additional effort, only show values flowing by.
+Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any Observable code at all.
 The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.
@@
 
-The visualizer helps to understand the behavior of a specific part of an Observable stream in a sandboxed environment.
+The visualizer helps to understand the behavior of a specific part of an Observable stream in a sandboxed environment as a reported war story stated.
 
@@
 Most of them provide a way to represent the data flow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy).
-They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in the visual graph representation (Reactive Inspector).
-Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-based tools have a more specific feature set: They usually work outside of an IDE and/or need manual code instrumentation like rxjs-spy.
+They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector).
+Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-related tools provide small, differentiated feature sets.
+Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).
 
-\textbf{TODO Compare im Table?}
+\begin{table}[]
+	\begin{tabular}{@{}lllll@{}}
+	\toprule
+					   & \textbf{UI} & \textbf{Integration}                                               & \textbf{Flow Control}                                              & \textbf{RP Runtime} \\ \midrule
+	Reactive Inspector & Visual      & Eclipse IDE                                                        & Full                                                               & REScala             \\ \midrule
+	rxfiddle           & Visual      & Web App                                                            & \begin{tabular}[c]{@{}l@{}}Visualization\\ Only\end{tabular}       & RxJS                \\ \midrule
+	rxjs-spy           & Textual     & \begin{tabular}[c]{@{}l@{}}Browser Console\\and Code\end{tabular} & \begin{tabular}[c]{@{}l@{}}Limited (No\\ Breakpoints)\end{tabular} & RxJS                \\ \midrule
+	Rx Visualizer      & Visual      & Web App                                                            & \begin{tabular}[c]{@{}l@{}}Visualization\\ Only\end{tabular}       & RxJS                \\ \bottomrule
+	\end{tabular}
+	\caption{Feature Matrix for Discussed RP Debugging Tools}
+	\label{tab:compare-tools}
+	\end{table}
 

### 2020-05-28 paper.tex
Start Validation

@@
 Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes two recurrent challenges.
-With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of reactive applications as well as specific tooling.
+With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of RAs as well as specific tooling.
 We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.
@@
 
-Reactive programming \emph{(RP)} is a declarative programming paradigm.
+RP is a declarative programming paradigm.
 Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved.
@@
 
-\section{Study Design}
+\section{Validation of Interviews and War Stories}
 \label{sec:study}
 
+\subsection{Study Design}
+
 \begin{itemize}
-	\item Hypothesis
-	\item Use prev section to build experiments using existing tooling
-	\item Validate stories and interviews
+	\item \emph{Hypothesis: If frontend engineers are given an RxJS-based problem, then they will modify the code in order to understand its behavior.}
 \end{itemize}

### 2020-05-29 paper.tex
Add CH Reference
Study Desgin

@@
 		\vspace{2ex}
-		HSR University of Applied Sciences Rapperswil\\
+		HSR University of Applied Sciences Rapperswil, Switzerland\\
 		\par
@@
 Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
-We will validate their statement in a controlled experiment and answer our first research question:
+We will validate their statement in an observational study and answer our first research question:
 
@@
 
+We explored on how engineers use RP and what techniques they might apply for debugging problems earlier.
+We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
+It is further our own experience that this practice is often not a productive one since the evaluation of trace information, once available, can be cumbersome and time consuming.
+In addition, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
+We identified this circumstance as one of the main debugging practices when software engineers work with RxJS-based code.
+
+This is why we saw the need in validating earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate following hypothesis:
+
+\begin{itemize}
+	\item \emph{Hypothesis: If software engineers are given an RxJS-based problem, then they will modify the code in order to understand its behavior.}
+\end{itemize}
+
 \subsection{Study Design}
 
-\begin{itemize}
-	\item \emph{Hypothesis: If frontend engineers are given an RxJS-based problem, then they will modify the code in order to understand its behavior.}
-\end{itemize}
+The subjects for our study were required to have experience in developing applications with RxJS.
+We did not distinguish between frontend or backend engineers, hence in the end were able to find four subjects willing to participate in our experiment.
+We were interested to see how the subjects apply debugging techniques they would use in a common situation at their jobs.
+This is why we decided to conduct the experiment in a rather uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
+Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention the hypothesis by intention.
+
+We planned to have a one hour session for the actual experiment with each subject, followed-up by a short, unattended after-action survey.
+The experiment was executed in two consecutive blocks of 25 minutes each.
+We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of the experiment.
+Each of these applications were rigged with two to three bugs which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
+A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
+We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process.
+We refrained to answer any question related to the ``where'' a bug has to be expected though.
+
+A week before the actual experiment, we sent out the participant briefing document to all of our subjects.
+We outlined the course of action and provided them with an example ZIP file.
+This file contained the same setup as the file provided at the experiment and allowed the subjects to get accustomed with things like starting the web applications or running the test suites.
+
+We decided to monitor our subjects progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
+This allowed us to record the sessions with relative low technical effort for later evaluation.
+
+The quantitative after-action survey provided after the main part of the study was provided within 24 hours after a subjects participation.
+We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
+
+\subsection{Study Results}
 

### 2020-05-29 paper.tex
Add Results

@@
 
-We planned to have a one hour session for the actual experiment with each subject, followed-up by a short, unattended after-action survey.
+We planned to have a one hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
 The experiment was executed in two consecutive blocks of 25 minutes each.
@@
 We decided to monitor our subjects progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
-This allowed us to record the sessions with relative low technical effort for later evaluation.
+This allowed us further to record the sessions with relative low technical effort for later evaluation.
 
-The quantitative after-action survey provided after the main part of the study was provided within 24 hours after a subjects participation.
+The after-action survey was provided within 24 hours after a subjects participation in the main part of the study.
 We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
+The respective answers allowed us to put the observed actions into perspective and detect potential irregularities in case a subject acted differently as they would have in a ``real'' situation.
 
-\subsection{Study Results}
+\subsection{Results}
+
+After the subjects got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
+Two of them used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
+
+For both problems given, all subjects could be observed to make use of manually added trace log statements which they either added to existing arrow functions and/or using the \mintinline{TypeScript}{tap} operator.
+Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
+After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
+
+The responses on the survey showed that 75\% of the subject population had two or more years of experience with RxJS.
+Where all of them use RxJS to develop frontend applications, one subject declared having RxJS used for backend development as well.
+When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser and/or IDE.
+Two of them leverage additional tracing functionality of rxjs-spy and all of them use manual log statements.
+
+\subsection{Interpretation}
 

### 2020-05-29 paper.tex
Interpretation

@@
 Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
-We will validate their statement in an observational study and answer our first research question:
+We will validate their statements in an observational study and answer our first research question:
 
@@
 	Reactive Inspector & Visual      & Eclipse IDE                                                        & Full                                                               & REScala             \\ \midrule
-	rxfiddle           & Visual      & Web App                                                            & \begin{tabular}[c]{@{}l@{}}Visualization\\ Only\end{tabular}       & RxJS                \\ \midrule
-	rxjs-spy           & Textual     & \begin{tabular}[c]{@{}l@{}}Browser Console\\and Code\end{tabular} & \begin{tabular}[c]{@{}l@{}}Limited (No\\ Breakpoints)\end{tabular} & RxJS                \\ \midrule
-	Rx Visualizer      & Visual      & Web App                                                            & \begin{tabular}[c]{@{}l@{}}Visualization\\ Only\end{tabular}       & RxJS                \\ \bottomrule
+	rxfiddle           & Visual      & Web App                                                            & None       & RxJS                \\ \midrule
+	rxjs-spy           & Textual     & \begin{tabular}[c]{@{}l@{}}Browser Console\\and Code\end{tabular} & Limited & RxJS                \\ \midrule
+	Rx Visualizer      & Visual      & Web App                                                            & None       & RxJS                \\ \bottomrule
 	\end{tabular}
@@
 
+We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs at some point.
+All of them used the produced information in order to track down problems in the application.
+The the extraction to and reintegration from an external tool was not observed, though would have been yet another form of code modification.
+
+For any subject, we could not identify any contradiction between the observed behavior and the answers given within the after-action survey.
+
+All reviewed subjects exhibited the debugging behavior described in our initial hypothesis which we can accept based on the collected evidence.
+
 \section{Future Work}

### 2020-05-29 paper.tex
Answer RQ1

@@
 when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
+Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
 
@@
 Once a cell is changed, it is the task of the application to update all dependant cells accordingly and cascade the updates throughout the model.
-This data-flow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.
+This dataflow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.
 
@@
 The languages runtime system then figures out a way to satisfy that given description.
-RP environments specifically provide abstractions to describe a data-flow and dependency graph which is used to transform and direct data at runtime.
+RP environments specifically provide abstractions to describe a dataflow and dependency graph which is used to transform and direct data at runtime.
 
@@
 
-Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the data-flow graph.
+Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the dataflow graph.
 In fact, RP provides its own challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g.
-the predecessor transformation according to the data-flow graph.
+the predecessor transformation according to the dataflow graph.
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
@@
 It leverages on it and provides specialized tools e.g.
-to navigate or visualize the data-flow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
+to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
 We will have a closer look to available reactive debugger solutions.
@@
 It has the ability to push (``emit'') values to its dependents.
-Multiple depending Observables form and describe a data-flow graph eventually.
+Multiple depending Observables form and describe a dataflow graph eventually.
 A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable.
@@
 We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
-Hence a big strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL.
+Hence a big strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
 The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.
@@
 The overall sentiment in both the informal interviews as well as the reported war stories in regards of RP with RxJS was positive.
-The software engineers value how they can describe data flows using a DSL, even though the learning curve was perceived as steep.
+The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep.
 We heard some interesting reports how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
@@
 
-Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement data-flow graphs with minimal, formal correctness.
+Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness.
 We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though.
@@
 Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools.
-Most of them provide a way to represent the data flow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy).
+Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy).
 They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector).
@@
 
-\section{Future Work}
+Talking to professionals during the informal reviews, evaluating the war stories as well as the hands-on experiences and results from the observational study allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application: (i) They want to understand how and what data flows through the graph described by the RP DSL.
+This helps to identify simpler problems most of the time, though knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
+(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a deeper understanding of the RP runtimes internal graph-based model.
+This makes it hard for engineers to retrace dataflows and computations.
+And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to manually modify source code or extracting parts of it to an external utilities.
+These additional efforts cost time and bring their own risks since production code needs to be touched without adding actual value.
+
+\section{Outlook}
 \label{sec:future}
-
-\begin{itemize}
-	\item Pick up results from war stories and experiment
-	\item Propose one solution per challenges
-	\item Sketch how this could look like
-	\item This answers RQ2
-\end{itemize}
 
@@
 	\end{itemize}
+\end{itemize}
+
+\begin{itemize}
+	\item Pick up results from war stories and experiment
+	\item Propose one solution per challenges
+	\item Sketch how this could look like
+	\item This answers RQ2
 \end{itemize}

### 2020-05-29 paper.tex
Grammarly Run One

@@
 An application providing a user interface (\emph{UI}) can be seen as a reactive application (\emph{RA})\cite{7827078}: It handles a continuos flow of user- and system-created events in order to give the user a representation of the applications state in a timely fashion.
-One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design} where an observer (e.g.
+One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design}, where an observer (e.g.
 the UI rendering code) is repeatedly informed about changed application state.
 With Reactive Programming (\emph{RP}), another way to implement RAs is trending.
-RP is enabled by language extensions, runtime environments and domain specific languages (\emph{DSL}) and based on functional programming (\emph{FP}) principles.
+RP is enabled by language extensions, runtime environments and domain-specific languages (\emph{DSL}) and based on functional programming (\emph{FP}) principles.
 With RP, software engineers describe declaratively how data flows through a RA, how that data depends from each other and how it is to be transformed over time.
@@
 They help to understand the inner workings of unknown source code or allow to track down a problem which may lead to an unexpected failure in a particular situation.
-Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, software engineers have access to versatile debugging tools at present day.
+Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, software engineers have access to versatile debugging tools at the present day.
 
@@
 
-This lack specific knowledge about RP concepts in traditional debugging utilities does often end in time consuming debugging sessions as we can confirm from our own experience.
+This lack specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions as we can confirm from our own experience.
 Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
@@
 Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes two recurrent challenges.
-With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding debugging of RAs as well as specific tooling.
-We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.
+With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding the debugging of RAs as well as specific tooling.
+We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to the validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.
 
@@
 Salvaneschi et al.
-\cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of data.
+\cite{7827078} define a RA as a computer program which processes continuos, indefinite and ever-changing flows of data.
 Think of a spreadsheet application\cite{10.1145/2501654.2501666} that links the values of cells to other cells using formulas.
@@
 Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved.
-The languages runtime system then figures out a way to satisfy that given description.
+The language runtime system then figures out a way to satisfy that given description.
 RP environments specifically provide abstractions to describe a dataflow and dependency graph which is used to transform and direct data at runtime.
 
-An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a DSL\cite{10.1145/2577080.2577083}.
+An RP environment provides (i) language constructs to describe mentioned graphs, often in the form of a DSL\cite{10.1145/2577080.2577083}.
 At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent application state is maintained.
@@
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
-A \textbf{reactive debugger} has knowledge about the internal model of a RP runtime system.
+A \textbf{reactive debugger} has knowledge about the internal model of an RP runtime system.
 It leverages on it and provides specialized tools e.g.
@@
 
-The traditional as well as the reactive debugger works with the \emph{current} state of a programs execution only.
+The traditional as well as the reactive debugger works with the \emph{current} state of a program's execution only.
 They lack knowledge about what happened before or what is going to happen in the future.
@@
 
-``Reactive Extensions'' (\emph{ReactiveX}) is an open source project.
+``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project.
 Its members and contributors defined a generic description of an RP API.
@@
 
-All of the interviewees referred to RP with an overall positive experience.
+All of the interviewees referred to RP with overall positive experience.
 We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
@@
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs.
-After the time consuming log analysis, they finally were able resolve the bug.
+After the time-consuming log analysis, they finally were able resolve the bug.
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
@@
 Like in the report before, it was necessary to extracted code from the actual applications code base.
-But once done, the diagrams helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.
+But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.
 
@@
 Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3.
-One would need to place the breakpoint within the operators internal implementation instead which can be cumbersome in case the operator is used in a different stream as well.
-Once the breakpoint on Line~2 interrupts the programs execution, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
+One would need to place the breakpoint within the operators' internal implementation instead which can be cumbersome in case the operator is used in a different stream as well.
+Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
 
@@
 once a node retains a specific output value or an event of interest takes place in the runtime.
-Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.
+Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without re-execution.
 
@@
 
-The reference implementation supports event collection for RxJS-based code in version 4 and 5.
+The reference implementation supports event collection for RxJS-based code in versions 4 and 5.
 It can either be used on a local development machine on existing projects or by copying code snippets to the online available.
@@
 Ideally, ``tagged'' Observables are created during development when the data-flows are composed for the first time.
-Like tha \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.
+Like the \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.
 
@@
 We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
-It is further our own experience that this practice is often not a productive one since the evaluation of trace information, once available, can be cumbersome and time consuming.
+It is further our own experience that this practice is often not a productive one since the evaluation of trace information, once available, can be cumbersome and time-consuming.
 In addition, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
@@
 
-This is why we saw the need in validating earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate following hypothesis:
+This is why we saw the need in validating earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
 
@@
 
-We planned to have a one hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
+We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
 The experiment was executed in two consecutive blocks of 25 minutes each.
@@
 
-We decided to monitor our subjects progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
+We decided to monitor our subjects' progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
 This allowed us further to record the sessions with relative low technical effort for later evaluation.
 
-The after-action survey was provided within 24 hours after a subjects participation in the main part of the study.
+The after-action survey was provided within 24 hours after a subject's participation in the main part of the study.
 We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
@@
 All of them used the produced information in order to track down problems in the application.
-The the extraction to and reintegration from an external tool was not observed, though would have been yet another form of code modification.
+The extraction to and reintegration from an external tool was not observed, though would have been yet another form of code modification.
 
@@
 This makes it hard for engineers to retrace dataflows and computations.
-And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to manually modify source code or extracting parts of it to an external utilities.
-These additional efforts cost time and bring their own risks since production code needs to be touched without adding actual value.
+And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to manually modify source code or extracting parts of it to an external utility.
+These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
 

### 2020-05-30 paper.tex
Future Work

@@
 
-In response to this, we will present three potential solutions to resolve previously identified problems and give an answer on the second research question:
+In response to this, we will present concepts how to resolve previously identified problems and give an answer on the second research question:
 
@@
 
-We divide debugging concepts roughly into three categories: Traditional \textbf{imperative-focused debuggers} provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
+We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
@@
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
-A \textbf{reactive debugger} has knowledge about the internal model of an RP runtime system.
+A (ii) \emph{reactive debugger} has knowledge about the internal model of an RP runtime system.
 It leverages on it and provides specialized tools e.g.
@@
 This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
-An \textbf{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
+An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
 Instead, it records runtime telemetry and provides an interface for later inspection.
@@
 	\input{figures/marble-diagram.pdf_tex}
-	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-step-controls}}
+	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-debugging}.
+Each marble represents an emitted value.
+The vertical line at marble \mintinline{TypeScript}{6} indicates the Observable completed.}
 	\label{fig:marble-diagram}
@@
 
-\begin{table}[]
+\begin{table}[H]
 	\begin{tabular}{@{}lllll@{}}
@@
 
-Talking to professionals during the informal reviews, evaluating the war stories as well as the hands-on experiences and results from the observational study allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application: (i) They want to understand how and what data flows through the graph described by the RP DSL.
+Talking to professionals during the informal reviews, evaluating the war stories as well as the hands-on experiences and results from the observational study allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application, hence answering RQ1: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL.
 This helps to identify simpler problems most of the time, though knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
-(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a deeper understanding of the RP runtimes internal graph-based model.
+(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}.
 This makes it hard for engineers to retrace dataflows and computations.
-And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to manually modify source code or extracting parts of it to an external utility.
+And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
 These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
 
-\section{Outlook}
+\section{Future Work}
 \label{sec:future}
+
+Having the main challenges for engineers identified when debugging RxJS-code, we can propose areas to improve on now and resolve RQ2 by doing so.
+
+\subsection{Improve User Experience}
+
+Our overall goal is to support engineers during the process of understanding the inner workings of RxJS data-flow graph.
+Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw.
+Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly though.
+Filter and search functionalities help, but do not resolve all UX challenges here.
+
+Visualizing the data-flow graph work better to show dependencies within the model compared to textual trace logs.
+Previous research showed that the sheer amount of nodes in a complex graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other not as easy to implement efficiently since the mass of collected data grows quickly.
+Storing a representation of the graph rarely enough since engineers are interested in the values emitted over time per node as well.
+
+We intend to find a solution that combines the power of textual trace logs with the ease-of-use of a graphical visualization for RxJS data-flow graphs.
+
+\subsection{Automatic Code Augmentation}
+
+RP code needs to be augmented with additional functionality so telemetry data can be collected.
+This happens using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156} but especially in the area of RxJS often manually.
+We want to make the manual modification of production source code obsolete.
+rxfiddle proofed such mechanisms to work with RxJS 4 and 5, hence we intend to bring this feature to version 6, and possibly the upcoming major version 7 as well.
+
+\subsection{Integration}
+
+75\% of our observational study's participants used the debugging tools integrated to their browser and/or IDE.
+Even though some of them stated to know additional external utilities in the after-action survey, none of them went that far and used them during the one-hour session.
+
+We understand the observed behavior as evidence for a hurdle preventing engineers from using specialized RP debugging utilities.
+In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser and or an IDE.
+
+\subsection{Other Areas of Improvement}
+
+There are many more areas which could be worth a further look in the future.
+E.g.
+the combination of controlling the programs execution based on breakpoints within the data-flow graph or concepts of omniscient debugging are interesting fields as well.
+We will focus on the previous three topics for our own future work but keep other ideas in a backlog in case time allows additional iterations on them.
 

### 2020-05-30 paper.tex
Threats to Validity

@@
 
-We have worked together with 15 professional software engineers during the process of answering RQ1.
+We have worked together professional software engineers during the process of answering RQ1.
 On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people which understand themselves as users of RP and related technologies.
@@
 We understand the observed behavior as evidence for a hurdle preventing engineers from using specialized RP debugging utilities.
-In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser and or an IDE.
+In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser and/or an IDE as possible.
 
@@
 
-\begin{itemize}
-	\item Informal interviews/war stories only
-	\begin{itemize}
-		\item Describe background of interview partners
-		\item Only 4 Participants in experiment
-		\item Code in experiment carefully crafted... Though is it representative?
-		\
-	\end{itemize}
-\end{itemize}
+We worked with 15 professional software engineers while writing this paper.
+Six interview partners and five war story reports helped us building the foundation for a later observational study with a total of four participants.
 
-\begin{itemize}
-	\item Pick up results from war stories and experiment
-	\item Propose one solution per challenges
-	\item Sketch how this could look like
-	\item This answers RQ2
-\end{itemize}
+While we claim nine interviewees and reporters to have been an appropriate number of peers for collecting the initial data basis, we are aware that our four subjects for the observational study lead to a result lacking significance.
+
+More time given, we suggest to conduct the study with a bigger population again to obtain a result with higher significance and providing more insight in how engineers debug RxJS-based code.
+Doing so would possibly require a statistical evaluation of these results.
+
+Furthermore, this was the first time we designed an observational study of this kind.
+Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless.
+E.g.
+we would use a commonly known business domain for both problems presented to the subjects.
+This would allow them to understand the functionality of the application in front of them quicker and more focus on resolving the task given.
 
@@
 \label{sec:conclusion}
-
-\begin{itemize}
-	\item Recap Past Developments in RP Debugging
-	\item Recap proposed solution
-	\item Whats next in PA2?
-	\item Whats the vision for the thesis?
-\end{itemize}
 

### 2020-05-30 paper.tex
Conclusion

@@
 
+We have talked to multiple people in order to understand how software engineers, and in particular frontend engineers, apply RP using RxJS for building applications.
+We have looked into how they debug their applications and could identify several struggles they have to overcome in this process.
+We could proof that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal data-flow graph by observing professionals in an observational study.
+
+On the way, we elaborated on previous, academical as well as non-academical efforts in the field of reactive debugging.
+We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.
+
+With \emph{UX}, \emph{Automatic Code Augmentation} and \emph{Integration} we finally proposed three concepts where we intend to improve on a software engineers debugging experience, hence answering our second research question RQ2.
+We plan to work further on these topics in an upcoming workshop paper within the scope of our master's studies research.
+
 \bibliographystyle{splncs04}

### 2020-05-30 paper.tex
Grammarly to the rescue

@@
 
-An application providing a user interface (\emph{UI}) can be seen as a reactive application (\emph{RA})\cite{7827078}: It handles a continuos flow of user- and system-created events in order to give the user a representation of the applications state in a timely fashion.
-One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design}, where an observer (e.g.
-the UI rendering code) is repeatedly informed about changed application state.
+An application providing a user interface (\emph{UI}) can be seen as a reactive application (\emph{RA})\cite{7827078}: It handles a constant flow of user- and system-created events in order to give the user a representation of the applications state in a timely fashion.
+One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design}, where an observer (e.g., the UI rendering code) is repeatedly informed about the changed application state.
 With Reactive Programming (\emph{RP}), another way to implement RAs is trending.
-RP is enabled by language extensions, runtime environments and domain-specific languages (\emph{DSL}) and based on functional programming (\emph{FP}) principles.
-With RP, software engineers describe declaratively how data flows through a RA, how that data depends from each other and how it is to be transformed over time.
+Language extensions, runtime environments and domain-specific languages (\emph{DSL}) enable RP based on functional programming (\emph{FP}) principles.
+With RP, software engineers describe how data flows through a RA, how it depends on each other and how it needs to be transformed over time in a declarative way.
 
@@
 They help to understand the inner workings of unknown source code or allow to track down a problem which may lead to an unexpected failure in a particular situation.
-Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, software engineers have access to versatile debugging tools at the present day.
+Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level; software engineers have access to versatile debugging tools at the present day.
 
 Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs.
-E.g.
-when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.
+E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
+Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor; it will lead to the internals of the RP runtime instead.
 
-This lack specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions as we can confirm from our own experience.
+This lack of specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions, as we can confirm from our own experience.
 Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
@@
 
-Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight in the challenges they face in their day-to-day jobs.
+Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight into the challenges they face in their day-to-day jobs.
 We will validate their statements in an observational study and answer our first research question:
@@
 
-In response to this, we will present concepts how to resolve previously identified problems and give an answer on the second research question:
+In response to this, we will present concepts on how to resolve previously identified problems and answer the second research question:
 
@@
 
-The future implementation and validation of these proposals leads to our third research question which will be part of future work not part of this paper:
+The future implementation and validation of these proposals lead to our third research question which will be part of future work, not part of this paper:
 
@@
 \cite{7827078} divides RAs into three types: A (i) \emph{synthetic application} transforms and processes streams of information.
-RAs are prevalent in the area of UIs further: They can be identified as (ii) \emph{interactive applications} and help to describe and implement (iii) \emph{graphical animations}.
+RAs are prevalent in the area of UIs known as (ii) \emph{Interactive applications} and (iii) \emph{graphical animations}.
 
@@
 Various ways of implementing RAs emerged over time.
-The Observer design pattern \cite{gamma1995design}, originated in Object Oriented Programming \emph{(OOP)}, is a pattern we can find in many of todays API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior.
+The Observer design pattern \cite{gamma1995design}, originated in Object-Oriented Programming \emph{(OOP)}, is a pattern we can find in many of today's API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior.
 Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
@@
 RP is a declarative programming paradigm.
-Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved.
+Where engineers use imperative programming languages to tell a program every step \emph{how} to do something, a declarative language allows the description of \emph{what} wants to be achieved.
 The language runtime system then figures out a way to satisfy that given description.
-RP environments specifically provide abstractions to describe a dataflow and dependency graph which is used to transform and direct data at runtime.
+RP environments specifically provide abstractions to describe a dataflow and dependency graph used to transform and direct data at runtime.
 
-An RP environment provides (i) language constructs to describe mentioned graphs, often in the form of a DSL\cite{10.1145/2577080.2577083}.
-At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent application state is maintained.
+An RP environment provides (i) language constructs to describe the mentioned graphs, often in the form of a DSL\cite{10.1145/2577080.2577083}.
+At program runtime, a (ii) runtime system takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.
 
@@
 
-We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide basic functionality to interact with programs at runtime: Once code execution is paused (i.e.
+We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once code execution is paused (i.e.
 using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
-Manual control of the program execution allows to inspect its behavior step by step as well as assigning new values to variables ``on-the-fly''.
+Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
 
-Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the dataflow graph.
-In fact, RP provides its own challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g.
-the predecessor transformation according to the dataflow graph.
+Where traditional debuggers can inspect individual transformations of an RP based program, they fall short in understanding the underlying model of the dataflow graph.
+RP provides its challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
-A (ii) \emph{reactive debugger} has knowledge about the internal model of an RP runtime system.
-It leverages on it and provides specialized tools e.g.
-to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
-We will have a closer look to available reactive debugger solutions.
+A (ii) \emph{reactive debugger} knows about the internal model of an RP runtime system.
+It leverages on it and provides specialized tools e.g., to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
 
-The traditional as well as the reactive debugger works with the \emph{current} state of a program's execution only.
+Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only.
 They lack knowledge about what happened before or what is going to happen in the future.
-This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system.
+This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system.
 An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
@@
 The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables.
-It has the ability to push (``emit'') values to its dependents.
+It can push (``emit'') values to its dependents.
 Multiple depending Observables form and describe a dataflow graph eventually.
-A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable.
+A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable.
 This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables.
-Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+Values emitted from an Observable can be subscribed to at any point, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
@@
 
-We have worked together professional software engineers during the process of answering RQ1.
-On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people which understand themselves as users of RP and related technologies.
-E.g.
-even though Angular makes heavy use of RxJS, we will see that many engineers do not interact with its abstractions directly in order to build simple UIs.
-In the end we were able to gather interesting statements nonetheless which we are going to present in this section.
+We have worked together with professional software engineers during the process of answering RQ1.
+On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies.
+E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions to build simple UIs.
+In the end, we were able to gather compelling statements which we are going to present in this section.
 
@@
 
-We organized informal interviews which allowed us to gain first insight in how software engineers work with RP in their daily jobs.
+We organized informal interviews which allowed us to gain a first insight in how software engineers work with RP in their daily jobs.
 We talked to six engineers and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
 
-Four of our interview partners stated to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
+Four of our interview partners declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
 By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user.
-Our sixth interview partner finally was a backend engineer which used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
+Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
-All of the interviewees referred to RP with overall positive experience.
+All of the interviewees referred to RP with an overall positive experience.
 We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
-Hence a big strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.
+Hence a significant strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
+The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition'' an interview partner giving lectures in frontend web application development stated further.
 
 It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
-E.g.
-when using ngrx for state management, ``the framework takes care that you do not have to interact with Observables on your own, but leverages greatly on their potential behind the scenes''.
-As soon as our interviewees had to extend built-in functionalities with their own features, e.g.
-a new effect\footnote{\url{https://ngrx.io/guide/effects}}, all of them valued the possibility to interact with underlying Observables highly.
+E.g., when using ngrx for state management, ``The framework takes care that you do not have to interact with Observables on your own, but leverages greatly on their potential behind the scenes''.
+As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, they all valued the possibility to interact with underlying Observables highly.
 
-When asked explicitly about what they dislike the most about RP, all interview partners with no exception emphasized the debugging process of an RP application as a big nuisance.
+When asked explicitly about what they dislike the most about RP, all interview partners, with no exception, emphasized the debugging process of an RP application as a big nuisance.
 The fact that our interviewees remembered the search for a bug as something negative did not surprise us, hence a bug is usually something negative afflicted.
@@
 Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless.
-Further, one single partner mentioned marble diagrams as very valuable in order to understand how an RxJS Observable works, no matter if during development or when debugging code.
+Further, one single partner mentioned marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or when debugging code.
 Marble diagrams are a way to visualize and describe the value-emitting behavior of an Observable over time.
@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect five responses: One by an RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.
+After reaching out, we were able to collect five responses: One by an RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS, which includes the author of this paper.
 
 The reports provided us with diverse perspectives on various areas of software development using RxJS.
-One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it.
+One of the reporters focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it.
 I really disliked that once you started to test async code you were forced to pass a TestScheduler''.
-Allowing to pass the scheduler explicitly as parameter forced them to introduce code which was only necessary for testing reasons they stated further.
+Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
 With its current major version, RxJS 6 improved heavily on the  \mintinline{TypeScript}{TestScheduler}.
-The runtime environment itself can now be augmented with the scheduler which results in cleaner code.
+The runtime environment itself can be augmented with the scheduler now, which results in cleaner code.
 
@@
 One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
-On top, each data source could re-emit updated versions of previously requested information any time.
-Though tested thoroughly, after a week in production the first of many bugs was reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all'' they told us.
+On top, each data source could re-emit updated versions of previously requested information at any time.
+Though tested thoroughly, after a week in production, the first of many bugs were reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all,'' they told us.
 The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs.
-After the time-consuming log analysis, they finally were able resolve the bug.
+After the time-consuming log analysis, they finally were able to resolve the bug.
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
@@
 
-In a related war story, another engineer disclosed similar practices using  external tools: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create smaller working examples and iterate on them.'' Stackblitz\footnote{\url{https://stackblitz.com/}} is a full JavaScript development environment available online.
-They used this sandboxed setting to run, debug and iterate on selected pieces of a larger Observable stream.
+Another engineer disclosed similar practices using external tools in a related war story: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create smaller working examples and iterate on them.'' Stackblitz\footnote{\url{https://stackblitz.com/}} is a full JavaScript development environment available online.
+They used this sandboxed setting to run, debug, and iterate on selected pieces of a larger Observable stream.
 Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
@@
 A final story described yet another variation of the application of an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
-Like in the report before, it was necessary to extracted code from the actual applications code base.
+Like in the report before, it was necessary to extract parts from the codebase of the actual application.
 But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.
@@
 
-The overall sentiment in both the informal interviews as well as the reported war stories in regards of RP with RxJS was positive.
+The sentiment in both the informal interviews and the reported war stories in regards to RP with RxJS was positive.
 The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep.
-We heard some interesting reports how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
+We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
 Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.
 
-Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness.
-We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though.
-Yet traditional debugging tools allowed them to inspect various parts of a stream, they were unable to get more information about e.g.
-life cycles of the Observable composition.
+Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness.
+We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP.
+However, traditional debugging tools allowed them to inspect various parts of a stream, yet they were unable to get more information about, e.g., life cycles of the Observable composition.
 Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem.
@@
 Finally, we understood that if a problem is hard to replicate within the actual application, the engineers use external sandbox development environments to isolate specific parts of an Observable composition.
-This allows them to iterate on it faster than it would be possible otherwise.
+These allow them to iterate on it faster than it would be possible otherwise.
 
-After the evaluation of all reports and interviews it is our speculation that software engineers are truly lacking debugging tools which ``understand'' RP concepts provided through RxJS.
-Even though traditional debuggers might help up to some extent, they do not provide all the information an engineer is interested in.
-The last resort in this situation is often the manual augmentation, modification and extraction of source code as we saw repeatedly.
+After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools which ``understand'' RP concepts provided through RxJS.
+Even though traditional debuggers might help to some extent, they do not provide all the information an engineer is interested in.
+The last resort in this situation is often the manual augmentation, modification, and extraction of source code as we saw repeatedly.
 After a look at previous efforts in the area of reactive debugging, we are going to validate our assumption in Section~\ref{sec:study}.
@@
 Our work is not the first attempt to understand and improve on the struggles software engineers have to overcome when working on RP code, either with or without RxJS.
-We are going to present a selection of four utilities whereas two of them have an academic background.
-We either learned about them during our research because they were mentioned in one of the interviews and war stories or because we have worked with them personally in the past.
+We are going to present a selection of four utilities, whereas two of them have an academic background.
+We learned about them during our research either because they were mentioned in one of the interviews and war stories, or because we have worked with them in the past.
 
@@
 did focus their research heavily on RP during the mid 2010s.
-Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well.
+Besides the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well.
 In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al.
-elaborate on the necessity of RP debugging tools which can understand the specific language abstractions.
-With \emph{Reactive Inspector} plugin for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala.
+elaborate on the necessity of RP debugging tools that understand the specific language abstractions.
+They provide with the \emph{Reactive Inspector} plugin for the Eclipse IDE a reactive debugger for the REScala RP runtime environment.
 One of its main features is the visualization of the internal dependency graph.
-The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph.
-The execution of the program can be interrupted e.g.
-once a node retains a specific output value or an event of interest takes place in the runtime.
+The new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph.
+The execution of the program can be interrupted e.g., once a node retains a specific output value or an event of interest takes place in the runtime.
 Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without re-execution.
 
-Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with different RP runtime environment as well.
+Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with different RP runtime environments as well.
 
@@
 is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
-The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g.
-when an Observable is created or subscribed) which get collected.
+The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them.
 The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables.
-An additional marble diagram for every Observable is used to display the values it emits over time.
+A marble diagram visualizes the values emitted over time for every Observable further.
 
@@
 
-Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier.
-rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator along with sophisticated log monitoring facilities.
+Software engineers tend to augment their code with additional log trace statements where they have to track down a problem we saw earlier.
+rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator, along with sophisticated log monitoring facilities.
 Ideally, ``tagged'' Observables are created during development when the data-flows are composed for the first time.
@@
 Additional features are available through the debuggers console interface.
-A tagged Observable can be paused so it collects incoming values.
-They can then be emitted one after another or resumed at once.
+Engineers can ``pause'' a tagged Observable, so it collects incoming values.
+They can then be emitted one after another manually or resumed at once.
 
@@
 The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser.
-It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.
-Diagrams are generated over a given, fixed time and can be saved as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}}.
+It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization for snippets of RxJS code using marble diagrams.
+Diagrams are generated over a given, fixed time and available for download as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}} files.
 
@@
 They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector).
-Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-related tools provide small, differentiated feature sets.
+Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities, and full IDE integration.
+In contrast, the RxJS-related tools provide small, differentiated feature sets.
 Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).
@@
 
-We explored on how engineers use RP and what techniques they might apply for debugging problems earlier.
+We explored how engineers use RP and what techniques they might apply for debugging problems earlier.
 We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
-It is further our own experience that this practice is often not a productive one since the evaluation of trace information, once available, can be cumbersome and time-consuming.
-In addition, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
-We identified this circumstance as one of the main debugging practices when software engineers work with RxJS-based code.
+It is further our own experience that this practice is often not productive since the evaluation of trace information, once available, can be cumbersome and time-consuming.
+Also, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
+We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
 
-This is why we saw the need in validating earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
+That is why we saw the demand to validate earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
 
 \begin{itemize}
-	\item \emph{Hypothesis: If software engineers are given an RxJS-based problem, then they will modify the code in order to understand its behavior.}
+	\item \emph{Hypothesis: If software engineers must solve an RxJS-based problem, then they will modify the code in order to understand its behavior.}
 \end{itemize}
@@
 The subjects for our study were required to have experience in developing applications with RxJS.
-We did not distinguish between frontend or backend engineers, hence in the end were able to find four subjects willing to participate in our experiment.
-We were interested to see how the subjects apply debugging techniques they would use in a common situation at their jobs.
-This is why we decided to conduct the experiment in a rather uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
+We did not distinguish between frontend or backend engineers, so in the end, we were able to find four subjects willing to participate in our experiment.
+We were interested to see how the subjects apply debugging techniques they would use in an everyday situation at their jobs.
+Hence we decided to conduct the experiment in a somewhat uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
 Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention the hypothesis by intention.
@@
 We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
-The experiment was executed in two consecutive blocks of 25 minutes each.
+We executed the experiment two consecutive blocks of 25 minutes each.
 We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of the experiment.
-Each of these applications were rigged with two to three bugs which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
+Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
 A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
 We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process.
-We refrained to answer any question related to the ``where'' a bug has to be expected though.
+Though we refrained from answering any question related to the ``where'' a bug has to be expected.
 
-A week before the actual experiment, we sent out the participant briefing document to all of our subjects.
+A week before the experiment, we sent out the participant briefing document to all of our subjects.
 We outlined the course of action and provided them with an example ZIP file.
-This file contained the same setup as the file provided at the experiment and allowed the subjects to get accustomed with things like starting the web applications or running the test suites.
+This file contained the same setup as the file provided at the experiment and allowed the subjects to get accustomed to things like starting the web applications or running the test suites.
 
 We decided to monitor our subjects' progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
-This allowed us further to record the sessions with relative low technical effort for later evaluation.
+Furthermore, this allowed us to record the sessions with relatively low technical effort for later evaluation.
 
@@
 
-For both problems given, all subjects could be observed to make use of manually added trace log statements which they either added to existing arrow functions and/or using the \mintinline{TypeScript}{tap} operator.
+For both problems given, we could observe all subjects to make use of manually added trace log statements either with the existing arrow functions or using the \mintinline{TypeScript}{tap} operator.
 Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
@@
 
-The responses on the survey showed that 75\% of the subject population had two or more years of experience with RxJS.
+The survey responses showed that 75\% of the subject population had two or more years of experience with RxJS.
 Where all of them use RxJS to develop frontend applications, one subject declared having RxJS used for backend development as well.
-When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser and/or IDE.
-Two of them leverage additional tracing functionality of rxjs-spy and all of them use manual log statements.
+When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser or IDE.
+Two of them leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
 
@@
 
-We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs at some point.
+We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs.
 All of them used the produced information in order to track down problems in the application.
-The extraction to and reintegration from an external tool was not observed, though would have been yet another form of code modification.
+We could not observe the extraction to and reintegration from an external tool, which would have been yet another form of code modification.
 
@@
 
-All reviewed subjects exhibited the debugging behavior described in our initial hypothesis which we can accept based on the collected evidence.
+All reviewed subjects exhibited the debugging behavior described in the initial hypothesis, which we can accept based on the collected evidence.
 
-Talking to professionals during the informal reviews, evaluating the war stories as well as the hands-on experiences and results from the observational study allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application, hence answering RQ1: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL.
-This helps to identify simpler problems most of the time, though knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
+Talking to professionals during the informal reviews, evaluating the war stories and the hands-on experiences and results from the observational study, allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application, hence answering RQ1: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
+Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
 (ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}.
-This makes it hard for engineers to retrace dataflows and computations.
-And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
+This circumstance makes it hard for engineers to retrace dataflows and computations.
+And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
 These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
@@
 
-Our overall goal is to support engineers during the process of understanding the inner workings of RxJS data-flow graph.
+Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS data-flow graph.
 Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw.
-Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly though.
-Filter and search functionalities help, but do not resolve all UX challenges here.
+Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly, however.
+Filter and search functionalities help but do not resolve all UX challenges here.
 
 Visualizing the data-flow graph work better to show dependencies within the model compared to textual trace logs.
-Previous research showed that the sheer amount of nodes in a complex graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other not as easy to implement efficiently since the mass of collected data grows quickly.
+Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other not as easy to implement efficiently since the mass of collected data grows quickly.
 Storing a representation of the graph rarely enough since engineers are interested in the values emitted over time per node as well.
 
-We intend to find a solution that combines the power of textual trace logs with the ease-of-use of a graphical visualization for RxJS data-flow graphs.
+We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS data-flow graphs.
 
@@
 RP code needs to be augmented with additional functionality so telemetry data can be collected.
-This happens using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156} but especially in the area of RxJS often manually.
+Other solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156} but especially in the area of RxJS often manually.
 We want to make the manual modification of production source code obsolete.
@@
 
-75\% of our observational study's participants used the debugging tools integrated to their browser and/or IDE.
+75\% of our observational study participants used the debugging tools integrated into their browser or IDE.
 Even though some of them stated to know additional external utilities in the after-action survey, none of them went that far and used them during the one-hour session.
@@
 We understand the observed behavior as evidence for a hurdle preventing engineers from using specialized RP debugging utilities.
-In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser and/or an IDE as possible.
+In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser or an IDE as possible.
 
@@
 There are many more areas which could be worth a further look in the future.
-E.g.
-the combination of controlling the programs execution based on breakpoints within the data-flow graph or concepts of omniscient debugging are interesting fields as well.
-We will focus on the previous three topics for our own future work but keep other ideas in a backlog in case time allows additional iterations on them.
+E.g., the combination of controlling the program execution based on breakpoints within the dataflow graph or concepts of omniscient debugging are exciting fields as well.
+We will focus on the presented three topics for our future work, but keep additional ideas in a backlog for the case time allows additional iterations.
 
@@
 We worked with 15 professional software engineers while writing this paper.
-Six interview partners and five war story reports helped us building the foundation for a later observational study with a total of four participants.
+Six interview partners and five war story reports helped us build the foundation for a later observational study with four participants.
 
-While we claim nine interviewees and reporters to have been an appropriate number of peers for collecting the initial data basis, we are aware that our four subjects for the observational study lead to a result lacking significance.
+While we claim that nine interviewees and reporters have been an appropriate number of peers for collecting the initial data, we are aware that our four subjects for the observational study lead to a result lacking significance.
 
-More time given, we suggest to conduct the study with a bigger population again to obtain a result with higher significance and providing more insight in how engineers debug RxJS-based code.
+More time given, we suggest conducting the study with a bigger population again to obtain a result with higher significance and providing more insight into how engineers debug RxJS-based code.
 Doing so would possibly require a statistical evaluation of these results.
@@
 Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless.
-E.g.
-we would use a commonly known business domain for both problems presented to the subjects.
-This would allow them to understand the functionality of the application in front of them quicker and more focus on resolving the task given.
+E.g., we would use a commonly known business domain for both problems presented to the subjects, which would allow them to understand the functionality of the application in front of them quicker and more focus on resolving the task given.
 
@@
 We have looked into how they debug their applications and could identify several struggles they have to overcome in this process.
-We could proof that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal data-flow graph by observing professionals in an observational study.
+We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal data-flow graph by observing professionals in an observational study.
 
-On the way, we elaborated on previous, academical as well as non-academical efforts in the field of reactive debugging.
+On the way, we elaborated on previous, academic as well as non-academical efforts in the field of reactive debugging.
 We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.
@@
 With \emph{UX}, \emph{Automatic Code Augmentation} and \emph{Integration} we finally proposed three concepts where we intend to improve on a software engineers debugging experience, hence answering our second research question RQ2.
-We plan to work further on these topics in an upcoming workshop paper within the scope of our master's studies research.
+We plan to work further on these topics within the scope of our master's studies research and, eventually, answer our last research question RQ3.
 

### 2020-05-31 paper.tex
Its about RxJS-base Applicaitons in general

@@
 	\par
-	\Huge{Debugging of RxJS-based Frontend Web Applications}
+	\Huge{Debugging of RxJS-based Applications}
 	\par
@@
 Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
-We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs using RxJS for frontend web applications.
+We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs implemented using RxJS.
 
@@
 \begin{itemize}
-	\item \emph{RQ1: What challenges do frontend engineers face when debugging RxJS-based code?}
+	\item \emph{RQ1: What challenges do software engineers face when debugging RxJS-based code?}
 \end{itemize}
@@
 \begin{itemize}
-	\item \emph{RQ2: How can the experience of frontend engineers during debugging of RxJS-based code be improved?}
+	\item \emph{RQ2: How can the experience of software engineers during debugging of RxJS-based code be improved?}
 \end{itemize}
@@
 \begin{itemize}
-	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a frontend engineer?}
+	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a software engineer?}
 \end{itemize}
@@
 
-We have talked to multiple people in order to understand how software engineers, and in particular frontend engineers, apply RP using RxJS for building applications.
+We have talked to multiple people in order to understand how software engineers apply RP using RxJS for building applications.
 We have looked into how they debug their applications and could identify several struggles they have to overcome in this process.

### 2020-05-31 abstract.tex
Abstract

An application implemented using RxJS is unintuitive and time-consuming to debug. We will explore the practices software engineers apply during debugging and take a close look at the challenges they face in this workshop paper. On the way to propose solution concepts, we will validate previously identified struggles in an observational study and elaborate on previous research efforts in the field of reactive debugging.

### 2020-05-31 paper.tex
Edit Internal Review

@@
 
-Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs.
+Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} are specialized to work with imperative, control-flow oriented programs.
 E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
@@
 This lack of specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions, as we can confirm from our own experience.
-Previous research has looked into this field \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
+Various researchers have looked into this topic \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
 We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs implemented using RxJS.
@@
 \begin{itemize}
-	\item \emph{RQ3: What is the impact of a proposed solution on the debugging experience of a software engineer?}
+	\item \emph{RQ3: What is the impact of proposed solutions on the debugging experience of software engineers?}
 \end{itemize}
@@
 We will conclude this introductory section with the clarification of important terms.
-Section~\ref{sec:interviews} presents the interview results along with the collected war stories and formalizes two recurrent challenges.
-With Section~\ref{sec:discussion}, we will have a look at current and previous research regarding the debugging of RAs as well as specific tooling.
-We are going to validate presented challenges against currently available tools within Section~\ref{sec:study}, show threats to the validity of our work in Section~\ref{sec:threats} and finally end with an outlook on future iterations.
+Section~\ref{sec:interviews} presents the interview results along with the collected war story reports.
+Before we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}.
+We will finally answer RQ2 in Section~\ref{sec:future} ``Future Work''.
 
@@
 The Observer design pattern \cite{gamma1995design}, originated in Object-Oriented Programming \emph{(OOP)}, is a pattern we can find in many of today's API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior.
-Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
+FP shines with reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
 
@@
 The language runtime system then figures out a way to satisfy that given description.
-RP environments specifically provide abstractions to describe a dataflow and dependency graph used to transform and direct data at runtime.
+RP runtime environments specifically provide abstractions to describe dataflow and dependency graphs used to transform and direct data at runtime.
 
-An RP environment provides (i) language constructs to describe the mentioned graphs, often in the form of a DSL\cite{10.1145/2577080.2577083}.
-At program runtime, a (ii) runtime system takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.
+An RP language extension provides (i) language abstraction to describe the mentioned graph models, often in the form of a DSL\cite{10.1145/2577080.2577083}.
+At program runtime, a (ii) runtime environment takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.
 
@@
 
-We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once code execution is paused (i.e.
-using a breakpoint at a given statement), we gain access to the current call stack and the values assigned to variables of a given frame.
+We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, we gain access to the current call stack and the values assigned to variables of a given frame.
 Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
@@
 This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows.
-As a result, user experience (\emph{UX}) concepts need to be aligned for the engineers being able to interact with the sheer amount\cite{5287015} of information efficiently.
+As a result, user experience (\emph{UX}) concepts need to be aligned so engineers can interact with the sheer amount of information efficiently\cite{5287015}.
 
@@
 Its members and contributors defined a generic description of an RP API.
-They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}.
+They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, and JavaScript\footnote{\url{http://reactivex.io/languages.html}}.
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
@@
 
-We have worked together with professional software engineers during the process of answering RQ1.
 On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies.
-E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions to build simple UIs.
+E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs.
 In the end, we were able to gather compelling statements which we are going to present in this section.
@@
 By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user.
-Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
+Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
@@
 	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-debugging}.
-Each marble represents an emitted value.
-The vertical line at marble \mintinline{TypeScript}{6} indicates the Observable completed.}
+From left to right, each marble represents an emitted value.
+The vertical line at the last marble indicates that the Observable completed.}
 	\label{fig:marble-diagram}
@@
 
-Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
+Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
 One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
@@
 
-A final story described yet another variation of the application of an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
+A final story described yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
@@
 We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
-Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.
+Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the pre-provided functionality needs to be extended.
 
@@
 		  tap(console.log)      // Manually added log statement
-		).subscribe(showValue); // 0, 2, 4, 6
+		).subscribe(showValue); // Emits: 0, 2, 4, 6
 	\end{minted}
-	\caption{Debugging of an RxJS Observable using breakpoints and log statements. \mintinline{TypeScript}{showValue} renders an emitted value to the UI.}
+	\caption{Debugging of an RxJS Observable using breakpoints and log statements.}
 	\label{lst:rxjs-debugging}
@@
 Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
+That is why the debuggers step controls cannot operate on the dataflow graph as well; It just does not know this level of abstraction.
 
@@
 They provide with the \emph{Reactive Inspector} plugin for the Eclipse IDE a reactive debugger for the REScala RP runtime environment.
-One of its main features is the visualization of the internal dependency graph.
-The new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph.
+One of its main features is the visualization of the internal dataflow graph.
+A new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph.
 The execution of the program can be interrupted e.g., once a node retains a specific output value or an event of interest takes place in the runtime.
@@
 
-Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with different RP runtime environments as well.
+Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with other RP runtime environments.
 
@@
 is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
-The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them.
+The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them.
 The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables.
@@
 The reference implementation supports event collection for RxJS-based code in versions 4 and 5.
-It can either be used on a local development machine on existing projects or by copying code snippets to the online available.
+It is available as an online application where snippets of code can be inspected.
+A proof-of-concept implementation which works on a local development machine is accessible via the projects Git repository.
 
-The architectural design of \emph{rxfiddle} is similar to the one \emph{Reactive Inspector}: Banken et al.
-provide a system where different RP runtime environments should be able to be attached to the Visualizer easily.
+The architectural design of \emph{rxfiddle} is similar to the one of \emph{Reactive Inspector}: Banken et al. claim that their Visualizer can be adapted to different RP runtime environments as well.
 
@@
 rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator, along with sophisticated log monitoring facilities.
-Ideally, ``tagged'' Observables are created during development when the data-flows are composed for the first time.
+Ideally, ``tagged'' Observables are created during development when the dataflows are composed for the first time.
 Like the \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.
@@
 	\begin{minted}{TypeScript}
+		import { create } from 'rxjs-spy';
+		import { tag } from 'rxjs-spy/operators';
+
+		const spy = create();  // Create monitor
+		spy.log(/multiply/);   // Log tags matching RegEx /multiply/
+
 		interval(1000).pipe(
 		  map(i => i * 2),
-		  tag('multiply'),
+		  tag('multiply'),     // Tag Observable with "multiply"
 		  map(i => i - 1),
@@
 
-Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any Observable code at all.
+Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console (see Listing~\ref{lst:rxjs-spy-log}) without touching any Observable code at all.
 The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.
@@
 The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser.
-It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization for snippets of RxJS code using marble diagrams.
+It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.
 Diagrams are generated over a given, fixed time and available for download as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}} files.
@@
 They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector).
-Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities, and full IDE integration.
-In contrast, the RxJS-related tools provide small, differentiated feature sets.
-Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).
+Whereas Reactive Inspector for REScala provides an ``all-in-one'' solution with reactive breakpoints, omniscient debugging capabilities, and full IDE integration, the RxJS-related tools provide smaller, differentiated feature sets in contrast.
+Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or call for manual code preparation or modification (rxjs-spy).
 
@@
 We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
-We executed the experiment two consecutive blocks of 25 minutes each.
-We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of the experiment.
+We executed the experiment in two consecutive blocks of 25 minutes each.
+We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
 Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
@@
 
-For both problems given, we could observe all subjects to make use of manually added trace log statements either with the existing arrow functions or using the \mintinline{TypeScript}{tap} operator.
+For both problems given, we could observe all subjects to make use of manually added trace log statements either within an existing arrow function or using the \mintinline{TypeScript}{tap} operator.
 Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
@@
 
-Talking to professionals during the informal reviews, evaluating the war stories and the hands-on experiences and results from the observational study, allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application, hence answering RQ1: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
+Interviewing professionals, evaluating the war stories, the hands-on experiences, and the results from the observational study allowed us to identify three main challenges, software engineers have to overcome when working with RxJS within an application: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
 Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
@@
 
-Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS data-flow graph.
+Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS dataflow graph.
 Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw.
 Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly, however.
-Filter and search functionalities help but do not resolve all UX challenges here.
+Filter and search functionalities help but cannot resolve all UX challenges.
 
-Visualizing the data-flow graph work better to show dependencies within the model compared to textual trace logs.
-Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other not as easy to implement efficiently since the mass of collected data grows quickly.
-Storing a representation of the graph rarely enough since engineers are interested in the values emitted over time per node as well.
+A graphical representation of the dataflow graph is better suited to show dependencies within the model then textual trace logs.
+Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other also not easily implement efficiently as well.
 
-We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS data-flow graphs.
+We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS dataflow graphs.
+We want this solution to handle small as well as large graphs.
 
@@
 RP code needs to be augmented with additional functionality so telemetry data can be collected.
-Other solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156} but especially in the area of RxJS often manually.
+Previous solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156}.
+For engineers themselves though, this is often a manual process.
 We want to make the manual modification of production source code obsolete.
@@
 Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless.
-E.g., we would use a commonly known business domain for both problems presented to the subjects, which would allow them to understand the functionality of the application in front of them quicker and more focus on resolving the task given.
+E.g., we would use a commonly known business domain for both problems presented to the subjects.
+This would allow them to understand the functionality of the application in front of them quicker and focus more on resolving the task given.
 
@@
 We have looked into how they debug their applications and could identify several struggles they have to overcome in this process.
-We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal data-flow graph by observing professionals in an observational study.
+We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal dataflow graph by observing professionals in an observational study.
 
-On the way, we elaborated on previous, academic as well as non-academical efforts in the field of reactive debugging.
+On the way, we elaborated on previous, academic as well as non-academic efforts in the field of reactive debugging.
 We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.

### 2020-08-06 paper.tex
Final Draft
Move Previous Efforts

@@
 
-\newcommand{\version}{v0.4 Draft}
+\newcommand{\version}{v0.9 Final Draft}
 
@@
 Section~\ref{sec:interviews} presents the interview results along with the collected war story reports.
-Before we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}.
+After we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}.
 We will finally answer RQ2 in Section~\ref{sec:future} ``Future Work''.
@@
 
+\section{Validation of Interviews and War Stories}
+\label{sec:study}
+
+We explored how engineers use RP and what techniques they might apply for debugging problems earlier.
+We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
+It is further our own experience that this practice is often not productive since the evaluation of trace information, once available, can be cumbersome and time-consuming.
+Also, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
+We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
+
+That is why we saw the demand to validate earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
+
+\begin{itemize}
+	\item \emph{Hypothesis: If software engineers must solve an RxJS-based problem, then they will modify the code in order to understand its behavior.}
+\end{itemize}
+
+\subsection{Study Design}
+
+The subjects for our study were required to have experience in developing applications with RxJS.
+We did not distinguish between frontend or backend engineers, so in the end, we were able to find four subjects willing to participate in our experiment.
+We were interested to see how the subjects apply debugging techniques they would use in an everyday situation at their jobs.
+Hence we decided to conduct the experiment in a somewhat uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
+Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention the hypothesis by intention.
+
+We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
+We executed the experiment in two consecutive blocks of 25 minutes each.
+We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
+Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
+A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
+We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process.
+Though we refrained from answering any question related to the ``where'' a bug has to be expected.
+
+A week before the experiment, we sent out the participant briefing document to all of our subjects.
+We outlined the course of action and provided them with an example ZIP file.
+This file contained the same setup as the file provided at the experiment and allowed the subjects to get accustomed to things like starting the web applications or running the test suites.
+
+We decided to monitor our subjects' progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
+Furthermore, this allowed us to record the sessions with relatively low technical effort for later evaluation.
+
+The after-action survey was provided within 24 hours after a subject's participation in the main part of the study.
+We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
+The respective answers allowed us to put the observed actions into perspective and detect potential irregularities in case a subject acted differently as they would have in a ``real'' situation.
+
+\subsection{Results}
+
+After the subjects got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
+Two of them used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
+
+For both problems given, we could observe all subjects to make use of manually added trace log statements either within an existing arrow function or using the \mintinline{TypeScript}{tap} operator.
+Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
+After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
+
+The survey responses showed that 75\% of the subject population had two or more years of experience with RxJS.
+Where all of them use RxJS to develop frontend applications, one subject declared having RxJS used for backend development as well.
+When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser or IDE.
+Two of them leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
+
+\subsection{Interpretation}
+
+We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs.
+All of them used the produced information in order to track down problems in the application.
+We could not observe the extraction to and reintegration from an external tool, which would have been yet another form of code modification.
+
+For any subject, we could not identify any contradiction between the observed behavior and the answers given within the after-action survey.
+
+All reviewed subjects exhibited the debugging behavior described in the initial hypothesis, which we can accept based on the collected evidence.
+
+Interviewing professionals, evaluating the war stories, the hands-on experiences, and the results from the observational study allowed us to identify three main challenges, software engineers have to overcome when working with RxJS within an application: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
+Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
+(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}.
+This circumstance makes it hard for engineers to retrace dataflows and computations.
+And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
+These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
+
 \section{Discussion of Previous Efforts}
@@
 
-\section{Validation of Interviews and War Stories}
-\label{sec:study}
-
-We explored how engineers use RP and what techniques they might apply for debugging problems earlier.
-We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
-It is further our own experience that this practice is often not productive since the evaluation of trace information, once available, can be cumbersome and time-consuming.
-Also, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
-We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
-
-That is why we saw the demand to validate earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
-
-\begin{itemize}
-	\item \emph{Hypothesis: If software engineers must solve an RxJS-based problem, then they will modify the code in order to understand its behavior.}
-\end{itemize}
-
-\subsection{Study Design}
-
-The subjects for our study were required to have experience in developing applications with RxJS.
-We did not distinguish between frontend or backend engineers, so in the end, we were able to find four subjects willing to participate in our experiment.
-We were interested to see how the subjects apply debugging techniques they would use in an everyday situation at their jobs.
-Hence we decided to conduct the experiment in a somewhat uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
-Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention the hypothesis by intention.
-
-We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
-We executed the experiment in two consecutive blocks of 25 minutes each.
-We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
-Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
-A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
-We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process.
-Though we refrained from answering any question related to the ``where'' a bug has to be expected.
-
-A week before the experiment, we sent out the participant briefing document to all of our subjects.
-We outlined the course of action and provided them with an example ZIP file.
-This file contained the same setup as the file provided at the experiment and allowed the subjects to get accustomed to things like starting the web applications or running the test suites.
-
-We decided to monitor our subjects' progress remotely using voice chat and screen sharing due to the COVID-19 situation at the time of our study.
-Furthermore, this allowed us to record the sessions with relatively low technical effort for later evaluation.
-
-The after-action survey was provided within 24 hours after a subject's participation in the main part of the study.
-We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
-The respective answers allowed us to put the observed actions into perspective and detect potential irregularities in case a subject acted differently as they would have in a ``real'' situation.
-
-\subsection{Results}
-
-After the subjects got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
-Two of them used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
-
-For both problems given, we could observe all subjects to make use of manually added trace log statements either within an existing arrow function or using the \mintinline{TypeScript}{tap} operator.
-Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
-After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
-
-The survey responses showed that 75\% of the subject population had two or more years of experience with RxJS.
-Where all of them use RxJS to develop frontend applications, one subject declared having RxJS used for backend development as well.
-When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser or IDE.
-Two of them leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
-
-\subsection{Interpretation}
-
-We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs.
-All of them used the produced information in order to track down problems in the application.
-We could not observe the extraction to and reintegration from an external tool, which would have been yet another form of code modification.
-
-For any subject, we could not identify any contradiction between the observed behavior and the answers given within the after-action survey.
-
-All reviewed subjects exhibited the debugging behavior described in the initial hypothesis, which we can accept based on the collected evidence.
-
-Interviewing professionals, evaluating the war stories, the hands-on experiences, and the results from the observational study allowed us to identify three main challenges, software engineers have to overcome when working with RxJS within an application: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
-Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
-(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}.
-This circumstance makes it hard for engineers to retrace dataflows and computations.
-And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
-These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
-
 \section{Future Work}

### 2020-08-06 paper.tex
Rewrite Interivew/War Story Quotes

@@
 All of the interviewees referred to RP with an overall positive experience.
-We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
+We heard statements similar to ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
 Hence a significant strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition'' an interview partner giving lectures in frontend web application development stated further.
+The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' an interview partner giving lectures in frontend web application development stated further.
 
 It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
-E.g., when using ngrx for state management, ``The framework takes care that you do not have to interact with Observables on your own, but leverages greatly on their potential behind the scenes''.
+E.g., when using ngrx for state management, ``The framework hides Observables from its main API surface carefully so you do not have to interact with them directly''.
 As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, they all valued the possibility to interact with underlying Observables highly.
 
-When asked explicitly about what they dislike the most about RP, all interview partners, with no exception, emphasized the debugging process of an RP application as a big nuisance.
-The fact that our interviewees remembered the search for a bug as something negative did not surprise us, hence a bug is usually something negative afflicted.
-What did alarm us though was the fact that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' were prevalent and gave insight in \emph{why} our interview partners dislike RP debugging in particular.
+When asked explicitly about what they dislike the most about RP, all interview partners, with no exception, emphasized the debugging process of an RP application as unsatisfactory.
+The fact that our interviewees remembered the search for a bug as something negative did not surprise, hence a bug is commonly something negative afflicted.
+It was remarkable though that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' were prevalent and showed \emph{why} our interview partners dislike RP debugging in particular.
 Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless.
@@
 The reports provided us with diverse perspectives on various areas of software development using RxJS.
-One of the reporters focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it.
-I really disliked that once you started to test async code you were forced to pass a TestScheduler''.
+One of the reporters focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
+I really disliked that, once you started to test asynchronous code, you were forced to pass a TestScheduler''.
 Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
@@
 On top, each data source could re-emit updated versions of previously requested information at any time.
-Though tested thoroughly, after a week in production, the first of many bugs were reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all,'' they told us.
+After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed information kept changing/flickering (where we expected it to change exactly once and stay), other data got not updated at all when it was supposed to,'' they told us.
 The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs.
-After the time-consuming log analysis, they finally were able to resolve the bug.
+After the time-consuming log analysis, they finally were able to resolve all bugs.
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
@@
 
-Another engineer disclosed similar practices using external tools in a related war story: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz which gave us ability to quickly create smaller working examples and iterate on them.'' Stackblitz\footnote{\url{https://stackblitz.com/}} is a full JavaScript development environment available online.
-They used this sandboxed setting to run, debug, and iterate on selected pieces of a larger Observable stream.
+Another engineer disclosed similar practices using external tools in a related war story: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed P999 to run, debug, and iterate on selected pieces of a larger Observable stream.
 Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
@@
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
-But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help to dig into detailed runtime and life cycle behaviors of the Observables.''.
+But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the Observable.''
 

### 2020-08-06 paper.tex
Rearrange Marble Diagrams And Code Participants

@@
 
-		of('hello', 'foo', 'world').pipe(  // Create Observable
-			filter(s => s !== 'foo'),        // filter Operator
-			map(s => s.toUpperCase())        // map Operator
-		).subscribe(console.log)           // Emits: HELLO, WORLD
+		of(0, 1, 2, 3, 4).pipe(	// Create Observable with Integers
+			filter(i => i < 4),		// Omit Integers >= 4
+			map(i => i * 2)				// Multiply Integer with 2
+		).subscribe(console.log) // Emits 0, 2, 4, 6
 	\end{minted}
-	\caption{Basic RxJS Example}
+	\caption{Basic RxJS example creating an Observable emitting five integers. Each integer is processed by two operators and finally written to the console.}
 	\label{lst:rp-with-rxjs}
 \end{listing}
+
+The RxJS community uses \emph{marble diagrams} as shown in Figure~\ref{fig:marble-diagram} to document \cite{marblediagrams} the runtime behavior of an Observable visually.
+Unit test libraries\cite{marbletesting} leverage on this abstraction to encode the behavior of mocked Observables or to describe assertions.
+
+\begin{figure}[H]
+	\centering
+	\input{figures/marble-diagram.pdf_tex}
+	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
+From left to right, each marble represents an emitted value.
+The vertical line at the last marble indicates that the Observable completed.}
+	\label{fig:marble-diagram}
+\end{figure}
 
@@
 We organized informal interviews which allowed us to gain a first insight in how software engineers work with RP in their daily jobs.
-We talked to six engineers and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
+We talked to five engineers (following identified by the codes \emph{I1} to \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
 
-Four of our interview partners declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
-By being the creator of a visualizer for RxJS Observables, the fifth interviewee was a proficient RxJS user.
-Our sixth interview partner was finally a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
+Our first three interview partners I1 to I3 declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
+By being the creator of a visualizer for RxJS Observables, I4 was a proficient RxJS user.
+Our fifth interview partner I5 was a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
@@
 Hence a significant strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' an interview partner giving lectures in frontend web application development stated further.
+The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' I3 giving lectures in frontend web application development stated further.
 
 It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
-E.g., when using ngrx for state management, ``The framework hides Observables from its main API surface carefully so you do not have to interact with them directly''.
-As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, they all valued the possibility to interact with underlying Observables highly.
+E.g., when using ngrx for state management, ``The framework hides Observables from its main API surface carefully so you do not have to interact with them directly'' (I2).
+As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying Observables highly.
 
@@
 The fact that our interviewees remembered the search for a bug as something negative did not surprise, hence a bug is commonly something negative afflicted.
-It was remarkable though that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' were prevalent and showed \emph{why} our interview partners dislike RP debugging in particular.
-Multiple interviewees mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as helpful when debugging Angular/ngrx applications nonetheless.
-Further, one single partner mentioned marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or when debugging code.
-Marble diagrams are a way to visualize and describe the value-emitting behavior of an Observable over time.
-They are accepted throughout the ReactiveX community for documentation\cite{marblediagrams} and testing reasons\cite{marbletesting}.
-
-\begin{figure}[H]
-	\centering
-	\input{figures/marble-diagram.pdf_tex}
-	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rxjs-debugging}.
-From left to right, each marble represents an emitted value.
-The vertical line at the last marble indicates that the Observable completed.}
-	\label{fig:marble-diagram}
-\end{figure}
+It was remarkable though that statements like ``In 99\% of all cases, I add console.log statements manually and run the program over and over again, trying to understand what is happening'' (I1) were prevalent and showed \emph{why} our partners dislike RP debugging in particular.
+I1 to I3 mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as particular helpful when debugging Angular/ngrx applications nonetheless.
+Further, I1 noted marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or when debugging code.
 
@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect five responses: One by an RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS, which includes the author of this paper.
+After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another two reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.
 
 The reports provided us with diverse perspectives on various areas of software development using RxJS.
-One of the reporters focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
-I really disliked that, once you started to test asynchronous code, you were forced to pass a TestScheduler''.
+R3 focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
+I really disliked [\dots] you was forced to pass a TestScheduler [explicitly]''.
 Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
@@
 
-Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
-One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
+Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
+R5 described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
 On top, each data source could re-emit updated versions of previously requested information at any time.
 After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed information kept changing/flickering (where we expected it to change exactly once and stay), other data got not updated at all when it was supposed to,'' they told us.
-The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints.
-``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs.
-After the time-consuming log analysis, they finally were able to resolve all bugs.
+The browsers debugger tools and its breakpoints did not help much since the operators were executed several times.
+Other parts of the stream were impossible to grasp, even with conditional breakpoints.
+``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs.
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
-(We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)
+(We will closer elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion}.)
 
-Another engineer disclosed similar practices using external tools in a related war story: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed P999 to run, debug, and iterate on selected pieces of a larger Observable stream.
+In a related war story, R2 discloses similar invasive practices using an external tool when they implemented logic to request and cache batches of a remote resource: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed them to run, debug, and iterate on selected pieces of a larger Observable stream.
 Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
 
-A final story described yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, this engineer used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
+A final story describes yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, R5 used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
@@
 
-We worked with 15 professional software engineers while writing this paper.
-Six interview partners and five war story reports helped us build the foundation for a later observational study with four participants.
+We worked with 14 professional software engineers while writing this paper.
+Five interview partners and five war story reports helped us build the foundation for a later observational study with four participants.
 

### 2020-08-06 paper.tex
Code Subjects

@@
 The subjects for our study were required to have experience in developing applications with RxJS.
-We did not distinguish between frontend or backend engineers, so in the end, we were able to find four subjects willing to participate in our experiment.
+We did not distinguish between frontend or backend engineers, so in the end, we were able to recruit four subjects willing to participate in our experiment.
 We were interested to see how the subjects apply debugging techniques they would use in an everyday situation at their jobs.
@@
 
-\subsection{Results}
+\subsection{Study Execution and Results}
 
-After the subjects got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
+After the subjects S1 to S4 got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
 Two of them used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
@@
 Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
-After the 25 minutes of the second block exceeded, one subject noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
+After the 25 minutes of the second block exceeded, S2 noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
 
-The survey responses showed that 75\% of the subject population had two or more years of experience with RxJS.
-Where all of them use RxJS to develop frontend applications, one subject declared having RxJS used for backend development as well.
-When asked what tools they usually use for debugging, only one subject told not to use the traditional debugger of their browser or IDE.
-Two of them leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
+The survey responses showed that 75\% (S2 to S4) of the subject population had two or more years of experience with RxJS.
+Where all of them use RxJS to develop frontend applications, S4 declared having RxJS used for backend development as well.
+When asked what tools they usually use for debugging, only S1 stated not to use the traditional debugger of their browser or IDE.
+S1 and S3 leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
 

### 2020-08-11 paper.tex
Introduction Stuff

@@
 
+\usepackage{./tikzit}
+
 \setminted{autogobble=true, tabsize=2, linenos=true, frame=single, breaklines=true}
 
-\newcommand{\version}{v0.9 Final Draft}
+\newcommand{\version}{v0.10}
 
@@
 
-An application providing a user interface (\emph{UI}) can be seen as a reactive application (\emph{RA})\cite{7827078}: It handles a constant flow of user- and system-created events in order to give the user a representation of the applications state in a timely fashion.
-One way of implementing reactive behavior is the Observer design pattern\cite{gamma1995design}, where an observer (e.g., the UI rendering code) is repeatedly informed about the changed application state.
-With Reactive Programming (\emph{RP}), another way to implement RAs is trending.
-Language extensions, runtime environments and domain-specific languages (\emph{DSL}) enable RP based on functional programming (\emph{FP}) principles.
-With RP, software engineers describe how data flows through a RA, how it depends on each other and how it needs to be transformed over time in a declarative way.
+Applications handling constant flow of data are known as Reactive Applications (\emph{RA})\cite{10.1145/2501654.2501666}.
+Based on this definition, a (graphical) user interface (\emph{UI} or \emph{GUI}) is a RA\cite{7827078}, as it continuously processes two dataflows as shown in Figure~\ref{fig:ui-data-flows}: External user input (e.g.
+mouse, touch or keyboard interaction) is interpreted and forwarded to the system.
+Once the system processed an interaction and updated its internal state accordingly, it notifies the UI about these changes which then are relayed to the user.
 
-Utilities to debug the runtime behavior of a program are essential to software engineers today.
-They help to understand the inner workings of unknown source code or allow to track down a problem which may lead to an unexpected failure in a particular situation.
-Debugging utilities took different forms over time: From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level; software engineers have access to versatile debugging tools at the present day.
+\begin{figure}[H]
+	\centering
+	\begin{tikzpicture}[node distance = 5.5cm, auto]
+		\tikzstyle{block} = [rectangle, draw,
+		text width=5em, text centered, rounded corners, minimum height=2.5em]
 
-Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} are specialized to work with imperative, control-flow oriented programs.
+		\node [block] (user) {User};
+		\node [block, right of=user] (ui) {UI};
+		\node [block, right of=ui] (system) {System};
+
+		\draw [<->] (user) to node {Interaction} (ui);
+		\draw [->] (ui.north) to[bend left=20] node {Interaction Data Flow} (system.north);
+		\draw [->] (system.south) to[bend left=20] node {Update Data Flow} (ui.south);
+	\end{tikzpicture}
+	\caption{Basic Data Flows in a UI Application.}
+	\label{fig:ui-data-flows}
+\end{figure}
+
+The Observer design pattern\cite{gamma1995design} facilitates the implementation of RAs.
+Variations of the pattern are omnipresent today and can be found in places like the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications}.
+With Reactive Programming (\emph{RP}), a programming paradigm rooted in functional programming (\emph{FP}), software engineers have yet another tool to choose from to implement a RA.
+With RP, a domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively.
+A runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.
+
+According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.
+
+Modern IDEs and web browsers ship with their own set of debugging tools.
+These traditional debuggers are specialized in working with imperative, controlflow oriented program code.
 E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor; it will lead to the internals of the RP runtime instead.
+Assuming a dataflow oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
 
-This lack of specific knowledge about RP concepts in traditional debugging utilities does often end in time-consuming debugging sessions, as we can confirm from our own experience.
-Various researchers have looked into this topic \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} over the past years already.
-We will revisit their results in this paper and elaborate on the particular challenges of debugging RAs implemented using RxJS.
-
-Transcripts of informal interviews and ``war stories'' collected from multiple practitioners will give insight into the challenges they face in their day-to-day jobs.
-We will validate their statements in an observational study and answer our first research question:
+These exemplary questions demonstrate the limits of a traditional controlflow focused debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry.
+Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities.
+For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs in applying RP.
+Based on this collected evidence, we will validate these statements in an observational study using RxJS and search for an answer to our first research question:
 
@@
 
-In response to this, we will present concepts on how to resolve previously identified problems and answer the second research question:
+In response to this, we are going to present concepts on how to resolve previously identified problems and answer the second research question:
 
@@
 
-\subsection{Debugging Concepts}
+\subsection{Debugging Model and Concepts}
+
+Where the standard glossary defines \emph{debug} as an activity ``to detect, locate, and correct faults in a computer program''\cite{ieeeglossary} in general, Layman et al.
+\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized the inner workings of the debugging process with the Debugging Process Model further:
+
+\begin{figure}[H]
+	\centering
+	\tikzfig{debugging-process-model}
+	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
+	\label{fig:debugging-process-model}
+\end{figure}
+
+TODO LEAD OVER
 
@@
 	\centering
-	\input{figures/marble-diagram.pdf_tex}
+	\tikzfig{marble-diagram}
 	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
@@
 
-\subsection{Informal Interviews}
+\subsection{Interviews}
 
@@
 
-After we built a first intuition for how software engineers work with RP by evaluating the informal interviews, we were interested in more RxJS-specific, hands-on experiences.
+After we built a first intuition for how software engineers work with RP by evaluating the interviews, we were interested in more RxJS-specific, hands-on experiences.
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
@@
 
-The sentiment in both the informal interviews and the reported war stories in regards to RP with RxJS was positive.
+The sentiment in both the interviews and the reported war stories in regards to RP with RxJS was positive.
 The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep.
@@
 The survey responses showed that 75\% (S2 to S4) of the subject population had two or more years of experience with RxJS.
-Where all of them use RxJS to develop frontend applications, S4 declared having RxJS used for backend development as well.
-When asked what tools they usually use for debugging, only S1 stated not to use the traditional debugger of their browser or IDE.
-S1 and S3 leverage additional tracing functionality of rxjs-spy, and all of them use manual log statements.
+Where all of them use RxJS to develop frontend applications, S4 declared having used  RxJS for backend development as well.
+When asked what tools they usually use for debugging, only S1 stated not to use the traditional debugger of their IDE at all.
+S1 and S3 leverage additional tracing functionality of rxjs-spy, and all four of our subjects use manual log statements.
 

### 2020-08-11 paper.tex
Model Stuff

@@
 With Reactive Programming (\emph{RP}), a programming paradigm rooted in functional programming (\emph{FP}), software engineers have yet another tool to choose from to implement a RA.
-With RP, a domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively.
-A runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.
+A domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively.
+A RP runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.
 
-According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.
+According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.
 
@@
 
-These exemplary questions demonstrate the limits of a traditional controlflow focused debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry.
+These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry.
 Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities.
-For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs in applying RP.
-Based on this collected evidence, we will validate these statements in an observational study using RxJS and search for an answer to our first research question:
+For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when applying RP.
+Based on this collected evidence, we will validate their statements in an observational study using RxJS and search for an answer to our first research question:
 
@@
 
-In response to this, we are going to present concepts on how to resolve previously identified problems and answer the second research question:
+In response to this, we are going to present concepts on how to resolve previously identified challenges and answer the second research question:
 
@@
 
-The future implementation and validation of these proposals lead to our third research question which will be part of future work, not part of this paper:
+The implementation and validation of these proposals lead to our third and last research question, whose answer will be part of a future iteration on the topic:
 
@@
 We will conclude this introductory section with the clarification of important terms.
-Section~\ref{sec:interviews} presents the interview results along with the collected war story reports.
-After we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}.
-We will finally answer RQ2 in Section~\ref{sec:future} ``Future Work''.
+Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports.
+We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RAs in Section~\ref{sec:discussion}.
+Finally, we will answer RQ2 in Section~\ref{sec:future} ``Future Work''.
 
@@
 
-\subsection{Debugging Model and Concepts}
+\subsection{Debugging Process Model}
 
 Where the standard glossary defines \emph{debug} as an activity ``to detect, locate, and correct faults in a computer program''\cite{ieeeglossary} in general, Layman et al.
-\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized the inner workings of the debugging process with the Debugging Process Model further:
+\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized debugging as an iterative hypothesis refinement process, as shown in Figure~\ref{fig:debugging-process-model}:
 
@@
 
-TODO LEAD OVER
+The process consists of three steps and includes a feedback loop: After the engineer (i) gathered sufficient context information (e.g.
+ways to reproduce the failure or details about external factors) and understands the situation satisfactory, they generate a hypothesis on the origins of the bug or what impact a change made to the program might have.
+With the intent to proof their hypothesis, the engineer then (ii) instruments the defective program using suitable tools (e.g.
+adding log statements, setting breakpoints or removing code parts).
+Finally, the instrumented system gets (iii) challenged against the formed hypothesis.
+E.g., code statements are executed step by step using a debugger or trace logs are analyzed and compared against expected behavior.
+If the hypothesis turns out to be correct, the debugging process stops.
+If not, the newly gained knowledge about the problem is used to build a refined hypothesis and start a new iteration.
 
-We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, we gain access to the current call stack and the values assigned to variables of a given frame.
+\subsection{Debugger Concepts}
+
+Software engineers have an arsenal of tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
+Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.
+
+We will use three main concepts to categorize debugging utilities in this paper: Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
 Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
 
-Where traditional debuggers can inspect individual transformations of an RP based program, they fall short in understanding the underlying model of the dataflow graph.
 RP provides its challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
-A (ii) \emph{reactive debugger} knows about the internal model of an RP runtime system.
-It leverages on it and provides specialized tools e.g., to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
+A (ii) \emph{reactive debugger} can interpret the underlying graph model of a RP runtime system.
+It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
 
-Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only.
-They lack knowledge about what happened before or what is going to happen in the future.
+Traditional, as well as reactive debuggers work with the \emph{current} state of a programs execution only.
+They lack information about what happened before or what is going to happen in the future.
 This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system.
@@
 Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
-This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows.
-As a result, user experience (\emph{UX}) concepts need to be aligned so engineers can interact with the sheer amount of information efficiently\cite{5287015}.
 
@@
 ``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project.
-Its members and contributors defined a generic description of an RP API.
-They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, and JavaScript\footnote{\url{http://reactivex.io/languages.html}}.
+Its members and contributors created a generic description of a RP API.
+They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, or JavaScript\footnote{\url{http://reactivex.io/languages.html}}.
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
@@
 This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables.
-Values emitted from an Observable can be subscribed to at any point, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+Values emitted from an Observable can be subscribed, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
-RxJS is the reference implementation of the ReactiveX API specification for JavaScript and TypeScript.
-It is used thoroughly by large projects like Angular\cite{angualrrxjs}.
+RxJS is the reference implementation of the ReactiveX API specification for JavaScript.
+Its current major version 6 is implemented using TypeScript and is used thoroughly by large projects like Angular\cite{angualrrxjs}.
 Listing \ref{lst:rp-with-rxjs} shows an example of RP using RxJS in TypeScript.
@@
 	\end{minted}
-	\caption{Basic RxJS example creating an Observable emitting five integers. Each integer is processed by two operators and finally written to the console.}
+	\caption{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
 	\label{lst:rp-with-rxjs}
@@
 From left to right, each marble represents an emitted value.
-The vertical line at the last marble indicates that the Observable completed.}
+The vertical line at the last marble indicates that the Observable completed after emitting \mintinline{TypeScript}{6}.}
 	\label{fig:marble-diagram}

### 2020-08-15 paper.tex
Get rid of Reactive Application

@@
 
-\newcommand{\version}{v0.10}
+\newcommand{\version}{v0.11}
 
@@
 
-Applications handling constant flow of data are known as Reactive Applications (\emph{RA})\cite{10.1145/2501654.2501666}.
-Based on this definition, a (graphical) user interface (\emph{UI} or \emph{GUI}) is a RA\cite{7827078}, as it continuously processes two dataflows as shown in Figure~\ref{fig:ui-data-flows}: External user input (e.g.
+The (graphical) user interface (\emph{UI} or \emph{GUI}) of an application handles two constant flows of data: External user input (e.g.
 mouse, touch or keyboard interaction) is interpreted and forwarded to the system.
-Once the system processed an interaction and updated its internal state accordingly, it notifies the UI about these changes which then are relayed to the user.
+Once the system processed an interaction and updated its internal state accordingly, it notifies the UI about these changes which are relayed to the user.
 
@@
 	\end{tikzpicture}
-	\caption{Basic Data Flows in a UI Application.}
+	\caption{Basic Dataflows in a UI Application.}
 	\label{fig:ui-data-flows}
@@
 
-The Observer design pattern\cite{gamma1995design} facilitates the implementation of RAs.
-Variations of the pattern are omnipresent today and can be found in places like the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications}.
-With Reactive Programming (\emph{RP}), a programming paradigm rooted in functional programming (\emph{FP}), software engineers have yet another tool to choose from to implement a RA.
-A domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively.
-A RP runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.
+To implement the dataflows as shown in Figure~\ref{fig:ui-data-flows} to drive a UI, the Observer design pattern\cite{gamma1995design} is often used and variations of the pattern are omnipresent today.
+E.g., a modified version of the pattern can be found in the event handling interface of the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications} implemented by internet browsers.
+
+The Observer design pattern has its roots in the Object Oriented Programming paradigm (\emph{OOP}), hence relies on imperative code constructs to form a dataflow.
+Reactive Programming (\emph{RP}) is another approach to realize such flows: It inherits the declarative way of implementing functionality from Functional Programming (\emph{FP}), i.e.
+dataflows are rather described than implemented step by step\cite{10.1145/2501654.2501666}.
+RP functionality is usually available in the form of a library providing necessary abstractions, either for imperative or declarative programming languages.
 
@@
 
-Modern IDEs and web browsers ship with their own set of debugging tools.
+Modern IDEs and internet browsers ship with their own set of debugging tools.
 These traditional debuggers are specialized in working with imperative, controlflow oriented program code.
@@
 Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports.
-We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RAs in Section~\ref{sec:discussion}.
+We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RP code in Section~\ref{sec:discussion}.
 Finally, we will answer RQ2 in Section~\ref{sec:future} ``Future Work''.
-
-\subsection{Reactive Applications}
-
-Salvaneschi et al.
-\cite{7827078} define a RA as a computer program which processes continuos, indefinite and ever-changing flows of data.
-Think of a spreadsheet application\cite{10.1145/2501654.2501666} that links the values of cells to other cells using formulas.
-Once a cell is changed, it is the task of the application to update all dependant cells accordingly and cascade the updates throughout the model.
-This dataflow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.
-
-The classification schema by Salvaneschi et al.
-\cite{7827078} divides RAs into three types: A (i) \emph{synthetic application} transforms and processes streams of information.
-RAs are prevalent in the area of UIs known as (ii) \emph{Interactive applications} and (iii) \emph{graphical animations}.
 
@@
 
-Various ways of implementing RAs emerged over time.
-The Observer design pattern \cite{gamma1995design}, originated in Object-Oriented Programming \emph{(OOP)}, is a pattern we can find in many of today's API implementations\cite{alabor:2019:reactiveappllications} allowing reactive behavior.
-FP shines with reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}.
+RP is a declarative programming paradigm originated in FP.
+Where engineers use imperative programming languages to specify every step \emph{how} a program has to do something, declarative languages allow to describe \emph{what} the program should achieve ultimately.
+A runtime system then figures out a way to satisfy that description and executes it.
+RP functionality is usually provided in form of a language extension for a specific programming language (e.g.
+REScala for Scala\cite{10.1145/2577080.2577083}) or as a library (e.g.
+RxJS for JavaScript\cite{rxjs})
 
-RP is a declarative programming paradigm.
-Where engineers use imperative programming languages to tell a program every step \emph{how} to do something, a declarative language allows the description of \emph{what} wants to be achieved.
-The language runtime system then figures out a way to satisfy that given description.
-RP runtime environments specifically provide abstractions to describe dataflow and dependency graphs used to transform and direct data at runtime.
-
-An RP language extension provides (i) language abstraction to describe the mentioned graph models, often in the form of a DSL\cite{10.1145/2577080.2577083}.
-At program runtime, a (ii) runtime environment takes care that changed values get propagated through the graph, transformations recomputed where necessary and a consistent application state is maintained.
+Either way, both usually provide a (i) domain specific language (\emph{DSL}) to describe dataflow graphs, how they depend on each other and how data flowing through should be transformed.
+At program execution, a (ii) runtime environment evaluates these descriptions and creates a representation of the specified graphs.
+It then takes care that values are processed and propagated correctly through them as well as that a consistent system state\cite{10.1145/2501654.2501666} is always maintained.
 
@@
 
-RxJS is the reference implementation of the ReactiveX API specification for JavaScript.
+RxJS\cite{rxjs} is the reference implementation of the ReactiveX API specification for JavaScript.
 Its current major version 6 is implemented using TypeScript and is used thoroughly by large projects like Angular\cite{angualrrxjs}.

### 2020-08-15 paper.tex
Polish

@@
 We organized informal interviews which allowed us to gain a first insight in how software engineers work with RP in their daily jobs.
-We talked to five engineers (following identified by the codes \emph{I1} to \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
+We talked to five engineers (following identified using the codes \emph{I1} through \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
+All interviews were conducted remotely using video chat.
 
@@
 All of the interviewees referred to RP with an overall positive experience.
-We heard statements similar to ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions.
+We heard statements similar to ``RP is a good way for composing multiple data sources'' (I1) and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' (I5) repeatedly throughout all sessions.
 Hence a significant strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' I3 giving lectures in frontend web application development stated further.
+The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' stated I3, giving lectures in frontend web application development.
 
@@
 E.g., when using ngrx for state management, ``The framework hides Observables from its main API surface carefully so you do not have to interact with them directly'' (I2).
-As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying Observables highly.
+As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying Observables though.
 
-When asked explicitly about what they dislike the most about RP, all interview partners, with no exception, emphasized the debugging process of an RP application as unsatisfactory.
+When asked explicitly about what they dislike the most about RP, all interview partners, with no exception, emphasized the debugging process of an RP program as unsatisfactory.
 The fact that our interviewees remembered the search for a bug as something negative did not surprise, hence a bug is commonly something negative afflicted.
@@
 I1 to I3 mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as particular helpful when debugging Angular/ngrx applications nonetheless.
-Further, I1 noted marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or when debugging code.
+Further, I1 noted marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or debugging.
 
@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another two reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.
+After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another set of reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.
 
@@
 R3 focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
-I really disliked [\dots] you was forced to pass a TestScheduler [explicitly]''.
+I really disliked [\dots] you were forced to pass a TestScheduler''.
 Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
@@
 
-Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners.
+Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our interview partners.
 R5 described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
@@
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
-(We will closer elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion}.)
+(We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion}.)
 
@@
 
-A final story describes yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, R5 used Rx Visualizer, an online visualization utilities to generate previously presented marble diagrams from real code.
+A final story describes yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, R5 used Rx Visualizer, an online visualization utilities to generate marble diagrams from real code.
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
@@
 Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness.
-We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP.
-However, traditional debugging tools allowed them to inspect various parts of a stream, yet they were unable to get more information about, e.g., life cycles of the Observable composition.
-Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem.
-Listing~\ref{lst:rxjs-debugging} exemplifies challenges when debugging a stream of Observables with traditional debugging tools.
+When the engineers needed to interact with these graphs at runtime, e.g.
+to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided with through traditional debugging tools.
+The lack of knowledge about RP abstractions like transformation operators or life cycles was reported as irritating.
+Almost all of the reporters referred to the practice of modifying their source code manually and adding log statements where they assumed a problem to overcome the feature gaps in traditional debuggers.
+Listing~\ref{lst:rxjs-debugging} exemplifies two challenges when debugging a stream of Observables with imperative-focused debugging tools.
 
@@
 Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
-That is why the debuggers step controls cannot operate on the dataflow graph as well; It just does not know this level of abstraction.
+That is why a traditional debuggers step controls cannot operate on the dataflow graph; It just can not interpret this level of abstraction.
 
@@
 
-After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools which ``understand'' RP concepts provided through RxJS.
+After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools which can handle RP concepts provided through RxJS.
 Even though traditional debuggers might help to some extent, they do not provide all the information an engineer is interested in.
@@
 
-We explored how engineers use RP and what techniques they might apply for debugging problems earlier.
+We explored how engineers use RP and what techniques they apply during debugging.
 We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
@@
 
-That is why we saw the demand to validate earlier statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
+That is why we saw demand in validating these statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
 
@@
 Hence we decided to conduct the experiment in a somewhat uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
-Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention the hypothesis by intention.
+Our objective for the experiment was communicated as broad as possible to prevent bias: ``We are interested in how \emph{you} debug a problem'' did not mention our hypothesis by intention.
 
@@
 We executed the experiment in two consecutive blocks of 25 minutes each.
-We provided a ZIP file containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
+We provided a ZIP file\footnote{\url{https://github.com/swissmanu/mse-pa1-experiment/archive/v1.0.2.zip}} containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
 Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
@@
 
-After the subjects S1 to S4 got a feel for what the application on front of them should do, all of them used the provided test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
-Two of them used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
+After the subjects got accustomed with the application provided and understood its purpose, all of them used the test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
+S2 and S4 used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
 

### 2020-08-16 paper.tex
Add Study Result Data

@@
 
-\subsection{Wrap Up}
+\subsection{Interpretation and Take Away}
 
@@
 
-We explored how engineers use RP and what techniques they apply during debugging.
-We observed that most participants tend to modify their source code manually with trace logs and alike in order to find underlying problems.
-It is further our own experience that this practice is often not productive since the evaluation of trace information, once available, can be cumbersome and time-consuming.
-Also, removing log statements after the debugging process might leave new bugs in production code if not done carefully.
+Almost all participants from our interviews and war story reports showed a tendency to manually modify source code with trace logs during the hypothesis instrumentation phase when debugging RxJS code.
+It is further our own experience that this practice is often not productive since the evaluation and interpretation of trace information tends to be cumbersome and very time-consuming.
+Also, removing log statements after a successful debugging process might leave new bugs in production code if not done carefully.
 We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
 
-That is why we saw demand in validating these statements about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
+That is why we saw demand in validating this statement about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
 
 \begin{itemize}
-	\item \emph{Hypothesis: If software engineers must solve an RxJS-based problem, then they will modify the code in order to understand its behavior.}
+	\item \emph{Hypothesis: If software engineers must solve an RxJS-based problem, then they will instrument the code manually in order to understand its behavior.}
 \end{itemize}
@@
 
-The after-action survey was provided within 24 hours after a subject's participation in the main part of the study.
-We asked the subjects about (i) the number of years they have experience with RxJS, (ii) if they currently use RxJS on or off their jobs, (iii) in which field (like frontend, backend or others) they use RxJS and finally (iv) which tools and techniques they use to debug RxJS-based code.
+The after-action survey\footnote{\url{https://github.com/swissmanu/mse-pa1-experiment/blob/f70102885be86fb2323b9516005e1d6dfeb9795b/after-action-survey-questions.md}} was provided within 24 hours after a subject's participation in the main part of the study.
+We asked the subjects about (Q1) if they currently use RxJS on or off their jobs, the (Q2) number of years they have experience with RxJS, in (Q3) which field (like frontend, backend or others) they use RxJS and finally (Q4) which tools and techniques they use to debug RxJS-based code.
 The respective answers allowed us to put the observed actions into perspective and detect potential irregularities in case a subject acted differently as they would have in a ``real'' situation.
@@
 
-After the subjects got accustomed with the application provided and understood its purpose, all of them used the test suite to check what features do not work as expected and tried to recreate the failing behavior in the UI manually.
-S2 and S4 used the traditional debugging tools provided by their browser or IDE to step through parts of the code trying to understand what triggered the failing tests.
+After the subjects got themselves accustomed with the application provided and understood its purpose, all of them used the test suite to \emph{gather context} about what features do not work as expected initially.
+Further, all of them tried to recreate the failing behavior in the UI manually.
+We could not observe any of the subjects using external tools to inspect specific code parts in isolation in later iterations of the debugging process.
+Though, S2 noted that they would have usually started to decompose the problem into smaller pieces and observe their behavior in specific after the 25 minutes of the second block expired.
 
-For both problems given, we could observe all subjects to make use of manually added trace log statements either within an existing arrow function or using the \mintinline{TypeScript}{tap} operator.
-Neither did any of them use additional external tools where they put parts of the code for further inspection nor did they install additional debugging libraries like rxjs-spy.
-After the 25 minutes of the second block exceeded, S2 noted that they would have usually started now to decompose the problem into smaller pieces and observe their behavior more isolated.
+While all subjects added manual trace log statements to existing arrow functions or by adding \mintinline{TypeScript}{tap} operators in the \emph{instrument hypothesis} phase, none of them used additional libraries like rxjs-spy for doing so.
+S2 and S4 used the traditional debugging tools provided by their browser or IDE to add breakpoints.
+Both of them commented on the inability of stack traces to interpret RP abstractions as unsatisfying.
+We could further observe a ``trail-and-error'' approach in later iterations of the debugging process.
+The subjects started to introduce modifications to the system which they immediately \emph{tested} against their latest hypothesis.
+Table~\ref{tab:subject-techniques} provides an overview on the complete collected data regarding used technniques and tools.
 
-The survey responses showed that 75\% (S2 to S4) of the subject population had two or more years of experience with RxJS.
+\begin{table}[H]
+	\begin{tabular}{ p{0.2\linewidth}  p{0.225\linewidth}  p{0.225\linewidth} p{0.225\linewidth} }
+		\toprule
+		\textbf{Subject} & \textbf{Trace Logs} & \textbf{Debugger} & \textbf{Add. Tools} \\ \midrule
+		\textbf{S1}      & \small{X}           &                   &                     \\ \midrule
+		\textbf{S2}      & \small{X}           & \small{X}         &                     \\ \midrule
+		\textbf{S3}      & \small{X}           &                   &                     \\ \midrule
+		\textbf{S4}      & \small{X}           & \small{X}         & \small{Next step}   \\ \bottomrule
+	\end{tabular}
+	\caption{Observed practices and tool usage per subject.}
+	\label{tab:subject-techniques}
+\end{table}
+
+Only S4 was able to solve the first problem given as shown in Table~\ref{tab:subject-results}.
+None of the subjects was able to successfully identify and fix the bugs hidden in the second problem within time.
+
+\begin{table}[H]
+	\begin{tabular}{ p{0.2\linewidth}  p{0.355\linewidth}  p{0.355\linewidth} }
+		\toprule
+		\textbf{Subject} & \textbf{Problem 1} & \textbf{Problem 2} \\ \midrule
+		\textbf{S1}      & \small{Time expired}       & \small{Time expired}     \\ \midrule
+		\textbf{S2}      & \small{Time expired}       & \small{Time expired}     \\ \midrule
+		\textbf{S3}      & \small{Time expired}       & \small{Time expired}     \\ \midrule
+		\textbf{S4}      & \small{Solved}             & \small{Time expired}     \\ \bottomrule
+		\end{tabular}
+	\caption{Results per subject for each presented problem.}
+	\label{tab:subject-results}
+\end{table}
+
+The survey responses available in Table~\ref{tab:subject-survey} showed that 75\% (S2, S3 and S4) of the subject population had two or more years of experience with RxJS.
 Where all of them use RxJS to develop frontend applications, S4 declared having used  RxJS for backend development as well.
@@
 S1 and S3 leverage additional tracing functionality of rxjs-spy, and all four of our subjects use manual log statements.
+
+\begin{table}[H]
+	\begin{tabular}{ p{0.13\linewidth}  p{0.08\linewidth}  p{0.12\linewidth} p{0.23\linewidth} p{0.28\linewidth}}
+		\toprule
+		\textbf{Subject} & \textbf{Q1} & \textbf{Q2}         & \textbf{Q3}               & \textbf{Q4}                             \\ \midrule
+		\textbf{S1}      & \small{Yes} & \small{1 year}      & \small{Frontend}          & \small{Trace Logs, rxjs-spy}            \\ \midrule
+		\textbf{S2}      & \small{No}  & \small{$>$ 3 years} & \small{Frontend}          & \small{Debugger, Trace Logs}            \\ \midrule
+		\textbf{S3}      & \small{Yes} & \small{2 years}     & \small{Frontend}          & \small{Debugger, Trace Logs, rxjs-spy}  \\ \midrule
+		\textbf{S4}      & \small{No}  & \small{2 years}     & \small{Frontend, Backend} & \small{Debugger, Trace Logs}            \\ \bottomrule
+	\end{tabular}
+	\caption{After-action survey responses per subject.}
+	\label{tab:subject-survey}
+\end{table}
 

### 2020-08-16 paper.tex
Rewrite RQ1 Answer

@@
 
-\subsection{Interpretation and Take Away}
+\subsection{Insights}
 
@@
 
-We were able to observe how all subjects modified the provided source code manually and augmented it with trace logs.
-All of them used the produced information in order to track down problems in the application.
-We could not observe the extraction to and reintegration from an external tool, which would have been yet another form of code modification.
+We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs.
+Only 50\% of them used traditional debugging utilities in order to inspect the programs state at runtime.
+All subjects used the new information gained to refine their hypothesis about underlying problems before they started a new iteration in the debugging process.
+We could not observe the extraction to and reintegration from an external tool.
 
-For any subject, we could not identify any contradiction between the observed behavior and the answers given within the after-action survey.
+Even though two subjects stated in the after-action survey to regularly use rxjs-spy for debugging RxJS programs, neither S1 nor S3 made use of this library during the experiment part of the study.
+That aside, we could not identify any contradiction between the observed behavior and the answers provided in the survey.
 
-All reviewed subjects exhibited the debugging behavior described in the initial hypothesis, which we can accept based on the collected evidence.
+All reviewed subjects exhibited the debugging behavior described in our hypothesis, which we can accept based on the collected evidence.
 
-Interviewing professionals, evaluating the war stories, the hands-on experiences, and the results from the observational study allowed us to identify three main challenges, software engineers have to overcome when working with RxJS within an application: (i) They want to \emph{understand how and what data flows} through the graph described by the RP DSL, which helps to identify more straightforward problems most of the time.
-Though, knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables.
-(ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a \emph{deeper understanding of the RP runtimes internal graph-based model}.
-This circumstance makes it hard for engineers to retrace dataflows and computations.
-And finally, (iii) even though there are partial solutions available for the two previously described challenges, they usually require the software engineers either to \emph{manually modify source code or extracting parts of it} to an external utility.
-These additional efforts cost time and bring their own risks since the production code needs to be touched without adding actual value.
+Interviewing professionals, consolidating RxJS hands-on experiences from the war stories, and evaluating the results from our observational study showed us that software engineers use a variety of practices, tools and utilities to debug RP programs.
+Beside the habit of adding trace logs manually, we saw them evidently trying to answer their debugging hypotheses using traditional, imperative-focused debugger utilities.
+The later way of debugging was repeatedly commented as unsatisfying as these utilities cannot handle RP constructs and with this, cannot help to detect problems located within these at all.
+The former way, the introduction of manual log statements, was often described as ``the last resort'' when no other debugging technique helped before.
+This is because the analysis of log entries gets tedious as the number of those increases.
+
+We heard further how engineers isolate specific Observables from bigger dataflows and how they inspect those in sandboxed environments and visualizers.
+This helps them understanding lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
+
+We are able to postulate an answer to our first research question based on this collected information: The biggest challenge for software engineers debugging RxJS-based programs is knowing \emph{when} to apply \emph{what} tool in order to get the \emph{best possible} answer on their currently tested hypothesis.
+We further claim that, backed by the observation where two subjects knew about rxjs-spy but refrained from using it nonetheless, it is not enough to \emph{know} about the existence of a specific tool for a particular debugging situation.
+Such a tool should rather offer itself at the ``right'' moment so the engineer can benefit from its features.
+An example for such a behavior is the seamless integration of asynchronous call stacks in Google Chromes JavaScript debugger\cite{chromeasync}: Even though code was executed ``outside'' of the current call stack, the debugger shows these asynchronous stack frames as they were actually part of it.
 

### 2020-08-19 paper.tex
v0.12
Fix v0.12 Review Findings

@@
 
-\newcommand{\version}{v0.11}
+\newcommand{\version}{v0.12}
 
@@
 To implement the dataflows as shown in Figure~\ref{fig:ui-data-flows} to drive a UI, the Observer design pattern\cite{gamma1995design} is often used and variations of the pattern are omnipresent today.
-E.g., a modified version of the pattern can be found in the event handling interface of the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications} implemented by internet browsers.
 
@@
 
-Where the standard glossary defines \emph{debug} as an activity ``to detect, locate, and correct faults in a computer program''\cite{ieeeglossary} in general, Layman et al.
-\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized debugging as an iterative hypothesis refinement process, as shown in Figure~\ref{fig:debugging-process-model}:
+Layman et al.
+\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} use the debugging process model, an iterative, hypothesis refinement process, to formalize the general activity of debugging a computer program in their paper.
 
@@
 
-Software engineers have an arsenal of tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
+Software engineers have many tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
 Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.
 
-We will use three main concepts to categorize debugging utilities in this paper: Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
+Many of these specialized utilities differentiate themselves fundamentally in regards of the concepts they are built upon. We identified and will use the following three categories to structure them:
+
+ Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
 Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
 
-RP provides its challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
+RP provides its own set of challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
@@
 RxJS\cite{rxjs} is the reference implementation of the ReactiveX API specification for JavaScript.
-Its current major version 6 is implemented using TypeScript and is used thoroughly by large projects like Angular\cite{angualrrxjs}.
+Its current major version 6 is implemented using TypeScript and is used by large projects like Angular\cite{angualrrxjs}.
 Listing \ref{lst:rp-with-rxjs} shows an example of RP using RxJS in TypeScript.
@@
 The RxJS community uses \emph{marble diagrams} as shown in Figure~\ref{fig:marble-diagram} to document \cite{marblediagrams} the runtime behavior of an Observable visually.
-Unit test libraries\cite{marbletesting} leverage on this abstraction to encode the behavior of mocked Observables or to describe assertions.
+Unit test libraries\cite{marbletesting} use this abstraction to encode the behavior of mocked Observables or to describe assertions.
 
@@
 
-We organized informal interviews which allowed us to gain a first insight in how software engineers work with RP in their daily jobs.
+We organized informal interviews which allowed us to gain insight in how software engineers work with RP in their daily jobs.
 We talked to five engineers (following identified using the codes \emph{I1} through \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
@@
 
-The reports provided us with diverse perspectives on various areas of software development using RxJS.
-R3 focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
+In their report, R3 focused on how they built code with improved testability because of recent changes in RxJS: ``[\dots] in the beginning, it was very hard to write asynchronous tests [\dots].
 I really disliked [\dots] you were forced to pass a TestScheduler''.
@@
 On top, each data source could re-emit updated versions of previously requested information at any time.
-After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed information kept changing/flickering (where we expected it to change exactly once and stay), other data got not updated at all when it was supposed to,'' they told us.
+After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed numbers kept changing where we did not expected them to.
+In other places, they were not rerendered where they were supposed to after, e.g.
+after we changed them in the system,'' they told us.
 The browsers debugger tools and its breakpoints did not help much since the operators were executed several times.
@@
 
-Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness.
-When the engineers needed to interact with these graphs at runtime, e.g.
+Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness, as we heard in the interviews.
+When the engineers needed to interact with dataflow graphs at runtime, e.g.
 to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided with through traditional debugging tools.
@@
 
-We further learned that log statements as shown on Line~4 in Listing~\ref{lst:rxjs-debugging} indeed help to trace emitted values, but do not expose any information about the life cycle of an Observable.
-The reporter describing their usage of rxjs-spy explained, that the knowledge about when an Observable gets subscribed, unsubscribed or completed helped a lot in order to solve problems related to higher-order Observables.
+We learned that simple trace augmentation, like on Line~4 in Listing~\ref{lst:rxjs-debugging}, logs emitted values only.
+This might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying Observable completely.
+The importance of such information was confirmed by the reporter describing their usage of rxjs-spy: The knowledge about the Observable lifecycle, like when it gets subscribed, unsubscribed or completed, helped them to solve problems multiple times.
+When dealing with higher-order Observables, they value this information even as indispensable.
 
@@
 
-After the subjects got themselves accustomed with the application provided and understood its purpose, all of them used the test suite to \emph{gather context} about what features do not work as expected initially.
+After the subjects got themselves accustomed with the application provided and understood its purpose, all of them used the test suite to gather context about what features do not work as expected initially.
 Further, all of them tried to recreate the failing behavior in the UI manually.
-We could not observe any of the subjects using external tools to inspect specific code parts in isolation in later iterations of the debugging process.
+We could not observe any of the subjects using external tools, e.g.
+RxViz, to inspect specific code parts in isolation in later iterations of the debugging process.
 Though, S2 noted that they would have usually started to decompose the problem into smaller pieces and observe their behavior in specific after the 25 minutes of the second block expired.
 
-While all subjects added manual trace log statements to existing arrow functions or by adding \mintinline{TypeScript}{tap} operators in the \emph{instrument hypothesis} phase, none of them used additional libraries like rxjs-spy for doing so.
+While all subjects added manual trace log statements to existing arrow functions or by adding \mintinline{TypeScript}{tap} operators in the instrument hypothesis phase, none of them used additional libraries like rxjs-spy for doing so.
 S2 and S4 used the traditional debugging tools provided by their browser or IDE to add breakpoints.
@@
 We could further observe a ``trail-and-error'' approach in later iterations of the debugging process.
-The subjects started to introduce modifications to the system which they immediately \emph{tested} against their latest hypothesis.
-Table~\ref{tab:subject-techniques} provides an overview on the complete collected data regarding used technniques and tools.
+The subjects started to introduce modifications to the system which they immediately tested against their latest hypothesis.
+Table~\ref{tab:subject-techniques} provides an overview on the complete collected data regarding used techniques and tools.
 
@@
 Where all of them use RxJS to develop frontend applications, S4 declared having used  RxJS for backend development as well.
-When asked what tools they usually use for debugging, only S1 stated not to use the traditional debugger of their IDE at all.
+When asked what tools they usually use for debugging, S2, S3 and S4 stated to use the traditional debugger of their IDE.
 S1 and S3 leverage additional tracing functionality of rxjs-spy, and all four of our subjects use manual log statements.
@@
 We could not observe the extraction to and reintegration from an external tool.
+All subjects exhibited the debugging behavior described in our hypothesis.
 
 Even though two subjects stated in the after-action survey to regularly use rxjs-spy for debugging RxJS programs, neither S1 nor S3 made use of this library during the experiment part of the study.
-That aside, we could not identify any contradiction between the observed behavior and the answers provided in the survey.
-
-All reviewed subjects exhibited the debugging behavior described in our hypothesis, which we can accept based on the collected evidence.
 
@@
 
-We are able to postulate an answer to our first research question based on this collected information: The biggest challenge for software engineers debugging RxJS-based programs is knowing \emph{when} to apply \emph{what} tool in order to get the \emph{best possible} answer on their currently tested hypothesis.
+We are able to postulate an answer to our first research question based on this collected information: \textbf{TODO Haben wir wirklich evidenz?
+Wenig diskutiert?} The biggest challenge for software engineers debugging RxJS-based programs is knowing \emph{when} to apply \emph{what} tool in order to get the \emph{best possible} answer on their currently tested hypothesis.
 We further claim that, backed by the observation where two subjects knew about rxjs-spy but refrained from using it nonetheless, it is not enough to \emph{know} about the existence of a specific tool for a particular debugging situation.

### 2020-08-19 paper.tex
Restructure Prev Efforts

@@
 
+Before we elaborate on a solution in order to overcome the challenges identified with RQ1, we are going to present a selection of RP specific debugging utilities available today.
+
 \section{Discussion of Previous Efforts}
@@
 
-Our work is not the first attempt to understand and improve on the struggles software engineers have to overcome when working on RP code, either with or without RxJS.
-We are going to present a selection of four utilities, whereas two of them have an academic background.
-We learned about them during our research either because they were mentioned in one of the interviews and war stories, or because we have worked with them in the past.
-
-\subsection{Reactive Inspector}
-
-Salvaneschi et al.
-did focus their research heavily on RP during the mid 2010s.
-Besides the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they have looked into the debugging habits of software engineers as well.
-In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al.
-elaborate on the necessity of RP debugging tools that understand the specific language abstractions.
-They provide with the \emph{Reactive Inspector} plugin for the Eclipse IDE a reactive debugger for the REScala RP runtime environment.
-One of its main features is the visualization of the internal dataflow graph.
-A new class of  breakpoints, the ``reactive breakpoints,'' builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather than on a specific node in the graph.
-The execution of the program can be interrupted e.g., once a node retains a specific output value or an event of interest takes place in the runtime.
-Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without re-execution.
-
-Salvaneschi et al. claim\cite{10.1145/2577080.2577083} the architecture behind \emph{Reactive Inspector} to be flexible enough to be used with other RP runtime environments.
-
-\subsection{rxfiddle}
-
-\emph{rxfiddle} as proposed by Banken et al.
-is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}.
-The design defines two independent parts: The (i) \emph{Host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., when an Observable is created or subscribed) and collects them.
-The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables.
-A marble diagram visualizes the values emitted over time for every Observable further.
-
-The reference implementation supports event collection for RxJS-based code in versions 4 and 5.
-It is available as an online application where snippets of code can be inspected.
-A proof-of-concept implementation which works on a local development machine is accessible via the projects Git repository.
-
-The architectural design of \emph{rxfiddle} is similar to the one of \emph{Reactive Inspector}: Banken et al. claim that their Visualizer can be adapted to different RP runtime environments as well.
-
 \subsection{rxjs-spy}
 
-Software engineers tend to augment their code with additional log trace statements where they have to track down a problem we saw earlier.
-rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator, along with sophisticated log monitoring facilities.
-Ideally, ``tagged'' Observables are created during development when the dataflows are composed for the first time.
-Like the \mintinline{TypeScript}{tap} operator, a tag as shown in Listing~\ref{lst:rxjs-spy-tag} on Line~3 does not influence runtime behavior of an application.
+\emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables.
+Once an observable is tagged with an arbitrary string identifier, the library can generate trace logs when a value is emitted as well as when individual lifecycle events occur (i.e.
+subscribe, unsubscribe, complete and error).
+Ideally, tagged Observables are created during development when the dataflows are composed for the first time.
+Like the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-debugging}, the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-spy-tag} on Line~3 is completely transparent to the actual dataflow.
 
@@
 		  map(i => i - 1),
+		  tag('subtract')      // Tag Observable with "subtract"
 		  take(2)
@@
 	\end{minted}
-	\caption{Usage of \emph{rxjs-spy} \mintinline{TypeScript}{tag} Operator on Line~3}
+	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~9 and 11}
 	\label{lst:rxjs-spy-tag}
@@
 
-Once necessary, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console (see Listing~\ref{lst:rxjs-spy-log}) without touching any Observable code at all.
-The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.
+The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log}.
 
@@
 	\end{minted}
-	\caption{Trace log generated by \emph{rxjs-spy} \mintinline{TypeScript}{tag} from Listing~\ref{lst:rxjs-spy-tag}}
+	\caption{rxjs-spy execution trace log generated from Listing~\ref{lst:rxjs-spy-tag}}
 	\label{lst:rxjs-spy-log}
@@
 Additional features are available through the debuggers console interface.
-Engineers can ``pause'' a tagged Observable, so it collects incoming values.
-They can then be emitted one after another manually or resumed at once.
+E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately.
+The engineer can then emitted these values one after another manually or resume all of them at once.
 
-\subsection{Rx Visualizer}
+\subsection{rxfiddle}
 
-The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser.
-It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.
-Diagrams are generated over a given, fixed time and available for download as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}} files.
+\emph{rxfiddle} as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
+They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
+The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables, whereas a marble diagram visualizes the values emitted over time for every Observable.
 
-The visualizer helps to understand the behavior of a specific part of an Observable stream in a sandboxed environment as a reported war story stated.
+The reference implementation supports event processing for the (outdated) RxJS major versions 4 and 5 only and is available as an online application\footnote{\url{https://rxfiddle.net/}}.
+A proof-of-concept implementation which can be executed on a local computer is available through the projects Git repository\footnote{\url{https://github.com/hermanbanken/RxFiddle}}.
+
+\subsection{RxViz}
+
+\emph{RxViz} is a visualizer utility available online\footnote{\url{https://rxviz.com/}}.
+It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
+Engineers implement or copy-paste dataflows in an editor window using JavaScript.
+This code is then used to generate a diagram over a configurable time interval.
+The diagrams are rendered immediately and are available as downloadable SVG files.
+
+As of writing this paper, the projects owner stopped working actively on the open source application and is looking for a new maintainer\cite{rxviz-maintainer}.
+
+\subsection{Reactive Inspector}
+
+In the paper ``Debugging for Reactive Programming''\cite{10.1145/2884781.2884815}, Salvaneschi et al.
+elaborate on the necessity of RP debugging tools which can interpret its specific dataflow abstractions.
+Based on their research, developed a reactive debugger named \emph{Reactive Inspector}.
+This plugin for the Eclipse IDE provides reactive debugging functionality for programs using the REScala RP runtime.
+
+The plugins core feature is the visualization of the dataflow graph.
+This representation of the dataflow provides a UI to add a new kind of breakpoint, the ``reactive breakpoint,'' to individual nodes in the graph.
+With them, the execution of the program can be interrupted e.g., when a node retains a specific value or a particular lifecycle event takes place.
+
+Along with the RP specific debugging features, Reactive Inspector experiments with the omniscient debugging concept: The time travel functionality allows to record dataflows.
+These recordings can then be replayed and inspected at a later time without actual re-execution of the program.
 
 \subsection{Comparison}
+
+\textbf{TODO REWRITE HERE}
 
@@
 
+\textbf{TODO REWRITE HERE}
+
 Having the main challenges for engineers identified when debugging RxJS-code, we can propose areas to improve on now and resolve RQ2 by doing so.
@@
 A graphical representation of the dataflow graph is better suited to show dependencies within the model then textual trace logs.
-Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively \cite{10.1145/3180155.3180156} on one side, on the other also not easily implement efficiently as well.
+Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively\cite{10.1145/3180155.3180156} on one side, on the other also not easily implement efficiently as well.
 
@@
 
+\textbf{TODO REWRITE HERE}
+
 We worked with 14 professional software engineers while writing this paper.
@@
 \label{sec:conclusion}
+
+\textbf{TODO REWRITE HERE}
 

### 2020-08-20 paper.tex
Switch to acmart

@@
-\documentclass[12pt,a4paper]{article}
+\documentclass[sigplan,screen,review]{acmart}
 \usepackage{ccicons}
@@
 \usepackage{booktabs}
-
 \usepackage{./tikzit}
@@
 
-\newcommand{\version}{v0.12}
+% Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
+\acmConference[SPLASH '20]{SPLASH '20: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
+\title{Debugging of RxJS-based Applications}
+\author{Manuel Alabor}
+\affiliation{
+	\institution{HSR University of Applied Sciences}
+	\city{Rapperswil}
+	\country{Switzerland}
+}
+\email{malabor@hsr.ch}
 
 \begin{document}
-\begin{centering}
 
-	\Large{Workshop Paper}
-	\par
-	\Huge{Debugging of RxJS-based Applications}
-	\par
-	\vspace{2ex}
+\begin{abstract}
+	\input{./abstract.tex}
+\end{abstract}
 
-	\normalsize{
-		Manuel Alabor\\
-		Supervised by Prof. Dr. Markus Stolze\\
-		\par
-		\vspace{2ex}
-		HSR University of Applied Sciences Rapperswil, Switzerland\\
-		\par
-		\vspace{2ex}
-		\today{} (\version)
-	}
-	\par
-	\vspace{2ex}
+\begin{CCSXML}
+	<ccs2012>
+	   <concept>
+		   <concept_id>10011007</concept_id>
+		   <concept_desc>Software and its engineering</concept_desc>
+		   <concept_significance>500</concept_significance>
+		   </concept>
+	 </ccs2012>
+\end{CCSXML}
 
-	\begin{quotation}
-		\small{
-			\noindent\textbf{Abstract}---\input{./abstract.tex}
-		}
-		\par
-		\vspace{4ex}
-	\end{quotation}
-\end{centering}
+\ccsdesc[500]{Software and its engineering}
+
+\keywords{reactive programming, debugging, empirical software engineering}
+
+\maketitle
 
@@
 
-\begin{figure}[H]
+\begin{figure}
 	\centering
@@
 	\end{tikzpicture}
+	\Description{Basic Dataflows in a UI Application.}
 	\caption{Basic Dataflows in a UI Application.}
@@
 	\tikzfig{debugging-process-model}
+	\Description{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
 	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
@@
 	\end{minted}
+	\Description{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
 	\caption{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
@@
 	\tikzfig{marble-diagram}
+	\Description{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
+From left to right, each marble represents an emitted value of 0, 2, 4, and 6.
+The vertical line at the last marble indicates that the Observable completed after emitting value \mintinline{TypeScript}{6}.}
 	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
@@
 
-\begin{table}[H]
+\begin{table}
 	\begin{tabular}{ p{0.2\linewidth}  p{0.225\linewidth}  p{0.225\linewidth} p{0.225\linewidth} }
@@
 
-\begin{table}[H]
+\begin{table}
 	\begin{tabular}{ p{0.2\linewidth}  p{0.355\linewidth}  p{0.355\linewidth} }
@@
 
-\begin{table}[H]
+\begin{table}
 	\begin{tabular}{ p{0.13\linewidth}  p{0.08\linewidth}  p{0.12\linewidth} p{0.23\linewidth} p{0.28\linewidth}}
@@
 The diagrams are rendered immediately and are available as downloadable SVG files.
-
-As of writing this paper, the projects owner stopped working actively on the open source application and is looking for a new maintainer\cite{rxviz-maintainer}.
 
@@
 
-\bibliographystyle{splncs04}
-\bibliography{bibliography.bib}
+\begin{acks}
+	We would like to
+\end{acks}
 
-\section*{License}
-\ccby\thinspace\thinspace This work is licensed under a \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International License}.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibliography}
+
 \end{document}

### 2020-08-20 paper.tex
Fix Overlowing Elements

@@
 	\centering
-	\begin{tikzpicture}[node distance = 5.5cm, auto]
-		\tikzstyle{block} = [rectangle, draw,
-		text width=5em, text centered, rounded corners, minimum height=2.5em]
-
-		\node [block] (user) {User};
-		\node [block, right of=user] (ui) {UI};
-		\node [block, right of=ui] (system) {System};
-
-		\draw [<->] (user) to node {Interaction} (ui);
-		\draw [->] (ui.north) to[bend left=20] node {Interaction Data Flow} (system.north);
-		\draw [->] (system.south) to[bend left=20] node {Update Data Flow} (ui.south);
-	\end{tikzpicture}
-	\Description{Basic Dataflows in a UI Application.}
-	\caption{Basic Dataflows in a UI Application.}
+	\tikzfig{ui-data-flow}
+	\Description{Schematic visualization of a basic UI application.
+The user interacts with the UI.
+The UI component interprets and forwards interactions to the system.
+The system sends update notifactions to the UI eventually.}
+	\caption{Basic dataflows in a UI Application.}
 	\label{fig:ui-data-flows}
@@
 
-\begin{listing}[H]
+\begin{listing}
 	\begin{minted}{TypeScript}
@@
 
-		of(0, 1, 2, 3, 4).pipe(	// Create Observable with Integers
-			filter(i => i < 4),		// Omit Integers >= 4
-			map(i => i * 2)				// Multiply Integer with 2
-		).subscribe(console.log) // Emits 0, 2, 4, 6
+		of(0, 1, 2, 3, 4).pipe(  // Create Observable
+			filter(i => i < 4),	  // Omit integers >= 4
+			map(i => i * 2)        // Multiply int. by 2
+		).subscribe(console.log) // Logs: 0, 2, 4, 6
 	\end{minted}
@@
 
-\begin{listing}[H]
+\begin{listing}
 	\begin{minted}{TypeScript}
-		interval(1000).pipe(    // Increasing numbers with 1s delay
-		  map(i => i * 2),      // Breakpoint possible within i * 2
-		  take(4),              // No breakpoint possible
-		  tap(console.log)      // Manually added log statement
+		interval(1000).pipe(
+		  map(i => i * 2),
+		  take(4),
+		  tap(console.log)
 		).subscribe(showValue); // Emits: 0, 2, 4, 6
 	\end{minted}
-	\caption{Debugging of an RxJS Observable using breakpoints and log statements.}
+	\caption{An Observable emitting a sequence of increasing integers every second.
+Traditional breakpoints are possible inside the arrow function on Line~2.
+Though a breakpoint can be added on Line~3, it will never be hit during the actual execution of the \mintinline{TypeScript}{take} operator.
+Line~4 shows a manually introduced trace log statement using the \mintinline{TypeScript}{tap} side effect operator.}
 	\label{lst:rxjs-debugging}
@@
 	\end{minted}
-	\caption{Call Stack for Arrow Function on Line~2 in Listing~\ref{lst:rxjs-debugging}}
+	\caption{A call stack showing the internal RxJS execution stack for a breakpoint in the arrow function on Line~2 in Listing~\ref{lst:rxjs-debugging}.}
 	\label{lst:rxjs-call-stack}
@@
 \begin{table}
-	\begin{tabular}{ p{0.2\linewidth}  p{0.225\linewidth}  p{0.225\linewidth} p{0.225\linewidth} }
+	\caption{Observed practices and tool usage per subject.}
+	\begin{tabular}{ p{0.16\linewidth}  p{0.225\linewidth}  p{0.225\linewidth} p{0.225\linewidth} }
 		\toprule
@@
 	\end{tabular}
-	\caption{Observed practices and tool usage per subject.}
 	\label{tab:subject-techniques}
@@
 \begin{table}
-	\begin{tabular}{ p{0.2\linewidth}  p{0.355\linewidth}  p{0.355\linewidth} }
+	\caption{Results per subject for each presented problem.}
+	\begin{tabular}{ p{0.16\linewidth}  p{0.355\linewidth}  p{0.355\linewidth} }
 		\toprule
@@
 	\end{tabular}
-	\caption{Results per subject for each presented problem.}
 	\label{tab:subject-results}
@@
 \begin{table}
-	\begin{tabular}{ p{0.13\linewidth}  p{0.08\linewidth}  p{0.12\linewidth} p{0.23\linewidth} p{0.28\linewidth}}
+	\caption{After-action survey responses per subject.}
+	\begin{minipage}{\columnwidth}
+	\begin{tabular}{ p{0.12\linewidth}  p{0.06\linewidth}  p{0.14\linewidth} p{0.20\linewidth} p{0.26\linewidth}}
 		\toprule
@@
 	\end{tabular}
-	\caption{After-action survey responses per subject.}
+	\end{minipage}
 	\label{tab:subject-survey}
@@
 		const spy = create(); // Create monitor
-		spy.log(/multiply/);   // Log tags matching RegEx /multiply/
+		spy.log(/multiply/);  // Log tags matching
+		                      // RegEx /multiply/
 
@@
 		  map(i => i * 2),
-		  tag('multiply'),     // Tag Observable with "multiply"
+		  tag('multiply'),    // Tag with "multiply"
 		  map(i => i - 1),
-		  tag('subtract')      // Tag Observable with "subtract"
+		  tag('subtract')     // Tag with "subtract"
 		  take(2)
@@
 	\end{minted}
-	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~9 and 11}
+	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~10 and 12.}
 	\label{lst:rxjs-spy-tag}
@@
 	\end{minted}
-	\caption{rxjs-spy execution trace log generated from Listing~\ref{lst:rxjs-spy-tag}}
+	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
 	\label{lst:rxjs-spy-log}
@@
 \begin{table}[H]
+	\caption{Feature Matrix for Discussed RP Debugging Tools}
 	\begin{tabular}{@{}lllll@{}}
@@
 	\end{tabular}
-	\caption{Feature Matrix for Discussed RP Debugging Tools}
 	\label{tab:compare-tools}

### 2020-08-20 paper.tex
Restructure Intro

@@
 \documentclass[sigplan,screen,review]{acmart}
-\usepackage{ccicons}
+% Class options:
+%  - review: Show line numbers for reviewers
+%  - nonacm: Hide ACM conference stuff
+
 \usepackage{hyperref}
 \usepackage{minted}
-\usepackage{graphicx}
 \usepackage{booktabs}
@@
 % Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
-\acmConference[SPLASH '20]{SPLASH '20: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
+\acmConference[SPLASH '20]{SPLASH 2020: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
 \title{Debugging of RxJS-based Applications}
@@
 
-According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.
+According to the IEEE Standard Glossary of Software Engineering, \emph{debugging} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.
 
@@
 
-These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry.
+These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow-specific inquiry.
 Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities.
@@
 
-We will conclude this introductory section with the clarification of important terms.
+We will conclude this introductory section with the clarification of important terms and a view on known RP debugging utilities.
 Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports.
-We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RP code in Section~\ref{sec:discussion}.
-Finally, we will answer RQ2 in Section~\ref{sec:future} ``Future Work''.
+We present our observational study intended to validate former insights in Section~\ref{sec:study} which allow us to answer RQ1.
+Before our final conclusion, we will answer RQ2 in Section~\ref{sec:future} ``Future Work'' and review the threats to validity regarding our study in Section~\ref{sec:threats}.
 
@@
 It then takes care that values are processed and propagated correctly through them as well as that a consistent system state\cite{10.1145/2501654.2501666} is always maintained.
-
-\subsection{Debugging Process Model}
-
-Layman et al.
-\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} use the debugging process model, an iterative, hypothesis refinement process, to formalize the general activity of debugging a computer program in their paper.
-
-\begin{figure}[H]
-	\centering
-	\tikzfig{debugging-process-model}
-	\Description{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
-	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
-	\label{fig:debugging-process-model}
-\end{figure}
-
-The process consists of three steps and includes a feedback loop: After the engineer (i) gathered sufficient context information (e.g.
-ways to reproduce the failure or details about external factors) and understands the situation satisfactory, they generate a hypothesis on the origins of the bug or what impact a change made to the program might have.
-With the intent to proof their hypothesis, the engineer then (ii) instruments the defective program using suitable tools (e.g.
-adding log statements, setting breakpoints or removing code parts).
-Finally, the instrumented system gets (iii) challenged against the formed hypothesis.
-E.g., code statements are executed step by step using a debugger or trace logs are analyzed and compared against expected behavior.
-If the hypothesis turns out to be correct, the debugging process stops.
-If not, the newly gained knowledge about the problem is used to build a refined hypothesis and start a new iteration.
-
-\subsection{Debugger Concepts}
-
-Software engineers have many tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
-Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.
-
-Many of these specialized utilities differentiate themselves fundamentally in regards of the concepts they are built upon. We identified and will use the following three categories to structure them:
-
-Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
-Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
-
-RP provides its own set of challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
-Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
-A (ii) \emph{reactive debugger} can interpret the underlying graph model of a RP runtime system.
-It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
-
-Traditional, as well as reactive debuggers work with the \emph{current} state of a programs execution only.
-They lack information about what happened before or what is going to happen in the future.
-This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system.
-An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
-Instead, it records runtime telemetry and provides an interface for later inspection.
-Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
 
@@
 
+\subsection{Debugging Process Model}
+
+Layman et al.
+\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} use the debugging process model, an iterative, hypothesis refinement process, to formalize the general activity of debugging a computer program in their paper.
+
+\begin{figure}
+	\centering
+	\tikzfig{debugging-process-model}
+	\Description{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
+	\caption{Debugging Process Model after Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013}}
+	\label{fig:debugging-process-model}
+\end{figure}
+
+The process consists of three steps and includes a feedback loop: After the engineer (i) gathered sufficient context information (e.g.
+ways to reproduce the failure or details about external factors) and understands the situation satisfactory, they generate a hypothesis on the origins of the bug or what impact a change made to the program might have.
+With the intent to proof their hypothesis, the engineer then (ii) instruments the defective program using suitable tools (e.g.
+adding log statements, setting breakpoints or removing code parts).
+Finally, the instrumented system gets (iii) challenged against the formed hypothesis.
+E.g., code statements are executed step by step using a debugger or trace logs are analyzed and compared against expected behavior.
+If the hypothesis turns out to be correct, the debugging process stops.
+If not, the newly gained knowledge about the problem is used to build a refined hypothesis and start a new iteration.
+
+\subsection{Debugger Concepts}
+
+Software engineers have many tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
+Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.
+
+Many of these specialized utilities differentiate themselves fundamentally in regards of the concepts they are built upon. We identified and will use the following three categories to structure them:
+
+Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
+Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
+
+RP provides its own set of challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
+Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
+A (ii) \emph{reactive debugger} can interpret the underlying graph model of a RP runtime system.
+It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
+
+Traditional, as well as reactive debuggers work with the \emph{current} state of a programs execution only.
+They lack information about what happened before or what is going to happen in the future.
+This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system.
+An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly.
+Instead, it records runtime telemetry and provides an interface for later inspection.
+Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again.
+
+\subsection{RxJS Debugging Utilities}
+
+\subsubsection{rxjs-spy}
+
+\emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables.
+Once an observable is tagged with an arbitrary string identifier, a monitor can generate trace logs whenever a value is emitted as well as when individual lifecycle events occur (i.e.
+subscribe, unsubscribe, complete and error).
+Ideally, tagged Observables are created during development when the dataflows are composed for the first time.
+Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an Observable emits a value.
+It cannot modify or influence the emitted value nor the Observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual dataflow.
+
+\begin{listing}
+	\begin{minted}{TypeScript}
+		import { create } from 'rxjs-spy';
+		import { tag } from 'rxjs-spy/operators';
+
+		const spy = create(); // Create monitor
+		spy.log(/multiply/);  // Log tags matching
+		                      // RegEx /multiply/
+		interval(1000).pipe(
+		  map(i => i * 2),
+		  tag('multiply'),    // Tag with "multiply"
+		  map(i => i - 1),
+		  tag('subtract')     // Tag with "subtract"
+		  take(2)
+		).subscribe();
+	\end{minted}
+	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~9 and 11.}
+	\label{lst:rxjs-spy-tag}
+\end{listing}
+
+The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log} eventually.
+
+\begin{listing}
+	\begin{minted}{Typescript}
+		Tag = multiply; notification = subscribe
+		Tag = multiply; notification = next; value = 0
+		Tag = multiply; notification = next; value = 2
+		Tag = multiply; notification = unsubscribe
+	\end{minted}
+	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
+	\label{lst:rxjs-spy-log}
+\end{listing}
+
+Additional features are available through the libraries console interface.
+E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately.
+The engineer can then emitted these values one after another manually or resume all of them at once.
+
+\subsubsection{rxfiddle}
+
+\emph{rxfiddle} as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
+They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
+The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables, whereas a marble diagram visualizes the values emitted over time for every Observable.
+
+The reference implementation supports event processing for the (outdated) RxJS major versions 4 and 5 only and is available as an online application\footnote{\url{https://rxfiddle.net/}}.
+A proof-of-concept implementation which can be executed on a local computer is available through the projects Git repository\footnote{\url{https://github.com/hermanbanken/RxFiddle}}.
+
+\subsubsection{RxViz}
+
+\emph{RxViz} is a visualizer utility available online\footnote{\url{https://rxviz.com/}}.
+It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
+Engineers implement or copy-paste dataflows in an editor window using JavaScript.
+This code is then used to generate a diagram over a configurable time interval.
+The diagrams are rendered immediately and are available as downloadable SVG files.
+
 \section{Interviews and War Stories}
@@
 E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs.
-In the end, we were able to gather compelling statements which we are going to present in this section.
+In the end, we were able to conduct interviews with 10 engineers and collect war stories from another five.
 
@@
 
-After we built a first intuition for how software engineers work with RP by evaluating the interviews, we were interested in more RxJS-specific, hands-on experiences.
+After we built an intuition for how software engineers work with RP by evaluating the interviews, we were interested in more RxJS-specific, hands-on experiences.
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
@@
 The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
-(We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion}.)
 
@@
 
-\section{Validation of Interviews and War Stories}
+\section{Validation}
 \label{sec:study}
@@
 
-Before we elaborate on a solution in order to overcome the challenges identified with RQ1, we are going to present a selection of RP specific debugging utilities available today.
-
-\section{Discussion of Previous Efforts}
-\label{sec:discussion}
-
-\subsection{rxjs-spy}
-
-\emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables.
-Once an observable is tagged with an arbitrary string identifier, the library can generate trace logs when a value is emitted as well as when individual lifecycle events occur (i.e.
-subscribe, unsubscribe, complete and error).
-Ideally, tagged Observables are created during development when the dataflows are composed for the first time.
-Like the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-debugging}, the \mintinline{TypeScript}{tap} operator in Listing~\ref{lst:rxjs-spy-tag} on Line~3 is completely transparent to the actual dataflow.
-
-\begin{listing}[H]
-	\begin{minted}{TypeScript}
-		import { create } from 'rxjs-spy';
-		import { tag } from 'rxjs-spy/operators';
-
-		const spy = create(); // Create monitor
-		spy.log(/multiply/);  // Log tags matching
-		                      // RegEx /multiply/
-
-		interval(1000).pipe(
-		  map(i => i * 2),
-		  tag('multiply'),    // Tag with "multiply"
-		  map(i => i - 1),
-		  tag('subtract')     // Tag with "subtract"
-		  take(2)
-		).subscribe();
-	\end{minted}
-	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~10 and 12.}
-	\label{lst:rxjs-spy-tag}
-\end{listing}
-
-The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log}.
-
-\begin{listing}[H]
-	\begin{minted}{Typescript}
-		Tag = multiply; notification = subscribe
-		Tag = multiply; notification = next; value = 0
-		Tag = multiply; notification = next; value = 2
-		Tag = multiply; notification = unsubscribe
-	\end{minted}
-	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
-	\label{lst:rxjs-spy-log}
-\end{listing}
-
-Additional features are available through the debuggers console interface.
-E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately.
-The engineer can then emitted these values one after another manually or resume all of them at once.
-
-\subsection{rxfiddle}
-
-\emph{rxfiddle} as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
-They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
-The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables, whereas a marble diagram visualizes the values emitted over time for every Observable.
-
-The reference implementation supports event processing for the (outdated) RxJS major versions 4 and 5 only and is available as an online application\footnote{\url{https://rxfiddle.net/}}.
-A proof-of-concept implementation which can be executed on a local computer is available through the projects Git repository\footnote{\url{https://github.com/hermanbanken/RxFiddle}}.
-
-\subsection{RxViz}
-
-\emph{RxViz} is a visualizer utility available online\footnote{\url{https://rxviz.com/}}.
-It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
-Engineers implement or copy-paste dataflows in an editor window using JavaScript.
-This code is then used to generate a diagram over a configurable time interval.
-The diagrams are rendered immediately and are available as downloadable SVG files.
-
-\subsection{Reactive Inspector}
-
-In the paper ``Debugging for Reactive Programming''\cite{10.1145/2884781.2884815}, Salvaneschi et al.
-elaborate on the necessity of RP debugging tools which can interpret its specific dataflow abstractions.
-Based on their research, developed a reactive debugger named \emph{Reactive Inspector}.
-This plugin for the Eclipse IDE provides reactive debugging functionality for programs using the REScala RP runtime.
-
-The plugins core feature is the visualization of the dataflow graph.
-This representation of the dataflow provides a UI to add a new kind of breakpoint, the ``reactive breakpoint,'' to individual nodes in the graph.
-With them, the execution of the program can be interrupted e.g., when a node retains a specific value or a particular lifecycle event takes place.
-
-Along with the RP specific debugging features, Reactive Inspector experiments with the omniscient debugging concept: The time travel functionality allows to record dataflows.
-These recordings can then be replayed and inspected at a later time without actual re-execution of the program.
-
-\subsection{Comparison}
-
-\textbf{TODO REWRITE HERE}
-
-Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools.
-Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy).
-They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector).
-Whereas Reactive Inspector for REScala provides an ``all-in-one'' solution with reactive breakpoints, omniscient debugging capabilities, and full IDE integration, the RxJS-related tools provide smaller, differentiated feature sets in contrast.
-Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or call for manual code preparation or modification (rxjs-spy).
-
-\begin{table}[H]
-	\caption{Feature Matrix for Discussed RP Debugging Tools}
-	\begin{tabular}{@{}lllll@{}}
-	\toprule
-					   & \textbf{UI} & \textbf{Integration}                                               & \textbf{Flow Control}                                              & \textbf{RP Runtime} \\ \midrule
-	Reactive Inspector & Visual      & Eclipse IDE                                                        & Full                                                               & REScala             \\ \midrule
-	rxfiddle           & Visual      & Web App                                                            & None       & RxJS                \\ \midrule
-	rxjs-spy           & Textual     & \begin{tabular}[c]{@{}l@{}}Browser Console\\and Code\end{tabular} & Limited & RxJS                \\ \midrule
-	Rx Visualizer      & Visual      & Web App                                                            & None       & RxJS                \\ \bottomrule
-	\end{tabular}
-	\label{tab:compare-tools}
-\end{table}
-
 \section{Future Work}
@@
 
+\textbf{TODO Refer to Reactive Inspector}
+
+\subsubsection{Reactive Inspector}
+
+In the paper ``Debugging for Reactive Programming''\cite{10.1145/2884781.2884815}, Salvaneschi et al.
+elaborate on the necessity of RP debugging tools which can interpret its specific dataflow abstractions.
+Based on their research, developed a reactive debugger named \emph{Reactive Inspector}.
+This plugin for the Eclipse IDE provides reactive debugging functionality for programs using the REScala RP runtime.
+
+The plugins core feature is the visualization of the dataflow graph.
+This representation of the dataflow provides a UI to add a new kind of breakpoint, the ``reactive breakpoint,'' to individual nodes in the graph.
+With them, the execution of the program can be interrupted e.g., when a node retains a specific value or a particular lifecycle event takes place.
+
+Along with the RP specific debugging features, Reactive Inspector contains  omniscient debugging concepts as well: The time travel functionality allows to record dataflows.
+These recordings can be replayed and inspected at a later time without an actual re-execution of the recorded program.
+
 \section{Threats to Validity}
@@
 \begin{acks}
-	We would like to
+
 \end{acks}

### 2020-08-21 paper.tex
Threats to validity

@@
 
-\textbf{TODO REWRITE HERE}
+\subsection{External Validity}
 
-We worked with 14 professional software engineers while writing this paper.
-Five interview partners and five war story reports helped us build the foundation for a later observational study with four participants.
+The data we collected from interviews, war story reports, and the observational study is based on a sample population with 14 individuals.
+Hence, the results we conveyed in this paper are not representative and are not transferable to the entire software engineering population.
 
-While we claim that nine interviewees and reporters have been an appropriate number of peers for collecting the initial data, we are aware that our four subjects for the observational study lead to a result lacking significance.
+\subsection{Internal Validity}
 
-More time given, we suggest conducting the study with a bigger population again to obtain a result with higher significance and providing more insight into how engineers debug RxJS-based code.
-Doing so would possibly require a statistical evaluation of these results.
+The observational study was executed in an uncontrolled environment.
+All subjects used their personal computers, running their own software development environments.
+We have no comparative data to measure how this design influenced the observed outcome, assuming that this setup diminishes the reproducibility of the experiment.
 
-Furthermore, this was the first time we designed an observational study of this kind.
-Though we carefully peer-reviewed the study design with our advisor, we cannot claim the design and its execution to be flawless.
-E.g., we would use a commonly known business domain for both problems presented to the subjects.
-This would allow them to understand the functionality of the application in front of them quicker and focus more on resolving the task given.
+We noticed that the subjects needed time to understand the intention of the applications they were provided with before they were able to start with the actual debugging process.
+Since the amount of required time was different from subject to subject, we suspect it influenced the result of the experiment.
+
+\subsection{Construct Validity}
+
+The time limit of 25 minutes per experiment block bears the potential to put the subjects under time pressure.
+This risk might explain why we could not observe any more time-consuming debugging techniques (e.g., installing additional utilities like rxjs-spy) during the study.
 
@@
 \begin{acks}
-
+	We want to thank the engineers who were part of this study for their time and insight throughout the different parts of our research.
+Further, we would like to thank Prof.
+Dr.
+Markus Stolze for his mentorship and many tireless review sessions.
 \end{acks}

### 2020-08-21 paper.tex
Rephrase Interpretation

@@
 \subsection{RxJS Debugging Utilities}
+\label{sec:rxjs-debugging-utilities}
 
@@
 RxViz, to inspect specific code parts in isolation in later iterations of the debugging process.
-Though, S2 noted that they would have usually started to decompose the problem into smaller pieces and observe their behavior in specific after the 25 minutes of the second block expired.
+Though, S4 noted that they would have usually started to decompose the problem into smaller pieces and observe their behavior in specific after the 25 minutes of the second block expired.
 
@@
 The later way of debugging was repeatedly commented as unsatisfying as these utilities cannot handle RP constructs and with this, cannot help to detect problems located within these at all.
-The former way, the introduction of manual log statements, was often described as ``the last resort'' when no other debugging technique helped before.
-This is because the analysis of log entries gets tedious as the number of those increases.
+The former way, the introduction of manual log statements, was both described as the prevalent way of debugging RxJS or as ``the last resort'' when no other debugging technique helped before.
 
 We heard further how engineers isolate specific Observables from bigger dataflows and how they inspect those in sandboxed environments and visualizers.
-This helps them understanding lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
+This helps them understanding the Observable lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
 
-We are able to postulate an answer to our first research question based on this collected information: \textbf{TODO Haben wir wirklich evidenz?
-Wenig diskutiert?} The biggest challenge for software engineers debugging RxJS-based programs is knowing \emph{when} to apply \emph{what} tool in order to get the \emph{best possible} answer on their currently tested hypothesis.
+More than 50\% of our peers throughout the interviews, war story reports and the experiments after-action survey stated to know about specific RxJS RP debugging tools.
+It was apparent that all subjects during the observational study refrained from using any of them though.
+It is our speculation that the subjects knowing about specific tools held themselves back from using them because they perceived the effort of setting them up (e.g., installing and configuring rxjs-spy) as too time-consuming.
+Not having the ``right'' tool available without significant additional effort is also what we interpret from the statement by S4: Though they would have started to extract parts of the dataflow and inspect it with other tools, they would have done so only after the 25 minutes of the block expired; Hence a more accessible way allowing such analyses would have influenced the subjects behavior.
+
+The best RP debugging tools are useless if either the hurdle to use them is too high or engineers do not understand in which particular part of the debugging process they can benefit from them.
+Hence, we are able to postulate an answer to our first research question: Biggest challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{best} way possible.
+
+\section{Future Work}
+\label{sec:future}
+
 We further claim that, backed by the observation where two subjects knew about rxjs-spy but refrained from using it nonetheless, it is not enough to \emph{know} about the existence of a specific tool for a particular debugging situation.
@@
 An example for such a behavior is the seamless integration of asynchronous call stacks in Google Chromes JavaScript debugger\cite{chromeasync}: Even though code was executed ``outside'' of the current call stack, the debugger shows these asynchronous stack frames as they were actually part of it.
-
-\section{Future Work}
-\label{sec:future}
 

### 2020-08-21 paper.tex
Future Work

@@
 
+\usepackage{tcolorbox}
 \usepackage{hyperref}
@@
 
-In response to this, we are going to present concepts on how to resolve previously identified challenges and answer the second research question:
+In response to this, we are going to present a concept on how to resolve previously identified challenges and answer the second research question:
 
@@
 We executed the experiment in two consecutive blocks of 25 minutes each.
-We provided a ZIP file\footnote{\url{https://github.com/swissmanu/mse-pa1-experiment/archive/v1.0.2.zip}} containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a test suite at the start of a session.
+We provided a ZIP file\footnote{\url{https://github.com/swissmanu/mse-pa1-experiment/archive/v1.0.2.zip}} containing the source code for two frontend web applications implemented using TypeScript and RxJS along with a Jest test suite at the start of a session.
 Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
+Where the first application required less complex intervention to resolve the contained bugs, the second application demanded for heavy modifications in the dataflow as it made heavy use of higher-order observables.
+The provided test suite allowed the subjects to understand the functional requirements of each application as well as to quickly verify their changes to be successful (or not).
+
 A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
@@
 
-A week before the experiment, we sent out the participant briefing document to all of our subjects.
+We sent out the participant briefing document to all of our subjects a week before the experiment.
 We outlined the course of action and provided them with an example ZIP file.
@@
 The best RP debugging tools are useless if either the hurdle to use them is too high or engineers do not understand in which particular part of the debugging process they can benefit from them.
-Hence, we are able to postulate an answer to our first research question: Biggest challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{best} way possible.
+Hence, we are able to postulate an answer to our first research question: The biggest challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
 
@@
 
-We further claim that, backed by the observation where two subjects knew about rxjs-spy but refrained from using it nonetheless, it is not enough to \emph{know} about the existence of a specific tool for a particular debugging situation.
-Such a tool should rather offer itself at the ``right'' moment so the engineer can benefit from its features.
-An example for such a behavior is the seamless integration of asynchronous call stacks in Google Chromes JavaScript debugger\cite{chromeasync}: Even though code was executed ``outside'' of the current call stack, the debugger shows these asynchronous stack frames as they were actually part of it.
+We see the biggest shortcoming of current RxJS-oriented debugging solutions like rxjs-spy, RxFiddle, or RxViz in the fact, that they are not integrated in established development environments (e.g., IDEs or internet browser developer tools).
+This leads to the practice of manually augmenting code itself rather then working with it in a less obtrusive, fully integrated way as we were able to proof in our observational study.
+Using specialized utilities is an extra effort an engineer has to invest every time they want to debug a dataflow: Either tagging an observable for rxjs-spy or extracting parts of it to an external environment, all of these practices require engineers to ``go the extra mile'' in order to inspect the runtime behavior of an RxJS-based application.
+The additional effort might be neglectable when treating a rather complex dataflow composition, but it holds back engineers from applying the tools to simple observables like in the first block of the experiment we conducted.
 
-\textbf{TODO REWRITE HERE}
+The observation that 50\% of our study subjects tried to debug an RP application with traditional, imperative-centric debugger utilities, as well as related statements from the interviews and war stories, strengthened our assumption regarding tool integration.
+Engineers expect the debugging tools they know and rely on to give correct insight on every program, no matter the paradigm (imperative or declarative) with which it got implemented.
 
-Having the main challenges for engineers identified when debugging RxJS-code, we can propose areas to improve on now and resolve RQ2 by doing so.
+This leads us to the answer on our second research question: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
 
-\subsection{Improve User Experience}
+\begin{tcolorbox}[colback=white,boxrule=1pt,sharp corners=all,left=1.8mm,boxsep=1mm]
+	\textbf{Example for Good Integration:} The debugger, part of the developer tools, in Google Chrome combines call stack frames of asynchronously executed functions\cite{chromeasync} seamlessly with those of synchronously executed code.
+This provides software engineers with a better understanding about which part of the program triggered the statement they currently inspect.
+\end{tcolorbox}
 
-Our overall goal is to support engineers during the process of understanding the inner workings of the RxJS dataflow graph.
-Existing tools and libraries provide some help, though often struggle by means of the UX: Trace logs with relevant information like life cycle events do improve the debugging process as we saw.
-Digging through hundreds of log entries containing information of multiple Observables gets tedious quickly, however.
-Filter and search functionalities help but cannot resolve all UX challenges.
+TODO
 
-A graphical representation of the dataflow graph is better suited to show dependencies within the model then textual trace logs.
-Previous research showed that the sheer amount of nodes in an elaborate graph is hard to visualize effectively\cite{10.1145/3180155.3180156} on one side, on the other also not easily implement efficiently as well.
-
-We intend to find a solution that combines the power of textual trace logs with the ease-of-use of graphical visualizations for RxJS dataflow graphs.
-We want this solution to handle small as well as large graphs.
-
-\subsection{Automatic Code Augmentation}
-
-RP code needs to be augmented with additional functionality so telemetry data can be collected.
-Previous solutions do this using code weaving\cite{10.1145/2884781.2884815} or similar techniques\cite{10.1145/3180155.3180156}.
-For engineers themselves though, this is often a manual process.
-We want to make the manual modification of production source code obsolete.
-rxfiddle proofed such mechanisms to work with RxJS 4 and 5, hence we intend to bring this feature to version 6, and possibly the upcoming major version 7 as well.
-
-\subsection{Integration}
-
-75\% of our observational study participants used the debugging tools integrated into their browser or IDE.
-Even though some of them stated to know additional external utilities in the after-action survey, none of them went that far and used them during the one-hour session.
-
-We understand the observed behavior as evidence for a hurdle preventing engineers from using specialized RP debugging utilities.
-In order to reach more software engineers with RP domain-specific debugging tools, we want to provide a potential solution as tightly integrated with the developer tools of the browser or an IDE as possible.
-
-\subsection{Other Areas of Improvement}
-
-There are many more areas which could be worth a further look in the future.
-E.g., the combination of controlling the program execution based on breakpoints within the dataflow graph or concepts of omniscient debugging are exciting fields as well.
-We will focus on the presented three topics for our future work, but keep additional ideas in a backlog for the case time allows additional iterations.
-
-\textbf{TODO Refer to Reactive Inspector}
-
-\subsubsection{Reactive Inspector}
-
-In the paper ``Debugging for Reactive Programming''\cite{10.1145/2884781.2884815}, Salvaneschi et al.
-elaborate on the necessity of RP debugging tools which can interpret its specific dataflow abstractions.
-Based on their research, developed a reactive debugger named \emph{Reactive Inspector}.
-This plugin for the Eclipse IDE provides reactive debugging functionality for programs using the REScala RP runtime.
-
-The plugins core feature is the visualization of the dataflow graph.
-This representation of the dataflow provides a UI to add a new kind of breakpoint, the ``reactive breakpoint,'' to individual nodes in the graph.
-With them, the execution of the program can be interrupted e.g., when a node retains a specific value or a particular lifecycle event takes place.
-
-Along with the RP specific debugging features, Reactive Inspector contains  omniscient debugging concepts as well: The time travel functionality allows to record dataflows.
-These recordings can be replayed and inspected at a later time without an actual re-execution of the recorded program.
+A future paper will elaborate on the selection of a RP debugging tool (e.g., a full reactive debugger or a visualizer using marble diagrams), how and where (i.e.
+which IDE or browser) it should be integrated in a future paper.
 
@@
 \label{sec:threats}
+
+This study is subject to the following threats and limitations:
 
@@
 \begin{acks}
-	We want to thank the engineers who were part of this study for their time and insight throughout the different parts of our research.
-Further, we would like to thank Prof.
-Dr.
-Markus Stolze for his mentorship and many tireless review sessions.
+	We want to thank the engineers who participated in our study for their time. Further, we would like to thank Prof. Dr. Markus Stolze for his mentorship and many tireless review sessions.
 \end{acks}

### 2020-08-21 paper.tex
Fix uppercase in Observable

@@
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
-The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables.
+The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.observable}}.}: An observable can be composed and linked (``piped'') with other observables.
 It can push (``emit'') values to its dependents.
-Multiple depending Observables form and describe a dataflow graph eventually.
-A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable.
-This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables.
-Values emitted from an Observable can be subscribed, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+Multiple depending observables form and describe a dataflow graph eventually.
+A multitude of operators allows the transformation of values and (higher-order\footnote{An observable emitted by another observable is considered a Higher-Order observable.
+This naming is related to the concept of higher-order functions in mathematics and computer science.}) observables.
+Values emitted from an observable can be subscribed, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
@@
 
-		of(0, 1, 2, 3, 4).pipe(  // Create Observable
+		of(0, 1, 2, 3, 4).pipe(  // Create observable
 			filter(i => i < 4),	  // Omit integers >= 4
@@
 	\end{minted}
-	\Description{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
-	\caption{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
+	\Description{Basic RxJS example creating an observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
+	\caption{Basic RxJS example creating an observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
 	\label{lst:rp-with-rxjs}
@@
 
-The RxJS community uses \emph{marble diagrams} as shown in Figure~\ref{fig:marble-diagram} to document \cite{marblediagrams} the runtime behavior of an Observable visually.
-Unit test libraries\cite{marbletesting} use this abstraction to encode the behavior of mocked Observables or to describe assertions.
+The RxJS community uses \emph{marble diagrams} as shown in Figure~\ref{fig:marble-diagram} to document \cite{marblediagrams} the runtime behavior of an observable visually.
+Unit test libraries\cite{marbletesting} use this abstraction to encode the behavior of mocked observables or to describe assertions.
 
@@
 	\tikzfig{marble-diagram}
-	\Description{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
+	\Description{A Marble Diagram visualizing the observable in Listing~\ref{lst:rp-with-rxjs}.
 From left to right, each marble represents an emitted value of 0, 2, 4, and 6.
-The vertical line at the last marble indicates that the Observable completed after emitting value \mintinline{TypeScript}{6}.}
-	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}.
+The vertical line at the last marble indicates that the observable completed after emitting value \mintinline{TypeScript}{6}.}
+	\caption{A Marble Diagram visualizing the observable in Listing~\ref{lst:rp-with-rxjs}.
 From left to right, each marble represents an emitted value.
-The vertical line at the last marble indicates that the Observable completed after emitting \mintinline{TypeScript}{6}.}
+The vertical line at the last marble indicates that the observable completed after emitting \mintinline{TypeScript}{6}.}
 	\label{fig:marble-diagram}
@@
 subscribe, unsubscribe, complete and error).
-Ideally, tagged Observables are created during development when the dataflows are composed for the first time.
-Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an Observable emits a value.
-It cannot modify or influence the emitted value nor the Observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual dataflow.
+Ideally, tagged observables are created during development when the dataflows are composed for the first time.
+Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an observable emits a value.
+It cannot modify or influence the emitted value nor the observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual dataflow.
 
@@
 Additional features are available through the libraries console interface.
-E.g., a tagged Observable can be paused so values get collected rather than being emitted immediately.
+E.g., a tagged observable can be paused so values get collected rather than being emitted immediately.
 The engineer can then emitted these values one after another manually or resume all of them at once.
@@
 They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
-The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an Observable is created and how it interacts with other Observables, whereas a marble diagram visualizes the values emitted over time for every Observable.
+The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an observable is created and how it interacts with other observables, whereas a marble diagram visualizes the values emitted over time for every observable.
 
@@
 \emph{RxViz} is a visualizer utility available online\footnote{\url{https://rxviz.com/}}.
-It is an ``animated playground for Rx Observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
+It is an ``animated playground for Rx observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
 Engineers implement or copy-paste dataflows in an editor window using JavaScript.
@@
 Our first three interview partners I1 to I3 declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
-By being the creator of a visualizer for RxJS Observables, I4 was a proficient RxJS user.
+By being the creator of a visualizer for RxJS observables, I4 was a proficient RxJS user.
 Our fifth interview partner I5 was a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
@@
 It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
-E.g., when using ngrx for state management, ``The framework hides Observables from its main API surface carefully so you do not have to interact with them directly'' (I2).
-As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying Observables though.
+E.g., when using ngrx for state management, ``The framework hides observables from its main API surface carefully so you do not have to interact with them directly'' (I2).
+As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying observables though.
 
@@
 I1 to I3 mentioned the Redux DevTools\footnote{\url{https://chrome.google.com/webstore/detail/redux-devtools}} as particular helpful when debugging Angular/ngrx applications nonetheless.
-Further, I1 noted marble diagrams as valuable in order to understand how an RxJS Observable works, whether during development or debugging.
+Further, I1 noted marble diagrams as valuable in order to understand how an RxJS observable works, whether during development or debugging.
 
@@
 Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, R2 as well as R4 and R5 commented on the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our interview partners.
-R5 described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators.
+R5 described a specific scenario where they suspected a problem within a complex observable composition: Having multiple asynchronous, remote data sources, they used observables to model the dependencies between them and implemented computations on their results as operators.
 On top, each data source could re-emit updated versions of previously requested information at any time.
@@
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs.
-The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future.
+The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the observable stream in case they might be needed again in the future.
 
-In a related war story, R2 discloses similar invasive practices using an external tool when they implemented logic to request and cache batches of a remote resource: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed them to run, debug, and iterate on selected pieces of a larger Observable stream.
+In a related war story, R2 discloses similar invasive practices using an external tool when they implemented logic to request and cache batches of a remote resource: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed them to run, debug, and iterate on selected pieces of a larger observable stream.
 Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
@@
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
-But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when Observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the Observable.''
+But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the observable.''
 
@@
 The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep.
-We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests.
+We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an observables behavior and are useful to implement tests.
 Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the pre-provided functionality needs to be extended.
@@
 Almost all of the reporters referred to the practice of modifying their source code manually and adding log statements where they assumed a problem to overcome the feature gaps in traditional debuggers.
-Listing~\ref{lst:rxjs-debugging} exemplifies two challenges when debugging a stream of Observables with imperative-focused debugging tools.
+Listing~\ref{lst:rxjs-debugging} exemplifies two challenges when debugging a stream of observables with imperative-focused debugging tools.
 
@@
 	\end{minted}
-	\caption{An Observable emitting a sequence of increasing integers every second.
+	\caption{An observable emitting a sequence of increasing integers every second.
 Traditional breakpoints are possible inside the arrow function on Line~2.
@@
 We learned that simple trace augmentation, like on Line~4 in Listing~\ref{lst:rxjs-debugging}, logs emitted values only.
-This might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying Observable completely.
-The importance of such information was confirmed by the reporter describing their usage of rxjs-spy: The knowledge about the Observable lifecycle, like when it gets subscribed, unsubscribed or completed, helped them to solve problems multiple times.
-When dealing with higher-order Observables, they value this information even as indispensable.
+This might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying observable completely.
+The importance of such information was confirmed by the reporter describing their usage of rxjs-spy: The knowledge about the observable lifecycle, like when it gets subscribed, unsubscribed or completed, helped them to solve problems multiple times.
+When dealing with higher-order observables, they value this information even as indispensable.
 
-Finally, we understood that if a problem is hard to replicate within the actual application, the engineers use external sandbox development environments to isolate specific parts of an Observable composition.
+Finally, we understood that if a problem is hard to replicate within the actual application, the engineers use external sandbox development environments to isolate specific parts of an observable composition.
 These allow them to iterate on it faster than it would be possible otherwise.
@@
 
-We heard further how engineers isolate specific Observables from bigger dataflows and how they inspect those in sandboxed environments and visualizers.
-This helps them understanding the Observable lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
+We heard further how engineers isolate specific observables from bigger dataflows and how they inspect those in sandboxed environments and visualizers.
+This helps them understanding the observable lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
 

### 2020-08-21 paper.tex
Future Work

@@
 
-TODO
-
-A future paper will elaborate on the selection of a RP debugging tool (e.g., a full reactive debugger or a visualizer using marble diagrams), how and where (i.e.
-which IDE or browser) it should be integrated in a future paper.
+The answer on which RP debugging tool exactly (e.g., a full reactive debugger or a visualizer using marble diagrams) we are going to integrate, how such an integration will look like in detail, as well as the answer on RQ3 will be part of our future work on the topic of ``Debugging RxJS-based Applications''.
 

### 2020-08-21 paper.tex
Conclusion

--- 2020-08-21 paper.tex
@@
 The (graphical) user interface (\emph{UI} or \emph{GUI}) of an application handles two constant flows of data: External user input (e.g.
-mouse, touch or keyboard interaction) is interpreted and forwarded to the system.
-Once the system processed an interaction and updated its internal state accordingly, it notifies the UI about these changes which are relayed to the user.
+mouse, touch, or keyboard interaction) is interpreted and forwarded to the system.
+Once the system processed an interaction and updated its internal state accordingly, it notifies the UI about these changes, which are relayed to the user.
 
@@
 The system sends update notifactions to the UI eventually.}
-	\caption{Basic dataflows in a UI Application.}
+	\caption{Basic data-flows in a UI Application.}
 	\label{fig:ui-data-flows}
@@
 
-To implement the dataflows as shown in Figure~\ref{fig:ui-data-flows} to drive a UI, the Observer design pattern\cite{gamma1995design} is often used and variations of the pattern are omnipresent today.
+To implement the data-flows as shown in Figure~\ref{fig:ui-data-flows} to drive a UI, the Observer design pattern\cite{gamma1995design} is often used and variations of the pattern are omnipresent today\cite{alabor:2019:reactiveappllications}.
 
-The Observer design pattern has its roots in the Object Oriented Programming paradigm (\emph{OOP}), hence relies on imperative code constructs to form a dataflow.
-Reactive Programming (\emph{RP}) is another approach to realize such flows: It inherits the declarative way of implementing functionality from Functional Programming (\emph{FP}), i.e.
-dataflows are rather described than implemented step by step\cite{10.1145/2501654.2501666}.
+The Observer design pattern has its roots in the Object Oriented Programming paradigm (\emph{OOP}), hence relies on imperative code constructs to form a data-flow.
+Reactive Programming (\emph{RP}) is another approach to realize such flows: It inherits the declarative way of implementing functionality from Functional Programming (\emph{FP}), i.e., data-flows are described rather than implemented step by step\cite{10.1145/2501654.2501666}.
 RP functionality is usually available in the form of a library providing necessary abstractions, either for imperative or declarative programming languages.
@@
 Modern IDEs and internet browsers ship with their own set of debugging tools.
-These traditional debuggers are specialized in working with imperative, controlflow oriented program code.
+These traditional debuggers are specialized in working with imperative, control-flow oriented program code.
 E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a dataflow oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
+Assuming a data-flow oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
 
-These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow-specific inquiry.
+These exemplary questions demonstrate the limits of a traditional control-flow oriented debugger, which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a data-flow-specific inquiry.
 Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities.
@@
 We will conclude this introductory section with the clarification of important terms and a view on known RP debugging utilities.
-Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports.
-We present our observational study intended to validate former insights in Section~\ref{sec:study} which allow us to answer RQ1.
+Section~\ref{sec:interviews} gives an overview on the insights from the conducted interviews and the collected war story reports.
+We present our observational study intended to validate former results in Section~\ref{sec:study} which allow us to answer RQ1.
 Before our final conclusion, we will answer RQ2 in Section~\ref{sec:future} ``Future Work'' and review the threats to validity regarding our study in Section~\ref{sec:threats}.
@@
 
-Either way, both usually provide a (i) domain specific language (\emph{DSL}) to describe dataflow graphs, how they depend on each other and how data flowing through should be transformed.
+Either way, both usually provide a (i) domain specific language (\emph{DSL}) to describe data-flow graphs, how they depend on each other and how data flowing through should be transformed.
 At program execution, a (ii) runtime environment evaluates these descriptions and creates a representation of the specified graphs.
@@
 It can push (``emit'') values to its dependents.
-Multiple depending observables form and describe a dataflow graph eventually.
+Multiple depending observables form and describe a data-flow graph eventually.
 A multitude of operators allows the transformation of values and (higher-order\footnote{An observable emitted by another observable is considered a Higher-Order observable.
@@
 Layman et al.
-\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} use the debugging process model, an iterative, hypothesis refinement process, to formalize the general activity of debugging a computer program in their paper.
+\cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} use the debugging process model, an iterative hypothesis refinement process, to formalize the general activity of debugging a computer program in their paper.
 
@@
 
-Software engineers have many tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program.
+Software engineers have many tools and utilities at hand, which help them to interact with and gain insight on the behavior of a defective program.
 Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.
@@
 
-Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
+Traditional (i) \emph{imperative-focused debuggers} provide the functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame.
 Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''
 
-RP provides its own set of challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph.
+RP provides its own set of challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than, e.g., the predecessor transformation according to the data-flow graph.
 Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph.
 A (ii) \emph{reactive debugger} can interpret the underlying graph model of a RP runtime system.
-It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
+It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the data-flow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.
 
-Traditional, as well as reactive debuggers work with the \emph{current} state of a programs execution only.
+Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only.
 They lack information about what happened before or what is going to happen in the future.
@@
 subscribe, unsubscribe, complete and error).
-Ideally, tagged observables are created during development when the dataflows are composed for the first time.
+Ideally, tagged observables are created during development when the data-flows are composed for the first time.
 Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an observable emits a value.
-It cannot modify or influence the emitted value nor the observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual dataflow.
+It cannot modify or influence the emitted value nor the observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual data-flow.
 
@@
 
-The dataflow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log} eventually.
+The data-flow configuration in Listing~\ref{lst:rxjs-spy-tag} will produce a trace log as shown in Listing~\ref{lst:rxjs-spy-log} eventually.
 
@@
 
-Additional features are available through the libraries console interface.
-E.g., a tagged observable can be paused so values get collected rather than being emitted immediately.
+Additional features are available through the library's console interface.
+E.g., a tagged observable can be paused, so values get collected rather than being emitted immediately.
 The engineer can then emitted these values one after another manually or resume all of them at once.
@@
 
-\emph{rxfiddle} as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
+\emph{rxfiddle}, as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
 They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
@@
 The reference implementation supports event processing for the (outdated) RxJS major versions 4 and 5 only and is available as an online application\footnote{\url{https://rxfiddle.net/}}.
-A proof-of-concept implementation which can be executed on a local computer is available through the projects Git repository\footnote{\url{https://github.com/hermanbanken/RxFiddle}}.
+A proof-of-concept implementation working on a local computer is available through the projects Git repository\footnote{\url{https://github.com/hermanbanken/RxFiddle}}.
 
@@
 It is an ``animated playground for Rx observables''\cite{rxviz} and allows the visualization of RxJS observables using marble diagrams.
-Engineers implement or copy-paste dataflows in an editor window using JavaScript.
-This code is then used to generate a diagram over a configurable time interval.
+Engineers implement or copy-paste data-flows in an editor window using JavaScript.
+A diagram is generated based on this code over a configurable time interval.
 The diagrams are rendered immediately and are available as downloadable SVG files.
@@
 
-We organized informal interviews which allowed us to gain insight in how software engineers work with RP in their daily jobs.
+We organized informal interviews, which allowed us to gain insight in how software engineers work with RP in their daily jobs.
 We talked to five engineers (following identified using the codes \emph{I1} through \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
@@
 
-Our first three interview partners I1 to I3 declared to currently work or more recently have worked with RxJS in conjunction with Angular and/or ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
+Our first three interview partners I1 to I3 declared to currently work or more recently have worked with RxJS in conjunction with Angular and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
 By being the creator of a visualizer for RxJS observables, I4 was a proficient RxJS user.
-Our fifth interview partner I5 was a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model dataflows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
+Our fifth interview partner I5 was a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
-All of the interviewees referred to RP with an overall positive experience.
+All of the interviewees referred to RP with a positive experience.
 We heard statements similar to ``RP is a good way for composing multiple data sources'' (I1) and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' (I5) repeatedly throughout all sessions.
-Hence a significant strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' stated I3, giving lectures in frontend web application development.
+Hence a significant strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL.
+The learning curve can be steep for a novice engineer, though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' stated I3, giving lectures in frontend web application development.
 
 It was interesting to hear that, especially in the area of developing web applications using Angular, our partners seemed not to have to work with pure RxJS code often.
-E.g., when using ngrx for state management, ``The framework hides observables from its main API surface carefully so you do not have to interact with them directly'' (I2).
-As soon as our interviewees had to extend built-in functionalities with their own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying observables though.
+E.g., when using ngrx for state management, ``The framework hides observables from its main API surface carefully, so you do not have to interact with them directly'' (I2).
+As soon as our interviewees had to extend built-in functionalities with own features, e.g., a new effect\footnote{\url{https://ngrx.io/guide/effects}}, I1 and I2 valued the possibility to interact with underlying observables though.
 
@@
 after we changed them in the system,'' they told us.
-The browsers debugger tools and its breakpoints did not help much since the operators were executed several times.
+The browser's debugger tools and its breakpoints did not help much since the operators were executed several times.
 Other parts of the stream were impossible to grasp, even with conditional breakpoints.
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs.
-The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the observable stream in case they might be needed again in the future.
+The log statements added were removed in the aftermath, though the rxjs-spy-related code was left in the observable stream in case they might be needed again in the future.
 
-In a related war story, R2 discloses similar invasive practices using an external tool when they implemented logic to request and cache batches of a remote resource: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed them to run, debug, and iterate on selected pieces of a larger observable stream.
+In a related war story, R2 discloses similar invasive practices using an external tool when they implemented logic to request and cache batches of a remote resource: ``It took quite some time to get it right and one of the most invaluable tools proved to be Stackblitz\footnote{Stackblitz is a full JavaScript development environment available online \url{https://stackblitz.com/}} which gave us the ability to quickly create smaller working examples and iterate on them.'' This sandboxed setting allowed them to run, debug, and iterate on selected pieces of a larger observable stream.
 Even though the final result had to be integrated back into the actual application, the extra effort was worth the result the engineer concluded in their report.
 
-A final story describes yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, R5 used Rx Visualizer, an online visualization utilities to generate marble diagrams from real code.
+A final story describes yet another way of utilizing an external tool: Rather than using a dedicated sandbox to develop pieces of a more complex system, R5 used Rx Visualizer, an online visualization utility to generate marble diagrams from real code.
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
-But once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the observable.''
+Once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the observable.''
 
@@
 The sentiment in both the interviews and the reported war stories in regards to RP with RxJS was positive.
-The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep.
+The software engineers value how they can describe data-flows using a DSL, even though the learning curve was perceived as steep.
 We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an observables behavior and are useful to implement tests.
@@
 
-Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness, as we heard in the interviews.
-When the engineers needed to interact with dataflow graphs at runtime, e.g.
+Most participants considered a statically typed language like TypeScript, a fundamental necessity allowing them to implement data-flow graphs with minimal, formal correctness, as we heard in the interviews.
+When the engineers needed to interact with data-flow graphs at runtime, e.g.
 to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided with through traditional debugging tools.
@@
 Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3.
-One would need to place the breakpoint within the operators' internal implementation instead which can be cumbersome in case the operator is used in a different stream as well.
+One would need to place the breakpoint within the operator's internal implementation instead, which can be cumbersome in case the operator is used in a different stream as well.
 Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
-That is why a traditional debuggers step controls cannot operate on the dataflow graph; It just can not interpret this level of abstraction.
+That is why a traditional debuggers step controls cannot operate on the data-flow graph; It just can not interpret this level of abstraction.
 
@@
 We learned that simple trace augmentation, like on Line~4 in Listing~\ref{lst:rxjs-debugging}, logs emitted values only.
-This might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying observable completely.
+Basic trace logs might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying observable completely.
 The importance of such information was confirmed by the reporter describing their usage of rxjs-spy: The knowledge about the observable lifecycle, like when it gets subscribed, unsubscribed or completed, helped them to solve problems multiple times.
@@
 
-After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools which can handle RP concepts provided through RxJS.
+After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools that can handle RP concepts provided through RxJS.
 Even though traditional debuggers might help to some extent, they do not provide all the information an engineer is interested in.
@@
 Each of these applications was rigged with two to three bugs, which we asked the subjects to identify and fix using whatever debugging techniques they prefer and commonly use.
-Where the first application required less complex intervention to resolve the contained bugs, the second application demanded for heavy modifications in the dataflow as it made heavy use of higher-order observables.
+Where the first application required less complicated intervention to resolve the contained bugs, the second application demanded substantial modifications in the data-flow as it made heavy use of higher-order observables.
 The provided test suite allowed the subjects to understand the functional requirements of each application as well as to quickly verify their changes to be successful (or not).
 
-A block was considered as complete once either the test suite signaled all bugs as resolved or the 25 minutes were exceeded.
+A block was considered as complete once the test suite signaled all bugs as resolved, or the 25 minutes expired.
 We asked our subjects to act like in a pair programming situation where they ``think out loud'' their thought process.
@@
 
-After the subjects got themselves accustomed with the application provided and understood its purpose, all of them used the test suite to gather context about what features do not work as expected initially.
+After the subjects got themselves accustomed to the application provided and understood its purpose, all of them used the test suite to gather context about what features do not work as expected initially.
 Further, all of them tried to recreate the failing behavior in the UI manually.
@@
 We could further observe a ``trail-and-error'' approach in later iterations of the debugging process.
-The subjects started to introduce modifications to the system which they immediately tested against their latest hypothesis.
+The subjects started to introduce modifications to the system, which they immediately tested against their latest hypothesis.
 Table~\ref{tab:subject-techniques} provides an overview on the complete collected data regarding used techniques and tools.
@@
 
-Only S4 was able to solve the first problem given as shown in Table~\ref{tab:subject-results}.
+Only S4 was able to solve the first problem given, as shown in Table~\ref{tab:subject-results}.
 None of the subjects was able to successfully identify and fix the bugs hidden in the second problem within time.
@@
 We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs.
-Only 50\% of them used traditional debugging utilities in order to inspect the programs state at runtime.
+Only 50\% of them used traditional debugging utilities in order to inspect the program's state at runtime.
 All subjects used the new information gained to refine their hypothesis about underlying problems before they started a new iteration in the debugging process.
@@
 Beside the habit of adding trace logs manually, we saw them evidently trying to answer their debugging hypotheses using traditional, imperative-focused debugger utilities.
-The later way of debugging was repeatedly commented as unsatisfying as these utilities cannot handle RP constructs and with this, cannot help to detect problems located within these at all.
+The later way of debugging was repeatedly commented as unsatisfying as these utilities cannot handle RP constructs, and with this, cannot help to detect problems located within these at all.
 The former way, the introduction of manual log statements, was both described as the prevalent way of debugging RxJS or as ``the last resort'' when no other debugging technique helped before.
 
-We heard further how engineers isolate specific observables from bigger dataflows and how they inspect those in sandboxed environments and visualizers.
+We heard further how engineers isolate specific observables from bigger data-flows and how they inspect those in sandboxed environments and visualizers.
 This helps them understanding the observable lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
@@
 More than 50\% of our peers throughout the interviews, war story reports and the experiments after-action survey stated to know about specific RxJS RP debugging tools.
-It was apparent that all subjects during the observational study refrained from using any of them though.
+It was apparent that all subjects during the observational study refrained from using any of them, though.
 It is our speculation that the subjects knowing about specific tools held themselves back from using them because they perceived the effort of setting them up (e.g., installing and configuring rxjs-spy) as too time-consuming.
-Not having the ``right'' tool available without significant additional effort is also what we interpret from the statement by S4: Though they would have started to extract parts of the dataflow and inspect it with other tools, they would have done so only after the 25 minutes of the block expired; Hence a more accessible way allowing such analyses would have influenced the subjects behavior.
+Not having the ``right'' tool available without significant additional effort is also what we interpret from the statement by S4: Though they would have started to extract parts of the data-flow and inspect it with other tools, they would have done so only after the 25 minutes of the block expired; Hence a more accessible way allowing such analyses would have influenced the behavior of the subject.
 
-The best RP debugging tools are useless if either the hurdle to use them is too high or engineers do not understand in which particular part of the debugging process they can benefit from them.
-Hence, we are able to postulate an answer to our first research question: The biggest challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
+The best RP debugging tools are useless if either the hurdle to use them is too high, or engineers do not understand in which particular part of the debugging process they can benefit from them.
+Hence, we are able to postulate an answer to our first research question: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
 
@@
 
-We see the biggest shortcoming of current RxJS-oriented debugging solutions like rxjs-spy, RxFiddle, or RxViz in the fact, that they are not integrated in established development environments (e.g., IDEs or internet browser developer tools).
+We see the biggest shortcoming of current RxJS-oriented debugging solutions like rxjs-spy, RxFiddle, or RxViz in fact that they are not integrated in established development environments (e.g., IDEs or internet browser developer tools).
 This leads to the practice of manually augmenting code itself rather then working with it in a less obtrusive, fully integrated way as we were able to proof in our observational study.
-Using specialized utilities is an extra effort an engineer has to invest every time they want to debug a dataflow: Either tagging an observable for rxjs-spy or extracting parts of it to an external environment, all of these practices require engineers to ``go the extra mile'' in order to inspect the runtime behavior of an RxJS-based application.
-The additional effort might be neglectable when treating a rather complex dataflow composition, but it holds back engineers from applying the tools to simple observables like in the first block of the experiment we conducted.
+Using specialized utilities is an extra effort an engineer has to invest every time they want to debug a data-flow: Either tagging an observable for rxjs-spy or extracting parts of it to an external environment, all of these practices require engineers to ``go the extra mile'' in order to inspect the runtime behavior of an RxJS-based application.
+The additional effort might be neglectable when treating a rather complex data-flow composition, but it holds back engineers from applying the tools to simple observables like in the first block of the experiment we conducted.
 
@@
 
-This leads us to the answer on our second research question: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
+This leads us to the answer to our second research question: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
 
@@
 
-TODO
-
-A future paper will elaborate on the selection of a RP debugging tool (e.g., a full reactive debugger or a visualizer using marble diagrams), how and where (i.e.
-which IDE or browser) it should be integrated in a future paper.
+The answer on which RP debugging tool exactly (e.g., a full reactive debugger or a visualizer using marble diagrams) we are going to integrate, how such integration will look like in detail, how it will support engineers in a particular step of the debugging process, as well as the answer on RQ3 will be part of our future work on the topic of ``Debugging RxJS-based Applications''.
 
@@
 
-\textbf{TODO REWRITE HERE}
+In this paper, we have explored how software engineers debug data-flow-oriented programs implemented using RxJS.
+We presented an observational study to validate a hypothesis based on the outcome of 10 individual interviews and hands-on experience reports from software engineering professionals.
+More than 50\% of the engineers we worked with during our research told us that they know of the existence of specific debugging tools for RP with RxJS.
+Nonetheless, the experiment conducted with four participants allowed us to prove that engineers augment source code manually with trace logs instead of using such specialized utilities.
 
-We have talked to multiple people in order to understand how software engineers apply RP using RxJS for building applications.
-We have looked into how they debug their applications and could identify several struggles they have to overcome in this process.
-We could prove that the engineers inevitably tend to modify their source code once they do not understand how the RP runtime propagates values through its internal dataflow graph by observing professionals in an observational study.
-
-On the way, we elaborated on previous, academic as well as non-academic efforts in the field of reactive debugging.
-We presented a selection of tools which leverage on their knowledge about the internal workings of RP runtime environments.
-
-With \emph{UX}, \emph{Automatic Code Augmentation} and \emph{Integration} we finally proposed three concepts where we intend to improve on a software engineers debugging experience, hence answering our second research question RQ2.
-We plan to work further on these topics within the scope of our master's studies research and, eventually, answer our last research question RQ3.
+We identified the fact that RxJS specific debugging tools are not tightly integrated with existing, traditional debuggers in IDEs and the developer tools of internet browsers as the main reason why software engineers do not use them more often.
+In order to lower the effort necessary to use specialized RP debugging tools for engineers, we declared the integration of such as the theme for our own future work on the topic.
 

### 2020-08-21 abstract.tex
Introduction Stuff
Abstract

@@
-An application implemented using RxJS is unintuitive and time-consuming to debug.
-We will explore the practices software engineers apply during debugging and take a close look at the challenges they face in this workshop paper.
-On the way to propose solution concepts, we will validate previously identified struggles in an observational study and elaborate on previous research efforts in the field of reactive debugging.
+RxJS is a popular library to implement data-flow-oriented applications with JavaScript using reactive programming principles.
+The library's domain specific language allows the sophisticated description of data-flow compositions.
+Though this way of programming bears new challenges for traditional debuggers: Their focus on imperative programming limits their applicability to problems originated in the declarative programming paradigm.
+The goals of this paper are: (i) to understand how software engineers debug RxJS-based applications, what tools do they use, what techniques they apply; (ii) to understand what are the most prevalent challenges they face while doing so; and (iii) to provide a course of action to resolve these challenges in a future iteration on this topic.
+We learned about the debugging habits of ten professionals using interviews, and hands-on war story reports.
+Based on this data, we designed and executed an observational study with four subjects to verify that engineers manually augment source code with trace logs instead of using specialized debugging utilities.
+In the end, we identified the lack of fully integrated RxJS-specific debugging solutions in existing development environments as the most significant reason why engineers do not make use of such tools.
+We decided to elaborate on how to resolve this situation in detail as the topic for our future work.

### 2020-08-21 paper.tex
Abstract

@@
 Modern IDEs and internet browsers ship with their own set of debugging tools.
-These traditional debuggers are specialized in working with imperative, control-flow oriented program code.
+These traditional debuggers are specialized in working with imperative, control-flow-oriented program code.
 E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a data-flow oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
+Assuming a data-flow-oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
 
@@
 E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs.
-In the end, we were able to conduct interviews with 10 engineers and collect war stories from another five.
+In the end, we were able to conduct interviews with ten engineers and collect war stories from another five.
 
@@
 In this paper, we have explored how software engineers debug data-flow-oriented programs implemented using RxJS.
-We presented an observational study to validate a hypothesis based on the outcome of 10 individual interviews and hands-on experience reports from software engineering professionals.
+We presented an observational study to validate a hypothesis based on the outcome of ten individual interviews and hands-on experience reports from software engineering professionals.
 More than 50\% of the engineers we worked with during our research told us that they know of the existence of specific debugging tools for RP with RxJS.

### 2020-08-21 abstract.tex
Fix Findings v0.12

@@
 RxJS is a popular library to implement data-flow-oriented applications with JavaScript using reactive programming principles.
-The library's domain specific language allows the sophisticated description of data-flow compositions.
-Though this way of programming bears new challenges for traditional debuggers: Their focus on imperative programming limits their applicability to problems originated in the declarative programming paradigm.
-The goals of this paper are: (i) to understand how software engineers debug RxJS-based applications, what tools do they use, what techniques they apply; (ii) to understand what are the most prevalent challenges they face while doing so; and (iii) to provide a course of action to resolve these challenges in a future iteration on this topic.
+This way of programming bears new challenges for traditional debuggers: Their focus on imperative programming limits their applicability to problems originated in the declarative programming paradigm.
+The goals of this paper are: (i) to understand how software engineers debug RxJS-based applications, what tools do they use, what techniques they apply; (ii) to understand what are the most prevalent challenges they face while doing so; and (iii) to provide a course of action to resolve these challenges in a future iteration on the topic.
 We learned about the debugging habits of ten professionals using interviews, and hands-on war story reports.
-Based on this data, we designed and executed an observational study with four subjects to verify that engineers manually augment source code with trace logs instead of using specialized debugging utilities.
+Based on this data, we designed and executed an observational study with four subjects to verify that engineers predominantly augment source code with manuel trace logs instead of using specialized debugging utilities.
 In the end, we identified the lack of fully integrated RxJS-specific debugging solutions in existing development environments as the most significant reason why engineers do not make use of such tools.

### 2020-08-21 paper.tex
Fix Findings v0.12

@@
 
-The Observer design pattern has its roots in the Object Oriented Programming paradigm (\emph{OOP}), hence relies on imperative code constructs to form a data-flow.
+The Observer design pattern has its roots in the Object Oriented Programming paradigm (\emph{OOP}), hence relies on imperative code constructs to handle a data-flow.
 Reactive Programming (\emph{RP}) is another approach to realize such flows: It inherits the declarative way of implementing functionality from Functional Programming (\emph{FP}), i.e., data-flows are described rather than implemented step by step\cite{10.1145/2501654.2501666}.
-RP functionality is usually available in the form of a library providing necessary abstractions, either for imperative or declarative programming languages.
+RP functionality is usually available in the form of a library providing necessary abstractions, for imperative as well as declarative programming languages.
 
@@
 Modern IDEs and internet browsers ship with their own set of debugging tools.
-These traditional debuggers are specialized in working with imperative, control-flow-oriented program code.
-E.g., when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question.
-Assuming a data-flow-oriented program implemented using RP, the question, what transformation function processed a piece of information right before another cannot be answered using a call stack anymore: The stack leads to the internals of the RP runtime environment rather than to the transformations logical predecessor.
+These debuggers are specialized in working with imperative, control-flow-oriented program code.
+The following example helps us to illustrate the implications of this: Assuming an engineer is inspecting a piece of code and wants to know which part of the program was executed right before.
+For a program implemented using the imperative paradigm, the call stack gives a clear answer to this question.
+Hence the stack frames represent each point in the program execution.
+In a data-flow-oriented program implemented using RP, the stack trace for a transformation function in the flow will not point to its logical predecessor.
+Instead, the stack frames lead to the internals of the RP runtime environment.
 
-These exemplary questions demonstrate the limits of a traditional control-flow oriented debugger, which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a data-flow-specific inquiry.
-Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities.
-For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when applying RP.
+This example demonstrates the limits of a traditional control-flow oriented debugger, which cannot interpret RP abstractions.
+As a result, these debuggers are not able to give the correct answer to a data-flow-specific inquiry.
+There have been numerous efforts to provide engineers with improved debugging utilities for RP \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156}.
+However, none of these have seen broad adoption by practitioners yet.
+To gain a better understanding of the underlying root causes, we conducted interviews with several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when using RP.
 Based on this collected evidence, we will validate their statements in an observational study using RxJS and search for an answer to our first research question:
@@
 \begin{itemize}
-	\item \emph{RQ1: What challenges do software engineers face when debugging RxJS-based code?}
+	\item \emph{RQ1: What challenges do software engineers face when debugging RxJS-based applications?}
 \end{itemize}
@@
 \begin{itemize}
-	\item \emph{RQ2: How can the experience of software engineers during debugging of RxJS-based code be improved?}
+	\item \emph{RQ2: How can the experience of software engineers during the debugging process of RxJS-based applications be improved?}
 \end{itemize}
 
-The implementation and validation of these proposals lead to our third and last research question, whose answer will be part of a future iteration on the topic:
+The implementation and validation of these proposals lead to our third and last research question, which will be investigated in our future research:
 
@@
 We will conclude this introductory section with the clarification of important terms and a view on known RP debugging utilities.
-Section~\ref{sec:interviews} gives an overview on the insights from the conducted interviews and the collected war story reports.
-We present our observational study intended to validate former results in Section~\ref{sec:study} which allow us to answer RQ1.
+Section~\ref{sec:interviews} gives an overview of the insights from the conducted interviews and the collected war story reports.
+We present our observational study intended to validate results from the interviews and reports in Section~\ref{sec:study}, which allow us to answer RQ1.
 Before our final conclusion, we will answer RQ2 in Section~\ref{sec:future} ``Future Work'' and review the threats to validity regarding our study in Section~\ref{sec:threats}.
@@
 
-RP is a declarative programming paradigm originated in FP.
-Where engineers use imperative programming languages to specify every step \emph{how} a program has to do something, declarative languages allow to describe \emph{what} the program should achieve ultimately.
+RP is a declarative programming paradigm that has its origins in FP.
+While engineers use imperative programming languages to specify every step \emph{how} a program has to do something, declarative languages allow to describe \emph{what} the program should achieve ultimately.
 A runtime system then figures out a way to satisfy that description and executes it.
@@
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
-The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.observable}}.}: An observable can be composed and linked (``piped'') with other observables.
-It can push (``emit'') values to its dependents.
-Multiple depending observables form and describe a data-flow graph eventually.
-A multitude of operators allows the transformation of values and (higher-order\footnote{An observable emitted by another observable is considered a Higher-Order observable.
+The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.observable}}.}: An observable can be composed with other observables to form a data-flow graph.
+Once an observable gets subscribed, it might push (``emit'') an arbitrary number of values to the subscriber until it either completes, fails, or gets unsubscribed again.
+There is a multitude of operator functions available which allow the transformation of values and composition with other (higher-order\footnote{An observable emitted by another observable is considered a Higher-Order observable.
 This naming is related to the concept of higher-order functions in mathematics and computer science.}) observables.
-Values emitted from an observable can be subscribed, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+The mechanism of subscribing to an observable is closely related to the Observer design patterns \mintinline{Java}{attach} method.
 
@@
 
-On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge on its own to find people who understand themselves as users of RP and related technologies.
+On the way of finding our interview partners and war stories reporters, we noticed it to be a challenge to find people who understand themselves as users of RP and related technologies.
 E.g., even though Angular makes heavy use of RxJS, we will see that many engineers do not directly interact with its abstractions when building ``basic'' UIs.
-In the end, we were able to conduct interviews with ten engineers and collect war stories from another five.
+In the end, we were able to conduct interviews with five engineers and collect reports on hands-on experiences from another five.
 
@@
 
-We organized informal interviews, which allowed us to gain insight in how software engineers work with RP in their daily jobs.
+We organized informal interviews, which allowed us to gain insight into how software engineers work with RP in their daily jobs.
 We talked to five engineers (following identified using the codes \emph{I1} through \emph{I5}) and asked them about the technologies they use, what their personal experience with RP was, what they most liked and most disliked about it.
-All interviews were conducted remotely using video chat.
+We used video chat to conduct all interviews remotely.
+The interview with I4 was done in English, all others in Swiss-German.
+Therefore, quotes by I1, I2, I3, and I5 are translated statements.
 
-Our first three interview partners I1 to I3 declared to currently work or more recently have worked with RxJS in conjunction with Angular and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
-By being the creator of a visualizer for RxJS observables, I4 was a proficient RxJS user.
+Our first three interview partners I1 to I3 stated to work currently or more recently have worked with RxJS in conjunction with Angular and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend web applications.
+I4 was a proficient RxJS user.
 Our fifth interview partner I5 was a backend engineer who used akka-streams\footnote{\url{https://akka.io/}} in Scala to model data-flows for a WebSocket-based\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/API/WebSockets_API}}, reactive API layer serving a web frontend application.
 
-All of the interviewees referred to RP with a positive experience.
-We heard statements similar to ``RP is a good way for composing multiple data sources'' (I1) and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' (I5) repeatedly throughout all sessions.
+All interview partners pointed out that they like RP because it provides them with ``[\dots] a good way for composing multiple data sources'' (I1) and, combined with ``[\dots] a statically typed language, RP guarantees some kind of basic formal correctness of a program'' (I5).
 Hence a significant strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL.
-The learning curve can be steep for a novice engineer, though: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' stated I3, giving lectures in frontend web application development.
+However, they also pointed to current challenges: The learning curve can be steep for a novice engineer: ``Being challenged with new abstractions [of Angular and ngrx] already, I experienced RxJS concepts and operators to be hard to convey'' stated I3, giving lectures in frontend web application development.
 
@@
 We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request.
-After reaching out, we were able to collect five responses: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another set of reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.
+After reaching out, we were able to collect five responses in English: One by an RxJS core team member (R1), two from Angular Google Developer Experts (R2/3) and another set of reports by two software engineers (R4/5) building web and mobile applications using React and RxJS, which includes the author of this paper.
 
@@
 Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
-With its current major version, RxJS 6 improved heavily on the  \mintinline{TypeScript}{TestScheduler}.
+With its current major version, RxJS 6 improved profoundly on the  \mintinline{TypeScript}{TestScheduler}.
 The runtime environment itself can be augmented with the scheduler now, which results in cleaner code.
@@
 The browser's debugger tools and its breakpoints did not help much since the operators were executed several times.
-Other parts of the stream were impossible to grasp, even with conditional breakpoints.
+Other parts of the stream were impossible to get a handle upon, even with conditional breakpoints.
 ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information.'' After a time-consuming log analysis, they finally were able to resolve all bugs.
@@
 Like in the report before, it was necessary to extract parts from the codebase of the actual application.
-Once done, the visualizations helped a lot in order to understand when values were emitted, when subscriptions changed and when observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the observable.''
+Once done, the visualizations helped to understand when values got emitted, when subscriptions changed and when observables completed: ``Marble diagrams were a huge help in order to understand detailed runtime and life cycle behaviors of the observable.''
 
@@
 
-The sentiment in both the interviews and the reported war stories in regards to RP with RxJS was positive.
-The software engineers value how they can describe data-flows using a DSL, even though the learning curve was perceived as steep.
+Software engineers value how they can describe data-flows using a RP DSL, even though the learning curve was perceived as steep.
 We heard some interesting reports on how RxJS is applied in a daily development environment: Marble diagrams help to understand  an observables behavior and are useful to implement tests.
@@
 to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided with through traditional debugging tools.
-The lack of knowledge about RP abstractions like transformation operators or life cycles was reported as irritating.
-Almost all of the reporters referred to the practice of modifying their source code manually and adding log statements where they assumed a problem to overcome the feature gaps in traditional debuggers.
+Almost all of the reporters referred to the practice of modifying their source code manually and adding trace log statements where they assumed a problem to overcome the feature gaps in traditional debuggers.
 Listing~\ref{lst:rxjs-debugging} exemplifies two challenges when debugging a stream of observables with imperative-focused debugging tools.
@@
 Once the breakpoint on Line~2 interrupts the execution of the program, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.
-That is why a traditional debuggers step controls cannot operate on the data-flow graph; It just can not interpret this level of abstraction.
+That is why a traditional debugger's step controls cannot operate on the data-flow graph; It just can not interpret this level of abstraction.
 
@@
 We learned that simple trace augmentation, like on Line~4 in Listing~\ref{lst:rxjs-debugging}, logs emitted values only.
-Basic trace logs might be helpful when debugging simple graph compositions, though it lacks lifecycle information of the underlying observable completely.
-The importance of such information was confirmed by the reporter describing their usage of rxjs-spy: The knowledge about the observable lifecycle, like when it gets subscribed, unsubscribed or completed, helped them to solve problems multiple times.
+Such trace logs might be helpful when debugging simple graph compositions.
+Though, they lack lifecycle information of the underlying observables completely.
+The importance of such information was emphasized by R5 describing their usage of rxjs-spy: To know, when an observable gets subscribed and unsubscribed, when it completes or fails, helped them to solve complex problems multiple times.
 When dealing with higher-order observables, they value this information even as indispensable.
@@
 
-After the evaluation of all reports and interviews, we speculate that software engineers truly lack debugging tools that can handle RP concepts provided through RxJS.
-Even though traditional debuggers might help to some extent, they do not provide all the information an engineer is interested in.
-The last resort in this situation is often the manual augmentation, modification, and extraction of source code as we saw repeatedly.
-After a look at previous efforts in the area of reactive debugging, we are going to validate our assumption in Section~\ref{sec:study}.
+After the evaluation of all reports and interviews, we speculate that software engineers truly lack, but also seldomly use debugging tools that can handle RP concepts provided by RxJS.
+Even though traditional debuggers might help to some extent, they do not provide all the information an engineer requires in a particular situation.
+Instead, they turn to manual trace log augmentation and extraction of source code as we saw repeatedly.
 
@@
 Almost all participants from our interviews and war story reports showed a tendency to manually modify source code with trace logs during the hypothesis instrumentation phase when debugging RxJS code.
-It is further our own experience that this practice is often not productive since the evaluation and interpretation of trace information tends to be cumbersome and very time-consuming.
+This practice is often not perceived as efficient since the evaluation and interpretation of trace information tends to be cumbersome and very time-consuming.
 Also, removing log statements after a successful debugging process might leave new bugs in production code if not done carefully.
@@
 The subjects for our study were required to have experience in developing applications with RxJS.
-We did not distinguish between frontend or backend engineers, so in the end, we were able to recruit four subjects willing to participate in our experiment.
-We were interested to see how the subjects apply debugging techniques they would use in an everyday situation at their jobs.
+We recruited four subjects willing to participate in our experiment.
+We were interested in seeing how the subjects apply debugging techniques they would use in everyday situations in their jobs.
 Hence we decided to conduct the experiment in a somewhat uncontrolled environment where the subjects used their own devices with their development environments of personal preference.
@@
 
-We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended quantitative after-action survey.
+We planned to have a one-hour session for the actual experiment with each subject, followed-up by an unattended after-action survey.
 We executed the experiment in two consecutive blocks of 25 minutes each.
@@
 We identified the fact that RxJS specific debugging tools are not tightly integrated with existing, traditional debuggers in IDEs and the developer tools of internet browsers as the main reason why software engineers do not use them more often.
-In order to lower the effort necessary to use specialized RP debugging tools for engineers, we declared the integration of such as the theme for our own future work on the topic.
+In order to lower the effort necessary to use specialized RP debugging tools for engineers, we declared the integration of such as the matter for our own future research.
 

### 2020-08-26 paper.tex
Review Stef

@@
 After a week in production, though tested thoroughly, the first of many bugs got reported: ``Displayed numbers kept changing where we did not expected them to.
-In other places, they were not rerendered where they were supposed to after, e.g.
+In other places, they were not rerendered where they were supposed to, e.g.
 after we changed them in the system,'' they told us.
@@
 When the engineers needed to interact with data-flow graphs at runtime, e.g.
-to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided with through traditional debugging tools.
+to debug the behavior of a specific part of a stream, we noticed throughout most reports that they were not 100\% satisfied with the feature set they were provided by traditional debugging tools.
 Almost all of the reporters referred to the practice of modifying their source code manually and adding trace log statements where they assumed a problem to overcome the feature gaps in traditional debuggers.
@@
 The best RP debugging tools are useless if either the hurdle to use them is too high, or engineers do not understand in which particular part of the debugging process they can benefit from them.
-Hence, we are able to postulate an answer to our first research question: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
+Hence, we are able to postulate an answer to our first research question RQ1: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
 
@@
 
-This leads us to the answer to our second research question: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
+This leads us to the answer to our second research question RQ2: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
 

### 2020-09-29 paper.tex
Change Author Institution
Add rxjs-playground

@@
 \affiliation{
-	\institution{HSR University of Applied Sciences}
+	\institution{Eastern Switzerland University of Applied Sciences}
 	\city{Rapperswil}
@@
 }
-\email{malabor@hsr.ch}
+\email{manuel.alabor@ost.ch}
 
@@
 \emph{rxjs-spy}\cite{rxjsspy} is a logging library specialized on RxJS observables.
-Once an observable is tagged with an arbitrary string identifier, a monitor can generate trace logs whenever a value is emitted as well as when individual lifecycle events occur (i.e.
+Once an observable is tagged with an arbitrary string identifier, a monitor can generate trace logs whenever a value is emitted as well as when individual life cycle events occur (i.e.
 subscribe, unsubscribe, complete and error).
@@
 \emph{rxfiddle}, as proposed by Banken et al.\cite{10.1145/3180155.3180156} is the first reference implementation of their RP debugger architecture for the ReactiveX API specification.
-They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or lifecycle events) and forwards them to the second component.
+They describe a software design consisting of two independent components: The (i) \emph{host instrumentation} augments a ReactiveX API implementation to emit events at runtime (e.g., emitting a value or life cycle events) and forwards them to the second component.
 The (ii) \emph{visualizer} interprets the events and displays them along two dimensions: The StoryFlow graph\cite{YWu2013a} shows when an observable is created and how it interacts with other observables, whereas a marble diagram visualizes the values emitted over time for every observable.
@@
 The diagrams are rendered immediately and are available as downloadable SVG files.
+
+\subsubsection{rxjs-playground}
+
+Building on the basic concept of marble diagrams, \emph{rxjs-playground}\footnote{\url{https://hediet.github.io/rxjs-playground}} is a sophisticated sandbox to simulate and visualize RxJS observables interactively in the browser.
+Users define editable and computed observables, represented as vertical marble diagrams: Values emitted by an editable observable can be created and modified either by interacting with its marble diagram directly or using a simple JSON syntax.
+The behavior of a computed observable is controlled by implementing functionality with TypeScript in the provided editor.
+
+rxjs-playground renders all observables\' values and life cycles in real-time allowing quick iteration on a specific piece of code.
 
@@
 Such trace logs might be helpful when debugging simple graph compositions.
-Though, they lack lifecycle information of the underlying observables completely.
+Though, they lack life cycle information of the underlying observables completely.
 The importance of such information was emphasized by R5 describing their usage of rxjs-spy: To know, when an observable gets subscribed and unsubscribed, when it completes or fails, helped them to solve complex problems multiple times.
@@
 We heard further how engineers isolate specific observables from bigger data-flows and how they inspect those in sandboxed environments and visualizers.
-This helps them understanding the observable lifecycle and value emitting behaviors better and iterate faster in order to resolve problems.
+This helps them understanding the observable life cycle and value emitting behaviors better and iterate faster in order to resolve problems.
 

### 2020-09-29 paper.tex
Remove Percentages

@@
 
-RP is a declarative programming paradigm that has its origins in FP.
+RP is a declarative programming paradigm that is strongly influenced by FP.
 While engineers use imperative programming languages to specify every step \emph{how} a program has to do something, declarative languages allow to describe \emph{what} the program should achieve ultimately.
@@
 
-The survey responses available in Table~\ref{tab:subject-survey} showed that 75\% (S2, S3 and S4) of the subject population had two or more years of experience with RxJS.
+The survey responses available in Table~\ref{tab:subject-survey} showed that S2, S3 and S4 had two or more years of experience with RxJS.
 Where all of them use RxJS to develop frontend applications, S4 declared having used  RxJS for backend development as well.
@@
 We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs.
-Only 50\% of them used traditional debugging utilities in order to inspect the program's state at runtime.
+Two of them used traditional debugging utilities in order to inspect the program's state at runtime.
 All subjects used the new information gained to refine their hypothesis about underlying problems before they started a new iteration in the debugging process.
@@
 
-More than 50\% of our peers throughout the interviews, war story reports and the experiments after-action survey stated to know about specific RxJS RP debugging tools.
+More than 50\% of our 14 peers throughout the interviews, war story reports and the experiments after-action survey stated to know about specific RxJS RP debugging tools.
 It was apparent that all subjects during the observational study refrained from using any of them, though.
@@
 
-The observation that 50\% of our study subjects tried to debug an RP application with traditional, imperative-centric debugger utilities, as well as related statements from the interviews and war stories, strengthened our assumption regarding tool integration.
+The observation that two out of four of our study subjects tried to debug an RP application with traditional, imperative-centric debugger utilities, as well as related statements from the interviews and war stories, strengthened our assumption regarding tool integration.
 Engineers expect the debugging tools they know and rely on to give correct insight on every program, no matter the paradigm (imperative or declarative) with which it got implemented.
@@
 We presented an observational study to validate a hypothesis based on the outcome of ten individual interviews and hands-on experience reports from software engineering professionals.
-More than 50\% of the engineers we worked with during our research told us that they know of the existence of specific debugging tools for RP with RxJS.
+More than 50\% of the 14 engineers we worked with during our research told us that they know of the existence of specific debugging tools for RP with RxJS.
 Nonetheless, the experiment conducted with four participants allowed us to prove that engineers augment source code manually with trace logs instead of using such specialized utilities.

### 2020-09-29 paper.tex
Better Integrate Main References

@@
 Also, removing log statements after a successful debugging process might leave new bugs in production code if not done carefully.
-We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
+We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code, which supports earlier results by Banken et al.
+\cite{10.1145/3180155.3180156} further.
 
-That is why we saw demand in validating this statement about manual code modification for debugging reasons with an observational study. Our study sought to validate the following hypothesis:
+That is why we saw demand in validating this statement and previous findings about manual code modification for debugging reasons with an observational study.
+Our study sought to validate the following hypothesis:
 
@@
 
-We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs.
-Two of them used traditional debugging utilities in order to inspect the program's state at runtime.
+We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs, which confirmed previous results by Banken et al.
+\cite{10.1145/3180155.3180156}.
+Two of the subjects used traditional debugging utilities in order to inspect the program's state at runtime in addition.
 All subjects used the new information gained to refine their hypothesis about underlying problems before they started a new iteration in the debugging process.
@@
 
-Even though two subjects stated in the after-action survey to regularly use rxjs-spy for debugging RxJS programs, neither S1 nor S3 made use of this library during the experiment part of the study.
+Even though S1 and S3 stated in the after-action survey to regularly use rxjs-spy for debugging RxJS programs, neither of them made use of this library during the experiment part of the study.
 
@@
 
-The best RP debugging tools are useless if either the hurdle to use them is too high, or engineers do not understand in which particular part of the debugging process they can benefit from them.
-Hence, we are able to postulate an answer to our first research question RQ1: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
+The best RP debugging tools are useless if either the hurdle to use them is too high, or engineers do not understand which particular part of the debugging process they can benefit from them.
+Salvaneschi et al.
+\cite{10.1145/2884781.2884815} provided in their previous study on \emph{Reactive Inspector} for \emph{REScala} evidence on the effectiveness of a fully integrated RP debugging solution, which supports developers in their daily work using the Eclipse IDE.
+Hence, we can postulate an answer to our first research question RQ1: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
 

### 2020-09-29 paper.tex
Editing

@@
 Also, removing log statements after a successful debugging process might leave new bugs in production code if not done carefully.
-We identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code, which supports earlier results by Banken et al.
-\cite{10.1145/3180155.3180156} further.
+Like Banken et al.
+\cite{10.1145/3180155.3180156} before, we identified this technique as one of the primary debugging practices when software engineers work with RxJS-based code.
 
@@
 
-We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs, which confirmed previous results by Banken et al.
-\cite{10.1145/3180155.3180156}.
+We were able to observe how all subjects predominantly used manual source code augmentation by adding trace logs.
 Two of the subjects used traditional debugging utilities in order to inspect the program's state at runtime in addition.
-All subjects used the new information gained to refine their hypothesis about underlying problems before they started a new iteration in the debugging process.
+All subjects used the new information gained to refine their hypothesis about underlying problems before starting a new iteration in the debugging process.
 We could not observe the extraction to and reintegration from an external tool.
 All subjects exhibited the debugging behavior described in our hypothesis.
+Further, we could verify previous results by Banken et al.
+\cite{10.1145/3180155.3180156} successfully as well.
 
@@
 Salvaneschi et al.
-\cite{10.1145/2884781.2884815} provided in their previous study on \emph{Reactive Inspector} for \emph{REScala} evidence on the effectiveness of a fully integrated RP debugging solution, which supports developers in their daily work using the Eclipse IDE.
+\cite{10.1145/2884781.2884815} provided in their previous study on the \emph{Reactive Inspector} for \emph{REScala} evidence on the effectiveness of a fully integrated RP debugging solution, which supports developers in their daily work using the Eclipse IDE.
 Hence, we can postulate an answer to our first research question RQ1: The most significant challenge software engineers face when debugging RxJS-based programs is to know \emph{when} they should apply \emph{what} tool to resolve their current problem in the \emph{most efficient} way.
@@
 Using specialized utilities is an extra effort an engineer has to invest every time they want to debug a data-flow: Either tagging an observable for rxjs-spy or extracting parts of it to an external environment, all of these practices require engineers to ``go the extra mile'' in order to inspect the runtime behavior of an RxJS-based application.
-The additional effort might be neglectable when treating a rather complex data-flow composition, but it holds back engineers from applying the tools to simple observables like in the first block of the experiment we conducted.
+The additional effort might be neglectable when treating a rather complex data-flow composition.
+However, it holds back engineers from applying the tools to simple observables like in the first block of the experiment we conducted.
 

### 2020-10-02 paper.tex
Fix Rendering
Add Co-Author

@@
 \email{manuel.alabor@ost.ch}
+
+\author{Prof. Dr. Markus Stolze}
+\affiliation{
+	\institution{Eastern Switzerland University of Applied Sciences}
+	\city{Rapperswil}
+	\country{Switzerland}
+}
+\email{markus.stolze@ost.ch}
 
@@
 
-rxjs-playground renders all observables\' values and life cycles in real-time allowing quick iteration on a specific piece of code.
+rxjs-playground renders the values and life cycle events for all observables in real-time, allowing quick iterations on a specific piece of code.
 
@@
 \begin{acks}
-	We want to thank the engineers who participated in our study for their time. Further, we would like to thank Prof. Dr. Markus Stolze for his mentorship and many tireless review sessions.
+	We want to thank the engineers who participated in our study for their time.
 \end{acks}

### 2020-10-15 abstract.tex
Review Stef
Add ACM References

@@
 We learned about the debugging habits of ten professionals using interviews, and hands-on war story reports.
-Based on this data, we designed and executed an observational study with four subjects to verify that engineers predominantly augment source code with manuel trace logs instead of using specialized debugging utilities.
+Based on this data, we designed and executed an observational study with four subjects to verify that engineers predominantly augment source code with manual trace logs instead of using specialized debugging utilities.
 In the end, we identified the lack of fully integrated RxJS-specific debugging solutions in existing development environments as the most significant reason why engineers do not make use of such tools.
-We decided to elaborate on how to resolve this situation in detail as the topic for our future work.
+We decided to elaborate on how to resolve this situation in our future work.

### 2020-10-15 paper.tex
Add ACM References

@@
-\documentclass[sigplan,screen,review]{acmart}
+\documentclass[sigplan,screen]{acmart}
 % Class options:
@@
 
+%%% The following is specific to REBLS '20 and the paper
+%%% 'Debugging RxJS-Based Applications'
+%%% by Manuel Alabor.
+%%%
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3427763.3428313}
+\acmYear{2020}
+\copyrightyear{2020}
+\acmSubmissionID{splashws20reblsmain-p3-p}
+\acmISBN{978-1-4503-8188-8/20/11}
+\acmConference[REBLS '20]{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems}{November 16, 2020}{Virtual, USA}
+\acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems (REBLS '20), November 16, 2020, Virtual, USA}
+
 % Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
-\acmConference[SPLASH '20]{SPLASH 2020: 7th Workshop on Reactive and Event-based Languages \& Systems}{November 15--20, 2020}{Chicago}
 \title{Debugging of RxJS-based Applications}

### 2020-10-15 paper.tex
Latest ACM

@@
-\documentclass[sigplan,screen]{acmart}
+\documentclass[sigplan,screen]{./acmart}
 % Class options:
@@
 %%%
+\copyrightyear{2020}
+\acmYear{2020}
 \setcopyright{acmlicensed}
+\acmConference[REBLS '20]{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems}{November 16, 2020}{Virtual, USA}
+\acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems (REBLS '20), November 16, 2020, Virtual, USA}
 \acmPrice{15.00}
 \acmDOI{10.1145/3427763.3428313}
-\acmYear{2020}
-\copyrightyear{2020}
-\acmSubmissionID{splashws20reblsmain-p3-p}
 \acmISBN{978-1-4503-8188-8/20/11}
-\acmConference[REBLS '20]{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems}{November 16, 2020}{Virtual, USA}
-\acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems (REBLS '20), November 16, 2020, Virtual, USA}
 
 % Top Matter: https://mirror.kumi.systems/ctan/macros/latex/contrib/acmart/acmart.pdf
-\title{Debugging of RxJS-based Applications}
+\title{Debugging of RxJS-Based Applications}
 \author{Manuel Alabor}
@@
 
-\author{Prof. Dr. Markus Stolze}
+\author{Markus Stolze}
 \affiliation{
@@
 			rxjs 6.5.2/internal/observable/interval.js:23
-			rxjs 6.5.2/internal/scheduler/AsyncAction.js:71
 			// ...

### 2020-10-15 paper.tex
Get fully acmart compliant
Dont use any fancy latex packages ;)

@@
-\documentclass[sigplan,screen]{./acmart}
+\documentclass[sigplan,screen]{acmart}
 % Class options:
@@
 
-\usepackage{tcolorbox}
-\usepackage{hyperref}
-\usepackage{minted}
-\usepackage{booktabs}
+\usepackage{fancybox}
 \usepackage{./tikzit}
 
-\setminted{autogobble=true, tabsize=2, linenos=true, frame=single, breaklines=true}
+\usepackage{listings}
+\lstdefinelanguage{TypeScript}{
+	keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, const, let, if, in, while, do, else, case, break},
+	keywordstyle=\color{blue}\bfseries,
+	commentstyle=\color{olive}\ttfamily,
+	stringstyle=\color{brown}\ttfamily,
+	ndkeywords={class, export, boolean, throw, implements, import, this, from, type, interface},
+	sensitive=false,
+	comment=[l]{//},
+	morecomment=[s]{/*}{*/},
+	morestring=[b]',
+	morestring=[b]",
+}
+\lstset{
+	language=TypeScript,
+	tabsize=2,
+	numbers=left,
+	numberstyle=\tiny,
+	numbersep=1.2em,
+	basicstyle=\small,
+	captionpos=b,
+	framexrightmargin=-0.5em,
+	xleftmargin=1.5em,
+	framexleftmargin=0em,
+	frame=single,
+	aboveskip=\bigskipamount,
+	belowskip=\bigskipamount,
+	flexiblecolumns=true,
+	floatplacement=h
+}
 
 %%% The following is specific to REBLS '20 and the paper
-%%% 'Debugging RxJS-Based Applications'
-%%% by Manuel Alabor.
+%%% 'Debugging of RxJS-Based Applications'
+%%% by Manuel Alabor and Markus Stolze.
 %%%
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3427763.3428313}
+\acmYear{2020}
 \copyrightyear{2020}
-\acmYear{2020}
-\setcopyright{acmlicensed}
+\acmSubmissionID{splashws20reblsmain-p3-p}
+\acmISBN{978-1-4503-8188-8/20/11}
 \acmConference[REBLS '20]{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems}{November 16, 2020}{Virtual, USA}
 \acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Reactive and Event-Based Languages and Systems (REBLS '20), November 16, 2020, Virtual, USA}
-\acmPrice{15.00}
-\acmDOI{10.1145/3427763.3428313}
-\acmISBN{978-1-4503-8188-8/20/11}
 
@@
 \begin{abstract}
-	\input{./abstract.tex}
+	RxJS is a popular library to implement data-flow-oriented applications with JavaScript using reactive programming principles.
+This way of programming bears new challenges for traditional debuggers: Their focus on imperative programming limits their applicability to problems originated in the declarative programming paradigm.
+The goals of this paper are: (i) to understand how software engineers debug RxJS-based applications, what tools do they use, what techniques they apply; (ii) to understand what are the most prevalent challenges they face while doing so; and (iii) to provide a course of action to resolve these challenges in a future iteration on the topic.
+We learned about the debugging habits of ten professionals using interviews, and hands-on war story reports.
+Based on this data, we designed and executed an observational study with four subjects to verify that engineers predominantly augment source code with manual trace logs instead of using specialized debugging utilities.
+In the end, we identified the lack of fully integrated RxJS-specific debugging solutions in existing development environments as the most significant reason why engineers do not make use of such tools.
+We decided to elaborate on how to resolve this situation in our future work.
 \end{abstract}
@@
 ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}.
-The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.observable}}.}: An observable can be composed with other observables to form a data-flow graph.
+The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\texttt{java.util.Observable}}.}: An observable can be composed with other observables to form a data-flow graph.
 Once an observable gets subscribed, it might push (``emit'') an arbitrary number of values to the subscriber until it either completes, fails, or gets unsubscribed again.
@@
 This naming is related to the concept of higher-order functions in mathematics and computer science.}) observables.
-The mechanism of subscribing to an observable is closely related to the Observer design patterns \mintinline{Java}{attach} method.
+The mechanism of subscribing to an observable is closely related to the Observer design patterns \texttt{attach} method.
 
@@
 
-\begin{listing}
-	\begin{minted}{TypeScript}
+\begin{lstlisting}[float,label=lst:rp-with-rxjs,caption={Basic RxJS example creating an observable emitting four integers.
+Each integer is processed by two operators and finally written to the console.}]
 import { of } from 'rxjs';
@@
 ).subscribe(console.log) // Logs: 0, 2, 4, 6
-	\end{minted}
-	\Description{Basic RxJS example creating an observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
-	\caption{Basic RxJS example creating an observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
-	\label{lst:rp-with-rxjs}
-\end{listing}
+\end{lstlisting}
 
@@
 From left to right, each marble represents an emitted value of 0, 2, 4, and 6.
-The vertical line at the last marble indicates that the observable completed after emitting value \mintinline{TypeScript}{6}.}
+The vertical line at the last marble indicates that the observable completed after emitting value \texttt{6}.}
 	\caption{A Marble Diagram visualizing the observable in Listing~\ref{lst:rp-with-rxjs}.
 From left to right, each marble represents an emitted value.
-The vertical line at the last marble indicates that the observable completed after emitting \mintinline{TypeScript}{6}.}
+The vertical line at the last marble indicates that the observable completed after emitting \texttt{6}.}
 	\label{fig:marble-diagram}
@@
 Ideally, tagged observables are created during development when the data-flows are composed for the first time.
-Like \mintinline{TypeScript}{tap}\footnote{RxJS' \mintinline{TypeScript}{tap} operator is used to execute a side effect whenever an observable emits a value.
-It cannot modify or influence the emitted value nor the observable in any way.}, the \mintinline{TypeScript}{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual data-flow.
+Like \texttt{tap}\footnote{RxJS' \texttt{tap} operator is used to execute a side effect whenever an observable emits a value.
+It cannot modify or influence the emitted value nor the observable in any way.}, the \texttt{tag} operator in Listing~\ref{lst:rxjs-spy-tag} is completely transparent to the actual data-flow.
 
-\begin{listing}
-	\begin{minted}{TypeScript}
+\begin{lstlisting}[float,label=lst:rxjs-spy-tag,caption={Application of rxjs-spy using its \texttt{tag} operator on Line~9 and 11.}]
 import { create } from 'rxjs-spy';
@@
 ).subscribe();
-	\end{minted}
-	\caption{Application of rxjs-spy using its \mintinline{TypeScript}{tag} operator on Line~9 and 11.}
-	\label{lst:rxjs-spy-tag}
-\end{listing}
+\end{lstlisting}
 
@@
 
-\begin{listing}
-	\begin{minted}{Typescript}
+\begin{lstlisting}[float,label=lst:rxjs-spy-log,caption={rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}]
 Tag = multiply; notification = subscribe
@@
 Tag = multiply; notification = unsubscribe
-	\end{minted}
-	\caption{rxjs-spy execution trace log generated by default monitor in Listing~\ref{lst:rxjs-spy-tag} on Line~4.}
-	\label{lst:rxjs-spy-log}
-\end{listing}
+\end{lstlisting}
 
@@
 Allowing to pass the scheduler explicitly as parameter forced them to introduce code, which was only necessary for testing reasons they stated further.
-With its current major version, RxJS 6 improved profoundly on the  \mintinline{TypeScript}{TestScheduler}.
+With its current major version, RxJS 6 improved profoundly on the  \texttt{TestScheduler}.
 The runtime environment itself can be augmented with the scheduler now, which results in cleaner code.
@@
 
-\begin{listing}
-	\begin{minted}{TypeScript}
+\begin{lstlisting}[float,label=lst:rxjs-debugging,caption={An observable emitting a sequence of increasing integers every second.
+Traditional breakpoints are possible inside the arrow function on Line~2.
+Though a breakpoint can be added on Line~3, it will never be hit during the actual execution of the \texttt{take} operator.
+Line~4 shows a manually introduced trace log statement using the \texttt{tap} side effect operator.}]
 interval(1000).pipe(
@@
 ).subscribe(showValue); // Emits: 0, 2, 4, 6
-	\end{minted}
-	\caption{An observable emitting a sequence of increasing integers every second.
-Traditional breakpoints are possible inside the arrow function on Line~2.
-Though a breakpoint can be added on Line~3, it will never be hit during the actual execution of the \mintinline{TypeScript}{take} operator.
-Line~4 shows a manually introduced trace log statement using the \mintinline{TypeScript}{tap} side effect operator.}
-	\label{lst:rxjs-debugging}
-\end{listing}
+\end{lstlisting}
 
-Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3.
+Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \texttt{take} on Line~3.
 One would need to place the breakpoint within the operator's internal implementation instead, which can be cumbersome in case the operator is used in a different stream as well.
@@
 
-\begin{listing}
-	\begin{minted}{TypeScript}
+\begin{lstlisting}[float,label=lst:rxjs-call-stack,caption={A call stack showing the internal RxJS execution stack for a breakpoint in the arrow function on Line~2 in Listing~\ref{lst:rxjs-debugging}.}]
 <anonymous> RxJS
@@
    // ...
-	\end{minted}
-	\caption{A call stack showing the internal RxJS execution stack for a breakpoint in the arrow function on Line~2 in Listing~\ref{lst:rxjs-debugging}.}
-	\label{lst:rxjs-call-stack}
-\end{listing}
+\end{lstlisting}
 
@@
 
-While all subjects added manual trace log statements to existing arrow functions or by adding \mintinline{TypeScript}{tap} operators in the instrument hypothesis phase, none of them used additional libraries like rxjs-spy for doing so.
+While all subjects added manual trace log statements to existing arrow functions or by adding \texttt{tap} operators in the instrument hypothesis phase, none of them used additional libraries like rxjs-spy for doing so.
 S2 and S4 used the traditional debugging tools provided by their browser or IDE to add breakpoints.
@@
 
+\balance
+
 \section{Future Work}
@@
 
-This leads us to the answer to our second research question RQ2: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools.
-
-\begin{tcolorbox}[colback=white,boxrule=1pt,sharp corners=all,left=1.8mm,boxsep=1mm]
-	\textbf{Example for Good Integration:} The debugger, part of the developer tools, in Google Chrome combines call stack frames of asynchronously executed functions\cite{chromeasync} seamlessly with those of synchronously executed code.
-This provides software engineers with a better understanding about which part of the program triggered the statement they currently inspect.
-\end{tcolorbox}
+This leads us to the answer to our second research question RQ2: We want to improve the experience of debugging RxJS-based applications by providing RP specific debugging utilities where software engineers expect them the most: Fully integrated with the traditional debugger they know from their IDE or browser developer tools\footnote{A positive example of such a seamless integration is the debugger of the Google Chrome developer tools: It combines call stack frames of asynchronously executed functions\cite{chromeasync} seamlessly with those of synchronously executed code.
+This provides software engineers with a better understanding about which part of the program triggered the statement they currently inspect.}.
 
@@
 
+\balance
+
 \bibliographystyle{ACM-Reference-Format}

### 2020-11-02 slides.tex
Prepare Slides

\documentclass[aspectratio=169]{beamer}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetheme[showmaxslides, darkmode, nofooter]{pureminimalistic}

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
% \pgfpagesuselayout{2 on 1}[a4paper,border shrink=10mm]

\usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2,style=numeric, sorting=none, url=false, eprint=false]{biblatex}
\addbibresource{demo_bib.bib}

% this makes it possible to add backup slides, without counting them
\usepackage{appendixnumberbeamer}
\renewcommand{\appendixname}{\texorpdfstring{\translate{appendix}}{appendix}}
\renewcommand{\logoheader}{}

% if loaded after begin{document} a warning will appear: "pdfauthor already used"
\title[Debugging RxJS-based Applications]{Getting Rid of the \texttt{console.log}:\\Debugging RxJS-based Applications}
\author{Manuel Alabor}
\institute{Eastern Switzerland University of Applied Sciences}
\date{\today}

\begin{document}
  \begin{frame}{Default Options}
    Lorem ipsum dolor sit amet, consetetur sadipscing elitr,
    sed diam nonumy eirmod tempor invidunt ut labore et dolore
    magna aliquyam erat, sed diam voluptua.
    At vero eos et accusam et justo duo dolores et ea rebum.
    Stet clita kasd gubergren, no sea takimata sanctus est
    Lorem ipsum dolor sit amet.
  \end{frame}
\end{document}

% \documentclass[aspectratio=169]{beamer}
% % should also look nice for the classic aspectratio
% % of course, than the text has to be refitted
% % \documentclass{beamer}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{tikz}
% \usetheme[darkmode, showmaxslides]{pureminimalistic}
% % \renewcommand{\pageword}{}
% % \renewcommand{\logoheader}{\vspace{1.5em}}
% \usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2,%
%             style=numeric, sorting=none, url=false, eprint=false]{biblatex}
% \addbibresource{bibliography.bib}
% % this makes it possible to add backup slides, without counting them
% \usepackage{appendixnumberbeamer}
% \renewcommand{\appendixname}{\texorpdfstring{\translate{appendix}}{appendix}}

% \title[short title]{This is the normal length of a research paper:
% always longer than you would expect}
% \author{Kai Norman Clasen}
% \institute{Insitute name}
% \date{\today}

% % \documentclass{beamer}
% % \documentclass[handout]{beamer}
% % \usepackage{pgfpages}
% % \setbeameroption{show notes on second screen}
% % \pgfpagesuselayout{2 on 1}[a4paper,border shrink=10mm]

% % \usepackage{ccicons}
% % \usepackage{fontawesome}
% % \usepackage{hyperref}

% % \usetheme{focus}
% % \usetheme{pureminimalistic}
% % \titlegraphic{\includegraphics[width=\textwidth/3]{./figures/brittney-weng-LycoZLp4jh0-unsplash}}
% % \title{Getting rid of the \texttt{console.log}}
% % \subtitle{Debugging RxJS-based Applications\\}
% % \author{\vspace{0.5cm}Manuel Alabor}
% % \institute{\small{Eastern Switzerland University\\ of Applied Sciences\\Rapperswil, Switzerland}}
% % \date{\vspace{0.9cm}\small{16 November 2020}}

% \begin{document}

% \section*{Welcome}

% \maketitle

% % \note[itemize]{
% % 	\item Welcome
% % 	\item Debugging of RxJS-based Applications
% % }

% % \begin{frame}{Who Am I?}
% % 	\begin{itemize}
% % 		\item Foobar\bigskip
% % 		\item Foobar\bigskip
% % 		\item Foobar\bigskip
% % 		\item Foobar\bigskip
% % 	\end{itemize}
% % \end{frame}

% % % \begin{frame}[focus]
% % % 	\includegraphics[width=\textwidth/3]{figures/rxjs-logo.png}
% % % 	\\
% % % 	\vspace{0.5cm}
% % % 	\Huge{RxJS}
% % % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \section*{Introduction}

% % \begin{frame}{}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \section{Interviews and War Stories}

% % \begin{frame}{Interviews}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{War Stories}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{Results}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \section{Observational Study}
% % \begin{frame}{Hypothesis}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{Design}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{Problem Example}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{Results}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \section{Conclusion}
% % \begin{frame}{}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % \begin{frame}{Future Work}

% % \end{frame}

% % \note[itemize] {
% % 	\item
% % }

% % % \note[itemize]{
% % % 	\item This was my first seminar paper done
% % % 	\item My goals:
% % % 	\item \textbf{Learn} scientific work style
% % % 	\item \textbf{Review} report
% % % 	\item \textbf{Transfer} a topic to frontend engineering
% % % 	\item \textbf{Write and present} my own research report\bigskip
% % % 	\item To do so, Markus came up with an interesting paper:
% % % }

% % % \begin{frame}[focus]
% % % 	\emph{``On the Positive Effect of Reactive Programming on Software Comprehension: An Empirical Study''}
% % % 	\\\bigskip
% % % 	\small{Salvaneschi et al. \cite{7827078}}
% % % \end{frame}

% % % \note[itemize]{
% % % 	\item On the positive effect of reactive programming on software comprehension
% % % 	\item An empirical study
% % % 	\item Research Group around Guido Salvaneschi
% % % 	\item 2015\bigskip
% % % }

% % % \subsection*{Research Questions}

% % % \begin{frame}{RQ1: Design Sciences and \\Empirical Software Engineering}
% % % 	\begin{enumerate}
% % % 		\item[RQ1.1] Which \textbf{empirical research methods}, approaches and concepts \textbf{were applied} by Salvaneschi et al. \cite{7827078}?\bigskip
% % % 		\item[RQ1.2] Are these research methods, approaches and concepts \textbf{applied well}, what could have been done better?\bigskip
% % % 		\item[RQ1.3] \textbf{Do} Salvaneschi et al. \cite{7827078} \textbf{meet FAIR principles} \cite{2019arXiv190805986H} \cite{wilkinson:2016} with their work?

% % % 	\end{enumerate}
% % % \end{frame}

% % % \note[itemize]{
% % % 	\item \textbf{Answer} 7 research questions
% % % 	\item Grouped in \textbf{2 sets} of research questions
% % % 	\item First: Research Principles
% % % 	\item Which \textbf{empirical research methods}?
% % % 	\item Are they \textbf{applied well}?
% % % 	\item Do they meet \textbf{FAIR} principles?
% % % }

% % % \section{Review of Salvaneschi et al.}
% % % \note[itemize]{
% % % 	\item In order to review Salvaneshi et al.:
% % % 	\item Present you two score cards
% % % 	\item One card per question set
% % % }

% % % \subsection*{Research Principle Score Card}
% % % \begin{frame}{Research Principle Score Card}
% % % 	\begin{exampleblock}{\faicon{thumbs-o-up} Empirical Software Engineering}
% % % 		\begin{itemize}
% % % 			\item \textbf{Careful} study \textbf{setup} and \textbf{execution}
% % % 			\item \textbf{Transparent} description of \textbf{threats to validity}
% % % 		\end{itemize}
% % % 	\end{exampleblock}

% % % 	\begin{alertblock}{\faicon{thumbs-o-down} Design Sciences}
% % % 		\textbf{No evidence} for the \textbf{application} of \textbf{design sciences} found
% % % 	\end{alertblock}

% % % 	\begin{alertblock}{\faicon{thumbs-o-down} FAIR Research Principles}
% % % 		Lack of \textbf{complementary information} violates \textbf{Reusability} principle
% % % 	\end{alertblock}
% % % \end{frame}

% % % \note[itemize]{
% % % 	\item \textbf{Empirical Software Engineering}: Well done, nothing to complain\bigskip
% % % 	\item \textbf{Design Sciences}: No clear evidence for use
% % % 	\item This is one study of many... Did they use Design Sciences for their high level planning?\bigskip
% % % 	\item \textbf{FAIR}: Good, except: Complementary information is missing
% % % 	\item Survey Tool and full code samples would improve value significantly
% % % }

% % % \subsection*{Synthetic Application}
% % % \begin{frame}{Synthetic Application: Scenario}
% % % 	\begin{columns}
% % % 		\column{0.6\textwidth}
% % % 			\begin{itemize}
% % % 				\item \textbf{Fetch} and \textbf{combine remote} profile and avatar \textbf{resource}\bigskip
% % % 				\item Avatar \textbf{depends} on profile resource\bigskip
% % % 			\end{itemize}

% % % 		\column{0.4\textwidth}
% % % 			\begin{block}{S1: \textbf{Modelling Data Flow}}
% % % 				\textbf{Fetch} a users \textbf{complete} profile \textbf{information} (name as well as their avatar picture) from \textbf{two} distinct \textbf{data sources}. The \textbf{profile} information \textbf{contains} required input to fetch the \textbf{avatar}. \textbf{Show} the combined information in the user interface \textbf{at once}.
% % % 			\end{block}
% % % 	\end{columns}
% % % \end{frame}

% % % \note[itemize]{
% % % 	\item Synthetic application: Model data flow
% % % 	\item Scenario: Show a users profile information with avatar
% % % 	\item Fetch an combine two remote data sources
% % % 	\item To fetch avatar, profile is required to be present\bigskip
% % % 	\item With reactive programming, this looks like:
% % % }

% % % \begin{frame}[focus]
% % % 	Q \& A
% % % \end{frame}

% % \appendix
% % \begin{frame}[allowframebreaks]{References}
% % 	\bibliographystyle{plain}
% % 	\bibliography{bibliography}
% % \end{frame}

% \end{document}

### 2020-11-02 slides.tex
Switch to Overleaf

@@
 \usepackage{tikz}
-\usetheme[showmaxslides, darkmode, nofooter]{pureminimalistic}
 
@@
 
-\usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2,style=numeric, sorting=none, url=false, eprint=false]{biblatex}
-\addbibresource{demo_bib.bib}
-
+\usetheme[darkmode, nofooterlogo]{pureminimalistic}
+% \renewcommand{\pageword}{}
+% \renewcommand{\logoheader}{\vspace{1.5em}}
+\usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2, style=numeric, sorting=none, url=false, eprint=false]{biblatex}
+\addbibresource{bibliography.bib}
 % this makes it possible to add backup slides, without counting them
@@
 \renewcommand{\appendixname}{\texorpdfstring{\translate{appendix}}{appendix}}
-\renewcommand{\logoheader}{}
 
-% if loaded after begin{document} a warning will appear: "pdfauthor already used"
-\title[Debugging RxJS-based Applications]{Getting Rid of the \texttt{console.log}:\\Debugging RxJS-based Applications}
+\title[Debugging RxJS-based Applications]{Getting rid of the \texttt{console.log}:\\Debugging RxJS-based Applications}
 \author{Manuel Alabor}
@@
 \begin{document}
-  \begin{frame}{Default Options}
-    Lorem ipsum dolor sit amet, consetetur sadipscing elitr,
-    sed diam nonumy eirmod tempor invidunt ut labore et dolore
-    magna aliquyam erat, sed diam voluptua.
-    At vero eos et accusam et justo duo dolores et ea rebum.
-    Stet clita kasd gubergren, no sea takimata sanctus est
-    Lorem ipsum dolor sit amet.
+
+\maketitle
+
+\note[itemize]{
+	\item Welcome
+	\item Debugging of RxJS-based Applications
+}
+
+\section*{Introduction}
+
+\begin{frame}{Who Am I?}
+	\begin{vfilleditems}
+		\item Foobar\bigskip
+		\item Foobar\bigskip
+		\item Foobar\bigskip
+		\item Foobar\bigskip
+	\end{vfilleditems}
 \end{frame}
+
+% \begin{frame}[focus]
+% 	\includegraphics[width=\textwidth/3]{figures/rxjs-logo.png}
+% 	\\
+% 	\vspace{0.5cm}
+% 	\Huge{RxJS}
+% \end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\section*{Introduction}
+
+\begin{frame}{}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\section{Interviews and War Stories}
+
+\begin{frame}{Interviews}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{War Stories}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{Results}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\section{Observational Study}
+\begin{frame}{Hypothesis}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{Design}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{Problem Example}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{Results}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\section{Conclusion}
+\begin{frame}{}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}{Future Work}
+
+\end{frame}
+
+\note[itemize] {
+	\item
+}
+
+\begin{frame}
+	Q \& A
+\end{frame}
+
+\appendix
+\begin{frame}[allowframebreaks]{References}
+	\bibliographystyle{plain}
+	\bibliography{bibliography}
+\end{frame}
+
 \end{document}
-
-% \documentclass[aspectratio=169]{beamer}
-% % should also look nice for the classic aspectratio
-% % of course, than the text has to be refitted
-% % \documentclass{beamer}
-% \usepackage[utf8]{inputenc}
-% \usepackage[T1]{fontenc}
-% \usepackage{tikz}
-% \usetheme[darkmode, showmaxslides]{pureminimalistic}
-% % \renewcommand{\pageword}{}
-% % \renewcommand{\logoheader}{\vspace{1.5em}}
-% \usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2,%
-%             style=numeric, sorting=none, url=false, eprint=false]{biblatex}
-% \addbibresource{bibliography.bib}
-% % this makes it possible to add backup slides, without counting them
-% \usepackage{appendixnumberbeamer}
-% \renewcommand{\appendixname}{\texorpdfstring{\translate{appendix}}{appendix}}
-
-% \title[short title]{This is the normal length of a research paper:
-% always longer than you would expect}
-% \author{Kai Norman Clasen}
-% \institute{Insitute name}
-% \date{\today}
-
-% % \documentclass{beamer}
-% % \documentclass[handout]{beamer}
-% % \usepackage{pgfpages}
-% % \setbeameroption{show notes on second screen}
-% % \pgfpagesuselayout{2 on 1}[a4paper,border shrink=10mm]
-
-% % \usepackage{ccicons}
-% % \usepackage{fontawesome}
-% % \usepackage{hyperref}
-
-% % \usetheme{focus}
-% % \usetheme{pureminimalistic}
-% % \titlegraphic{\includegraphics[width=\textwidth/3]{./figures/brittney-weng-LycoZLp4jh0-unsplash}}
-% % \title{Getting rid of the \texttt{console.log}}
-% % \subtitle{Debugging RxJS-based Applications\\}
-% % \author{\vspace{0.5cm}Manuel Alabor}
-% % \institute{\small{Eastern Switzerland University\\ of Applied Sciences\\Rapperswil, Switzerland}}
-% % \date{\vspace{0.9cm}\small{16 November 2020}}
-
-% \begin{document}
-
-% \section*{Welcome}
-
-% \maketitle
-
-% % \note[itemize]{
-% % 	\item Welcome
-% % 	\item Debugging of RxJS-based Applications
-% % }
-
-% % \begin{frame}{Who Am I?}
-% % 	\begin{itemize}
-% % 		\item Foobar\bigskip
-% % 		\item Foobar\bigskip
-% % 		\item Foobar\bigskip
-% % 		\item Foobar\bigskip
-% % 	\end{itemize}
-% % \end{frame}
-
-% % % \begin{frame}[focus]
-% % % 	\includegraphics[width=\textwidth/3]{figures/rxjs-logo.png}
-% % % 	\\
-% % % 	\vspace{0.5cm}
-% % % 	\Huge{RxJS}
-% % % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \section*{Introduction}
-
-% % \begin{frame}{}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \section{Interviews and War Stories}
-
-% % \begin{frame}{Interviews}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{War Stories}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{Results}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \section{Observational Study}
-% % \begin{frame}{Hypothesis}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{Design}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{Problem Example}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{Results}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \section{Conclusion}
-% % \begin{frame}{}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % \begin{frame}{Future Work}
-
-% % \end{frame}
-
-% % \note[itemize] {
-% % 	\item
-% % }
-
-% % % \note[itemize]{
-% % % 	\item This was my first seminar paper done
-% % % 	\item My goals:
-% % % 	\item \textbf{Learn} scientific work style
-% % % 	\item \textbf{Review} report
-% % % 	\item \textbf{Transfer} a topic to frontend engineering
-% % % 	\item \textbf{Write and present} my own research report\bigskip
-% % % 	\item To do so, Markus came up with an interesting paper:
-% % % }
-
-% % % \begin{frame}[focus]
-% % % 	\emph{``On the Positive Effect of Reactive Programming on Software Comprehension: An Empirical Study''}
-% % % 	\\\bigskip
-% % % 	\small{Salvaneschi et al. \cite{7827078}}
-% % % \end{frame}
-
-% % % \note[itemize]{
-% % % 	\item On the positive effect of reactive programming on software comprehension
-% % % 	\item An empirical study
-% % % 	\item Research Group around Guido Salvaneschi
-% % % 	\item 2015\bigskip
-% % % }
-
-% % % \subsection*{Research Questions}
-
-% % % \begin{frame}{RQ1: Design Sciences and \\Empirical Software Engineering}
-% % % 	\begin{enumerate}
-% % % 		\item[RQ1.1] Which \textbf{empirical research methods}, approaches and concepts \textbf{were applied} by Salvaneschi et al. \cite{7827078}?\bigskip
-% % % 		\item[RQ1.2] Are these research methods, approaches and concepts \textbf{applied well}, what could have been done better?\bigskip
-% % % 		\item[RQ1.3] \textbf{Do} Salvaneschi et al. \cite{7827078} \textbf{meet FAIR principles} \cite{2019arXiv190805986H} \cite{wilkinson:2016} with their work?
-
-% % % 	\end{enumerate}
-% % % \end{frame}
-
-% % % \note[itemize]{
-% % % 	\item \textbf{Answer} 7 research questions
-% % % 	\item Grouped in \textbf{2 sets} of research questions
-% % % 	\item First: Research Principles
-% % % 	\item Which \textbf{empirical research methods}?
-% % % 	\item Are they \textbf{applied well}?
-% % % 	\item Do they meet \textbf{FAIR} principles?
-% % % }
-
-% % % \section{Review of Salvaneschi et al.}
-% % % \note[itemize]{
-% % % 	\item In order to review Salvaneshi et al.:
-% % % 	\item Present you two score cards
-% % % 	\item One card per question set
-% % % }
-
-% % % \subsection*{Research Principle Score Card}
-% % % \begin{frame}{Research Principle Score Card}
-% % % 	\begin{exampleblock}{\faicon{thumbs-o-up} Empirical Software Engineering}
-% % % 		\begin{itemize}
-% % % 			\item \textbf{Careful} study \textbf{setup} and \textbf{execution}
-% % % 			\item \textbf{Transparent} description of \textbf{threats to validity}
-% % % 		\end{itemize}
-% % % 	\end{exampleblock}
-
-% % % 	\begin{alertblock}{\faicon{thumbs-o-down} Design Sciences}
-% % % 		\textbf{No evidence} for the \textbf{application} of \textbf{design sciences} found
-% % % 	\end{alertblock}
-
-% % % 	\begin{alertblock}{\faicon{thumbs-o-down} FAIR Research Principles}
-% % % 		Lack of \textbf{complementary information} violates \textbf{Reusability} principle
-% % % 	\end{alertblock}
-% % % \end{frame}
-
-% % % \note[itemize]{
-% % % 	\item \textbf{Empirical Software Engineering}: Well done, nothing to complain\bigskip
-% % % 	\item \textbf{Design Sciences}: No clear evidence for use
-% % % 	\item This is one study of many... Did they use Design Sciences for their high level planning?\bigskip
-% % % 	\item \textbf{FAIR}: Good, except: Complementary information is missing
-% % % 	\item Survey Tool and full code samples would improve value significantly
-% % % }
-
-% % % \subsection*{Synthetic Application}
-% % % \begin{frame}{Synthetic Application: Scenario}
-% % % 	\begin{columns}
-% % % 		\column{0.6\textwidth}
-% % % 			\begin{itemize}
-% % % 				\item \textbf{Fetch} and \textbf{combine remote} profile and avatar \textbf{resource}\bigskip
-% % % 				\item Avatar \textbf{depends} on profile resource\bigskip
-% % % 			\end{itemize}
-
-% % % 		\column{0.4\textwidth}
-% % % 			\begin{block}{S1: \textbf{Modelling Data Flow}}
-% % % 				\textbf{Fetch} a users \textbf{complete} profile \textbf{information} (name as well as their avatar picture) from \textbf{two} distinct \textbf{data sources}.
-The \textbf{profile} information \textbf{contains} required input to fetch the \textbf{avatar}.
-\textbf{Show} the combined information in the user interface \textbf{at once}.
-% % % 			\end{block}
-% % % 	\end{columns}
-% % % \end{frame}
-
-% % % \note[itemize]{
-% % % 	\item Synthetic application: Model data flow
-% % % 	\item Scenario: Show a users profile information with avatar
-% % % 	\item Fetch an combine two remote data sources
-% % % 	\item To fetch avatar, profile is required to be present\bigskip
-% % % 	\item With reactive programming, this looks like:
-% % % }
-
-% % % \begin{frame}[focus]
-% % % 	Q \& A
-% % % \end{frame}
-
-% % \appendix
-% % \begin{frame}[allowframebreaks]{References}
-% % 	\bibliographystyle{plain}
-% % 	\bibliography{bibliography}
-% % \end{frame}
-
-% \end{document}

### 2020-11-07 slides.tex
Intro

@@
 \usepackage{tikz}
+\usepackage{hyperref}
 
@@
 % \renewcommand{\pageword}{}
-% \renewcommand{\logoheader}{\vspace{1.5em}}
+\renewcommand{\logotitle}{}
+\renewcommand{\logoheader}{}
+\renewcommand{\logofooter}{}
 \usepackage[backend=biber, doi=false, maxbibnames=2, maxcitenames=2, style=numeric, sorting=none, url=false, eprint=false]{biblatex}
 \addbibresource{bibliography.bib}
-% this makes it possible to add backup slides, without counting them
 \usepackage{appendixnumberbeamer}
@@
 
-\title[Debugging RxJS-based Applications]{Getting rid of the \texttt{console.log}:\\Debugging RxJS-based Applications}
-\author{Manuel Alabor}
-\institute{Eastern Switzerland University of Applied Sciences}
+\title[Debugging of RxJS-based Applications]{Getting rid of the \texttt{console.log}:\\Debugging of RxJS-based Applications}
+\author{Manuel Alabor, Markus Stolze}
+\institute{OST in Rapperswil, Switzerland}
 \date{\today}
@@
 
-\note[itemize]{
-	\item Welcome
-	\item Debugging of RxJS-based Applications
+\note[itemize] {
+    \item Welcome everyone
+    \item I am Manuel
+    \item Frontend Software Engineer
+    \item Currently TypeScript, React, RxJS
+    \item Master Student
+    \item Research Topic: Debugging for Reactive Programming -> RxJS
+    \item -> Going to tell why this topic
 }
 
-\section*{Introduction}
+\section{Introduction}
 
-\begin{frame}{Who Am I?}
-	\begin{vfilleditems}
-		\item Foobar\bigskip
-		\item Foobar\bigskip
-		\item Foobar\bigskip
-		\item Foobar\bigskip
-	\end{vfilleditems}
-\end{frame}
-
-% \begin{frame}[focus]
-% 	\includegraphics[width=\textwidth/3]{figures/rxjs-logo.png}
-% 	\\
-% 	\vspace{0.5cm}
-% 	\Huge{RxJS}
-% \end{frame}
-
-\note[itemize] {
-	\item
-}
-
-\section*{Introduction}
-
-\begin{frame}{}
-
+\begin{frame}[fragile]{Reactive Programming for the Web}
+    \centering
+    \begin{figure}
+        \centering
+        \includegraphics[height=\textheight/2]{figures/rxjs-logo.png}
+        \caption{RxJS \cite{rxjs}}
+        \label{fig:rxjs}
+    \end{figure}
 \end{frame}
@@
 \note[itemize] {
-	\item
+    \item What is RxJS?
+    \item One of the larger representatives
+    \item RP with JavaScript
+    \item You might heard of Angular \cite{angualrrxjs} backed by Google
+    \item Angular is based on RxJS
+    \item All the things we love about RP: Check
+    \item All the things we... dont love about RP: Check
 }
 
-\begin{frame}{}
-
+\begin{frame}[fragile]{Debugging: First Intuition}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=\textheight/2]{figures/add-breakpoint.png}
+        \caption{Add a Breakpoint}
+    \end{figure}
 \end{frame}
@@
 \note[itemize] {
-	\item
+    \item How to debug a problem? Well...
+    \item Add a breakpoint
+    \item Imperative programming debugging
+    \item Helps sometimes (lambda functions, ...)
+    \item Often useless (stack traces, program step controls...)
+    \item What else?
+}
+
+\begin{frame}[fragile]{Debugging: The Reality}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=\textheight/2]{figures/consolelog.png}
+        \caption{Add Trace Logs}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item After some time with RxJS
+    \item Stuff gets more complex
+    \item Debugging === Trace Logs
+    \item Sprinkle log statements throughout the program
+    \item Hope to get the right hint
+    \item Remove (hopefully) everything again
+}
+
+\begin{frame}[fragile]{Debugging: Can this be it?}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=\textheight/2]{figures/person-shrugging_1f937.png}
+        \caption{\tiny{Source: \url{https://emojipedia.org/person-shrugging/}}}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item Frustration -> Reason Research topic
+    \item There are concepts and solutions for RP debugging
+    \item Reactive Debugger, Scala \cite{10.1145/2884781.2884815}
+    \item Improved logging tools rxjs-spy
+    \item Visualizers rxviz, rxjs visualizer
+    \item But why does nobody seem to use them?
+    \item Why are they not popular?
+    \item -> Research Questions
+}
+
+\begin{frame}[fragile]{Research Questions}
+    \begin{enumerate}
+        \vfill\item[RQ1] \textbf{What challenges} do software engineers face when debugging RxJS-based applications?
+        \vfill\item[RQ2] How can the \textbf{experience} of software engineers during the debugging process of RxJS-based applications \textbf{be improved}?
+        \vfill\item[\color{gray}{RQ3}] \color{gray}{What is the \textbf{impact of proposed solutions} on the debugging experience of software engineers?}
+    \end{enumerate}
+\end{frame}
+
+\note[itemize] {
+    \item First: Collect and Validate information
+    \item How do engineers debug RxJS?
+    \item What do they like? Dislike?
+    \item Validate this data in an observational study
+    \item Second: Propose a solution to improve
+    \item Third: Implement and validate solution
+    \item Not part of this work
+    \item Already started with in current iteration
 }
@@
 
-\begin{frame}{Interviews}
-
+\begin{frame}[fragile]{Call for ``War Stories''}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.6\textheight]{figures/tweet-interview.png}
+        \caption{\tiny{Source: \url{https://twitter.com/swissmanu/status/1242429409208029185}}}
+    \end{figure}
 \end{frame}
@@
 \note[itemize] {
-	\item
+    \item First: Collect and Validate information
+    \item How do engineers debug RxJS?
+    \item What do they like? Dislike?
+    \item Validate this data in an observational study
+    \item Second: Propose a solution to improve
+    \item Third: Implement and validate solution
+    \item Not part of this work
+    \item Already started with in current iteration
 }
 
-\begin{frame}{War Stories}
-
+\begin{frame}[fragile]{Getting Started with RQ1}
+	\begin{vfilleditems}
+		\item Talk to practitioners
+		\item 1:1 Interviews
+		\item Written ``War Story'' Reports
+	\end{vfilleditems}
 \end{frame}
 
-\note[itemize] {
-	\item
-}
+\begin{frame}{Interviews}
+	\begin{vfilleditems}
+		\item Informal interviews with five participants
+	\end{vfilleditems}
+\end{frame}
+
+\begin{frame}{War Stories}
+	\begin{vfilleditems}
+		\item Request via Twitter and direct emails
+		\item Responses from five engineers, including one RxJS core team member
+	\end{vfilleditems}
+\end{frame}
 
@@
 
-\note[itemize] {
-	\item
-}
-
 \section{Observational Study}
 \begin{frame}{Hypothesis}
-
+    If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
 \end{frame}
 
-\note[itemize] {
+\begin{frame}{Study Design}
+    \begin{vfilleditems}
+        \item Observational Study
+        \item Two Problems
+        \item 25 Minutes per Problem
         \item 
-}
-
-\begin{frame}{Design}
-
+    \end{vfilleditems}
 \end{frame}
-
-\note[itemize] {
-	\item
-}
 
@@
 
-\note[itemize] {
-	\item
-}
-
 \begin{frame}{Results}
@@
 \end{frame}
-
-\note[itemize] {
-	\item
-}
 
@@
 
-\note[itemize] {
-	\item
-}
-
-\begin{frame}{Future Work}
-
+\section{Future Work}
+\begin{frame}[fragile]{Future Work}
+    \begin{vfilleditems}
+        \item Manuel Alabor \cite{10.1145/2501654.2501666}
+        \item Manuel Alabor
+        \item Manuel Alabor
+    \end{vfilleditems}
 \end{frame}
 
-\note[itemize] {
-	\item
-}
-
-\begin{frame}
-	Q \& A
+\begin{frame}[plain, noframenumbering]
+  \centering
+  \vfill
+  {\fontsize{40}{50}\selectfont Q \& A}
+  \vfill
 \end{frame}
@@
 \begin{frame}[allowframebreaks]{References}
-	\bibliographystyle{plain}
-	\bibliography{bibliography}
+	\printbibliography
 \end{frame}

### 2020-11-08 slides.tex
Study

@@
 \usepackage{hyperref}
+\usepackage{pgfplots}
+\usepackage{pgfpages}
 
-\usepackage{pgfpages}
+\pgfplotsset{compat=1.17}
 \setbeameroption{show notes on second screen}
@@
 \title[Debugging of RxJS-based Applications]{Getting rid of the \texttt{console.log}:\\Debugging of RxJS-based Applications}
-\author{Manuel Alabor, Markus Stolze}
+\author{Manuel Alabor and Markus Stolze}
 \institute{OST in Rapperswil, Switzerland}
@@
 \section{Introduction}
+
+\subsection{Why?}
 
@@
 
+\subsection{Research Questions}
 \begin{frame}[fragile]{Research Questions}
@@
     \item First: Collect and Validate information
-    \item How do engineers debug RxJS?
-    \item What do they like? Dislike?
+    \item Interviews and War Story Reports
     \item Validate this data in an observational study
@@
 
+\subsection{Interviews}
+
+\begin{frame}[fragile]{One-to-One Interviews}
+	\begin{vfilleditems}
+		\item Interviewed five \textbf{Engineers}
+		\item How do they \textbf{use} Reactive Programming?
+		\item What do they \textbf{like},
+		\item ... and what \textbf{dislike} about it?
+	\end{vfilleditems}
+\end{frame}
+
+\note[itemize] {
+    \item We were interested in the overall, ...
+    \item the good ...
+    \item and bad experiences
+    \item Not yet debugging specific
+    \item Recruited through personal network
+    \item Talked to five engineers
+    \item 4 RxJS, 1 Scala Akka Streams
+    \item Most knew RxJS through Angular
+}
+
+\begin{frame}{Quotes}
+    \begin{vfilleditems}
+		\item \textit{``[Reactive Programming is] a good way for \textbf{composing} multiple data sources.''}
+		\item \textit{``In \textbf{99\%} of all cases, I add \textbf{console.log} statements [\dots] trying to understand what's happening.''}
+	\end{vfilleditems}
+\end{frame}
+
+\note[itemize] {
+    \item All interviewees were overall positive to RP
+    \item All value RP for complex data \textbf{dependencies}
+    \item and \textbf{transformations}
+    \item Concrete Quote about console log
+    \item Off the record: ``Debugging RP?''
+    \item ... well, then you are f***
+}
+
+\subsection{War Stories}
 \begin{frame}[fragile]{Call for ``War Stories''}
@@
 \note[itemize] {
-    \item First: Collect and Validate information
-    \item How do engineers debug RxJS?
-    \item What do they like? Dislike?
-    \item Validate this data in an observational study
-    \item Second: Propose a solution to improve
-    \item Third: Implement and validate solution
-    \item Not part of this work
-    \item Already started with in current iteration
+    \item After first "overall" data collection
+    \item Interested in specific data about RxJS Debugging
+    \item Twitter and direct-email
+    \item 5 replies
+    \item RxJS core team member
+    \item Google developer experts
+    \item Fin Tech Engineers
 }
 
-\begin{frame}[fragile]{Getting Started with RQ1}
+\begin{frame}[fragile]{Report Extracts}
 	\begin{vfilleditems}
-		\item Talk to practitioners
-		\item 1:1 Interviews
-		\item Written ``War Story'' Reports
+		\item Traditional debugging utilities are \textbf{seldom sufficient}
+		\item Tendency to \textbf{manual} code \textbf{modifications}:
+		\begin{vfilleditems}
+		    \item Add \textbf{trace logs} with \texttt{console.log}
+		    \item \textbf{Extract} code parts to \textbf{sand boxed} environments
+		\end{vfilleditems}
 	\end{vfilleditems}
@@
 
-\begin{frame}{Interviews}
+\note[itemize] {
+    \item Foo
+}
+
+\subsection{Insights}
+
+\begin{frame}[fragile]{Key Take Aways}
+    \begin{columns}[T]
+        \begin{column}{.6\linewidth}
+            \vspace{3em}
+            \begin{itemize}
+                \item The ``\textbf{to hammer a screw}''-problem\bigskip
+                \item Tendency to \textbf{manual} code \textbf{modifications}\bigskip
+                \item \textbf{Pessimistic} opinions on debugging of RxJS-based applications
+	        \end{itemize}
+        \end{column}
+        \begin{column}{.4\linewidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=.7\linewidth]{figures/hammer-screw.jpeg}
+                \caption{\tiny{Source: Manuel Alabor}}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\note[itemize] {
+    \item ... todo
+    \item Based on this
+    \item Observational study to validate data
+    \item Using the following hypothesis:
+}
+
+\section{Observational Study}
+\subsection{Study Design}
+
+\begin{frame}[fragile]{Hypothesis}
+    \begin{quote}
+        If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
+    \end{quote}
+\end{frame}
+
+\note[itemize] {
+    \item Foo
+}
+
+\begin{frame}{Study Design}
     \begin{vfilleditems}
-		\item Informal interviews with five participants
+        \item \textbf{Observational Study}
+        \item \textbf{Briefing} upfront for every subject
+        \item \textbf{Uncontrolled} Environment
+        \item \textbf{Remote} moderation and execution
+        \item ``After Action'' \textbf{Survey}
     \end{vfilleditems}
@@
 
-\begin{frame}{War Stories}
+\note[itemize] {
+    \item Remotely conducted observational study
+    \item Execution on subjects computer
+    \item Individual dev environment
+    \item Mimic "Daily dev routine"
+    \item Interested in "How would you tackle RxJS debugging"
+    \item Two problems, 25 minutes each
+    \item ~1h total
+    \item Unattended survey afterwards -> RxJS experience data
+}
+
+\begin{frame}{Problem 1: Example}
+    \begin{columns}[b]
+        \begin{column}{.5\linewidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=.9\linewidth]{figures/problem1-code.png}
+                \caption{RxJS Observable}
+            \end{figure}
+        \end{column}
+        \begin{column}{.5\linewidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=.9\linewidth]{figures/problem1-ui.png}
+                \caption{User Interface}
+            \end{figure}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\note[itemize] {
+    \item A problem consisted of:
+    \item Provided: \textbf{Code}, \textbf{Unit Tests}, \textbf{UI}
+    \item Task: Identify and Resolve Bugs
+    \item Task: "Speak out loud"
+    \item Of course: we did not care IF they actually resolve a problem
+    \item Collect as much data how subjects debug
+}
+
+\subsection{Study Execution}
+\begin{frame}{Subject Population}
     \begin{vfilleditems}
-		\item Request via Twitter and direct emails
-		\item Responses from five engineers, including one RxJS core team member
+        \item \textbf{Four} professional software engineers
+        \item Two currently \textbf{work} with RxJS
+        \item Three have \textbf{two or more years} of \textbf{experience} with RxJS
+        \item One of them uses RxJS to develop \textbf{backend applications}
     \end{vfilleditems}
@@
 
-\begin{frame}{Results}
+\note[itemize] {
+    \item Four engineers recruited from personal network
+    \item All use RxJS for frontend development
+    \item Only one for backend
+    \item Experienced population; most >2y
+    \item All: Trace logs
+    \item Two: Specialized loggers
+    \item Three: Debugger
+}
 
+\subsection{Study Results}
+
+\begin{frame}[fragile]{Observed Debugging Techniques}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+                    height=0.6\textheight,
+                    width=0.8\linewidth,
+                    xbar,
+                    xtick={0,1,2,3,4},
+                    xlabel={\# of Subjects},
+                    symbolic y coords={Add. Tools,Debugger,Trace Logs},
+                    ytick=data,
+                    enlarge y limits=0.5,
+                    enlargelimits=0.15,
+                    nodes near coords, nodes near coords align={horizontal}
+                ]
+                \addplot[
+                        fill=pureminimalistic@text@red,
+                        draw=pureminimalistic@text@red
+                    ]
+                    coordinates {
+                        (0,Add. Tools) (4,Trace Logs) (2,Debugger) 
+                    };
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
 \end{frame}
 
-\section{Observational Study}
-\begin{frame}{Hypothesis}
-    If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
-\end{frame}
+\note[itemize] {
+    \item All modified/augmented source code manually
+    \item -> Trace logs
+    \item Two tried traditional debugging
+    \item No one extracted code
+    \item No one used additional libraries
+    \item On told afterwards: Would have been their next step
+    \item Based on this evidence: Hypothesis is correct
+}
 
-\begin{frame}{Study Design}
+\begin{frame}[fragile]{Interpretation}
     \begin{vfilleditems}
-        \item Observational Study
-        \item Two Problems
-        \item 25 Minutes per Problem
-        \item 
+        \item \textbf{Four} professional software engineers
+        \item Two currently \textbf{work} with RxJS
+        \item Three have \textbf{two or more years} of \textbf{experience} with RxJS
+        \item One of them uses RxJS to develop \textbf{backend applications}
     \end{vfilleditems}
@@
 
-\begin{frame}{Problem Example}
-
-\end{frame}
-
-\begin{frame}{Results}
-
-\end{frame}
+\note[itemize] {
+    \item All used manual trace logs
+    \item Two tried traditional debugging
+    \item No one extracted code
+    \item No one used additional libraries
+    \item On told afterwards: Would have been their next step
+}
 

### 2020-11-08 slides.tex
Future Work

@@
     \item Frustration -> Reason Research topic
-    \item There are concepts and solutions for RP debugging
+    \item There are good concepts and solutions for RP debugging
     \item Reactive Debugger, Scala \cite{10.1145/2884781.2884815}
-    \item Improved logging tools rxjs-spy
-    \item Visualizers rxviz, rxjs visualizer
+    \item Improved logging tools rxjs-spy \cite{rxjsspy}
+    \item Visualizers rxviz \cite{rxviz}
+    \item Visualizers rxfiddle \cite{10.1145/3180155.3180156}
     \item But why does nobody seem to use them?
-    \item Why are they not popular?
+    \item Why unpopular?
     \item -> Research Questions
@@
     \begin{vfilleditems}
-		\item \textit{``[Reactive Programming is] a good way for \textbf{composing} multiple data sources.''}
-		\item \textit{``In \textbf{99\%} of all cases, I add \textbf{console.log} statements [\dots] trying to understand what's happening.''}
+		\item[\Large{``}] \textit{[Reactive Programming is] a good way for \textbf{composing} multiple data sources.}
+		\item[\Large{``}] \textit{In \textbf{99\%} of all cases, I add \textbf{console.log} statements [\dots] trying to understand what's happening.}
 	\end{vfilleditems}
@@
 \note[itemize] {
-    \item Foo
+    \item Again: Common practice to modify code
+    \item Not only trace logs
+    \item Extraction of code parts to extraneous tools
+    \item Visualizers rx-viz \cite{rxviz}
+    \item Code sandboxes StackBlitz
+    \item Try to use debugger
+    \item Debuggers often not that useful
 }
@@
                 \item Tendency to \textbf{manual} code \textbf{modifications}\bigskip
-                \item \textbf{Pessimistic} opinions on debugging of RxJS-based applications
+                \item Rather \textbf{negative} experience with debugging RxJS-based applications
 	        \end{itemize}
@@
 \note[itemize] {
-    \item ... todo
-    \item Based on this
-    \item Observational study to validate data
-    \item Using the following hypothesis:
+    \item Traditional debugger feels wrong
+    \item Ask a question which they cannot answer (hammer/screw) -> Stack trace
+    \item Fallback to manual code modification
+    \item Trace log statements
+    \item Extract pieces to sandboxes
+    \item Overall: Pessimistic view on debugging RxJS RP
+    \item Based on this:
+    \item Study to validate this...
+    \item Hypothesis coming up:
 }
@@
 \begin{frame}[fragile]{Hypothesis}
-    \begin{quote}
-        If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
-    \end{quote}
+    \begin{vfilleditems}
+        \item[\Large{``}] If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
+    \end{vfilleditems}
 \end{frame}
@@
 \note[itemize] {
-    \item Foo
+    \item Manual intervention repeating theme
+    \item Both Interviews and War Stories
+    \item Thats why focus on this in validation study
+    \item Dont care what kind (logs vs extraction)
 }
@@
     \item No one extracted code
-    \item No one used additional libraries
-    \item On told afterwards: Would have been their next step
-    \item Based on this evidence: Hypothesis is correct
+    \item No one used additional tools
+    \item One told afterwards: Would have been their next step
+    \item Based on experiment data: Hypothesis is correct
 }
+
+\section{Conclusion}
 
@@
     \begin{vfilleditems}
-        \item \textbf{Four} professional software engineers
-        \item Two currently \textbf{work} with RxJS
-        \item Three have \textbf{two or more years} of \textbf{experience} with RxJS
-        \item One of them uses RxJS to develop \textbf{backend applications}
+        \item Engineers \textbf{know} RP debugging tools, but \textbf{do not use} them
+        \item Engineers use \textbf{imperative}, but actually require ``\textbf{reactive}'' debuggers
+        \item \textbf{Manual intervention} is the ``\textbf{last resort}''
     \end{vfilleditems}
@@
 \note[itemize] {
-    \item All used manual trace logs
-    \item Two tried traditional debugging
-    \item No one extracted code
-    \item No one used additional libraries
-    \item On told afterwards: Would have been their next step
+    \item Interviews + War Stories + Study = ?
+    \item 7 of 14 peers know specific RxJS debugging tools
+    \item But they dont use them
+    \item Instead: Traditional tools \textbf{easily accessible}
+    \item Final way to go:
+    \item Trace logs and code extraction
 }
 
-\section{Conclusion}
-\begin{frame}{}
-
+\begin{frame}[fragile]{Answer}
+    \begin{enumerate}
+        \item[\color{gray}{RQ1}] \color{gray}{What challenges do software engineers face when debugging RxJS-based applications?}
+        \vfill\item[\Large{``}] \color{white}{The most significant challenge software engineers face [\dots] is to know \textbf{when} they should apply \textbf{what tool} to resolve their current problem in the \textbf{most efficient} way.}
+    \end{enumerate}
 \end{frame}
 
+\note[itemize] {
+    \item Challenge: What tool, when and how
+    \item Often: The "next best" as we saw: Traditional tools
+    \item Often: Modify code
+    \item Less often: Specific tools
+    \item What can we do about it?
+    \item Answer question 2
+}
+
+\begin{frame}[fragile]{Answer}
+    \begin{enumerate}
+        \item[\color{gray}{RQ2}] \color{gray}{How can the experience of software engineers during the debugging process of RxJS-based applications be improved?}
+        \vfill\item[\Large{``}] \color{white}{[\dots] \textbf{improve} the \textbf{experience} of debugging [\dots] by providing RP specific debugging utilities where software engineers expect them the most: \textbf{Fully integrated} [\dots] their IDE [\dots]}
+    \end{enumerate}
+\end{frame}
+
+\note[itemize] {
+    \item Improve the dev user experience
+    \item Lower the boundary to use the right tool
+    \item At the right time
+    \item Integrate the right tools where they are expected the most
+    \item Example: Async Stack Traces in Google Chrome \cite{chromeasync}
+}
+
 \section{Future Work}
+
 \begin{frame}[fragile]{Future Work}
     \begin{vfilleditems}
-        \item Manuel Alabor \cite{10.1145/2501654.2501666}
-        \item Manuel Alabor
-        \item Manuel Alabor
+        \item \textbf{Propose} and \textbf{implement} solution to \textbf{prevent} manual code \textbf{modification}
+        \item \textbf{Evaluate} proposal on \textbdf{effectiveness} to answer RQ3
     \end{vfilleditems}
 \end{frame}
+
+\note[itemize] {
+    \item Wraps up our plans for the future
+    \item Work on a concrete prototype
+    \item Find solution for trace log problem
+    \item Without manual intervention
+    \item Validate effectiveness and answer RQ3
+}
+
+\begin{frame}[fragile]{Sneak Preview}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.6\textheight]{figures/tweet-rxjs-debugging-extension.png}
+        \caption{\tiny{Source: \url{https://twitter.com/swissmanu/status/1313412280072232960}}}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item Small sneak preview
+    \item Things are progressing
+    \item Thank you for your time and attention!
+}
 
@@
   \vfill
-  {\fontsize{40}{50}\selectfont Q \& A}
+  {\fontsize{40}{50}\selectfont\color{white}{Q \& A}}
   \vfill
 \end{frame}
+
+\note[itemize] {
+    \item Happy for input
+    \item Happy to answer questions
+}
 

### 2020-11-11 slides.tex
Restructure

--- 2020-11-07 slides.tex
@@
 \usepackage{hyperref}
+\usepackage{pgfplots}
+\usepackage{pgf-pie}
+\usepackage{pgfpages}
 
-\usepackage{pgfpages}
+\pgfplotsset{compat=1.17}
+\usetikzlibrary{patterns}
 \setbeameroption{show notes on second screen}
@@
 
-\title[Debugging of RxJS-based Applications]{Getting rid of the \texttt{console.log}:\\Debugging of RxJS-based Applications}
-\author{Manuel Alabor, Markus Stolze}
+\title[Debugging of RxJS-based Applications]{The State of Debugging for RxJS\\``Debugging of RxJS-Based Applications''}
+\author{Manuel Alabor and Markus Stolze}
 \institute{OST in Rapperswil, Switzerland}
@@
 \note[itemize] {
-    \item Welcome everyone
+    \item Welcome
     \item I am Manuel
-    \item Frontend Software Engineer
-    \item Currently TypeScript, React, RxJS
-    \item Master Student
-    \item Research Topic: Debugging for Reactive Programming -> RxJS
-    \item -> Going to tell why this topic
+    \item Master Student, OST, Switzerland
+    \item I want to tell you a short story...
+    \item State of debugging for RxJS
+    \item You might think, well:
+    \item ``It has JS in its name''
+    \item JavaScript and Engineering?
 }
@@
 
+\subsection{Why?}
+
 \begin{frame}[fragile]{Reactive Programming for the Web}
-    \centering
     \begin{figure}
         \centering
-        \includegraphics[height=\textheight/2]{figures/rxjs-logo.png}
-        \caption{RxJS \cite{rxjs}}
+        \includegraphics[height=0.6\textheight]{figures/frontend-engineers.jpg}
         \label{fig:rxjs}
@@
 \note[itemize] {
-    \item What is RxJS?
-    \item One of the larger representatives
-    \item RP with JavaScript
-    \item You might heard of Angular \cite{angualrrxjs} backed by Google
-    \item Angular is based on RxJS
-    \item All the things we love about RP: Check
-    \item All the things we... dont love about RP: Check
+    \item Thats frontend stuff!
+    \item Are ``Frontend Engineers'' Engineers?
+    \item Because of things like RxJS...
+    \item Frontend Engineering came a long way!
+    \item RxJS is RP with JavaScript
+    \item Nice: TypeScript Interface
+    \item Nice: RxJS is core part of Googles Angular
+    \item Nice: Describe data flows
+    \item Wait? Sales pitch?
+    \item I have a story of frustration to tell: Debugging
 }
@@
 \note[itemize] {
-    \item How to debug a problem? Well...
-    \item Add a breakpoint
-    \item Imperative programming debugging
-    \item Helps sometimes (lambda functions, ...)
-    \item Often useless (stack traces, program step controls...)
+    \item Back when i started with RP
+    \item Debug? Ha! Breakpoint!
+    \item Helped sometimes (lambda functions, ...)
+    \item Often felt lost though
+    \item Stack traces
+    \item program step controls: Operator? No: Internals!
     \item What else?
@@
 \note[itemize] {
-    \item After some time with RxJS
-    \item Stuff gets more complex
+    \item Especially when stuff gets complicated
     \item Debugging === Trace Logs
-    \item Sprinkle log statements throughout the program
-    \item Hope to get the right hint
-    \item Remove (hopefully) everything again
+    \item Sprinkle log statements
+    \item Analyze flood of logs
+    \item Find problem
+    \item Remove log states
+    \item DONT remove the bug fix ;)
+    \item This left me...
 }
 
-\begin{frame}[fragile]{Debugging: Can this be it?}
+\begin{frame}[fragile]{Really?}
     \begin{figure}[H]
@@
 \note[itemize] {
-    \item Frustration -> Reason Research topic
-    \item There are concepts and solutions for RP debugging
-    \item Reactive Debugger, Scala \cite{10.1145/2884781.2884815}
-    \item Improved logging tools rxjs-spy
-    \item Visualizers rxviz, rxjs visualizer
-    \item But why does nobody seem to use them?
-    \item Why are they not popular?
-    \item -> Research Questions
+    \item Frustrated!
+    \item Mature Software Engineer has to feel different
+    \item Looked for existing solutions
+    \item Good concepts for other languages (Scala)
+    \item RxJS less stuff
+    \item Lets do this on our own!
+    \item -> Started with research questions
 }
+
+\subsection{Research Questions}
 
@@
         \vfill\item[RQ2] How can the \textbf{experience} of software engineers during the debugging process of RxJS-based applications \textbf{be improved}?
-        \vfill\item[\color{gray}{RQ3}] \color{gray}{What is the \textbf{impact of proposed solutions} on the debugging experience of software engineers?}
     \end{enumerate}
@@
 \note[itemize] {
-    \item First: Collect and Validate information
-    \item How do engineers debug RxJS?
-    \item What do they like? Dislike?
-    \item Validate this data in an observational study
-    \item Second: Propose a solution to improve
-    \item Third: Implement and validate solution
-    \item Not part of this work
-    \item Already started with in current iteration
+    \item Before we can improve, if at all
+    \item We need data on what we should improve
+    \item Thats question 1
+    \item Based on this
+    \item Propose a concept to make dev experience better
+    \item Thats question 2
+    \item Lets get started!
 }
 
-\section{Interviews and War Stories}
+\section{Studies}
 
-\begin{frame}[fragile]{Call for ``War Stories''}
+\begin{frame}[fragile]{Road Map}
+    \begin{vfilleditems}
+        \item Two qualitative studies
+        \begin{enumerate}
+            \vfill\item Preparatory study using \textbf{interviews} and \textbf{``war story'' reports}
+            \vfill\item \textbf{Controlled experiment} to validate previously identified findings
+        \end{enumerate}
+    \end{vfilleditems}
+\end{frame}
+
+\note[itemize] {
+    \item Worked with 14 software engineering professionals
+    \item Conducted interviews
+    \item Collected written reports
+    \item Then validated this with controlled experiment
+}
+
+\subsection{Study 1: Interviews and War Stories}
+
+\begin{frame}[plain, noframenumbering]{Study 1: Preparation}
+  \centering
+  \vfill
+  {\fontsize{30}{40}\selectfont\color{white}{Interviews\\and War Stories}}
+  \vfill
+\end{frame}
+
+\begin{frame}[fragile]{Call for Participants}
     \begin{figure}[H]
@@
 \note[itemize] {
-    \item First: Collect and Validate information
-    \item How do engineers debug RxJS?
-    \item What do they like? Dislike?
-    \item Validate this data in an observational study
-    \item Second: Propose a solution to improve
-    \item Third: Implement and validate solution
-    \item Not part of this work
-    \item Already started with in current iteration
+    \item Recruiting via Twitter and Email
+    \item Personal Network
+    \item 10 participants
+    \item Daily users of RP
+    \item One RxJS Core team member
+    \item Two Google Dev Experts on Angular
 }
 
-\begin{frame}[fragile]{Getting Started with RQ1}
+\begin{frame}[fragile]{Five One-to-One Interviews}
 	\begin{vfilleditems}
-		\item Talk to practitioners
-		\item 1:1 Interviews
-		\item Written ``War Story'' Reports
+		\item How do you \textbf{use} Reactive Programming?
+		\item Which aspects do you \textbf{like} on Reactive Programming?
+		\item Which do you \textbf{dislike}?
 	\end{vfilleditems}
@@
 
-\begin{frame}{Interviews}
+\note[itemize] {
+    \item Interviews had no focus on debugging yet
+    \item Interested in experience with RP
+    \item Pros
+    \item Cons
+    \item Pain Points
+}
+
+\begin{frame}[fragile]{Five Written War Story Reports}
 	\begin{vfilleditems}
-		\item Informal interviews with five participants
+		\item When did you \textbf{last} have to debug an RxJS-based Application?
+		\item Do you remember a specific \textbf{challenging} debugging situation?
+		\item How did you \textbf{solve} it? (If at all?)
 	\end{vfilleditems}
@@
 
-\begin{frame}{War Stories}
+\note[itemize] {
+    \item This time: Focused on debugging
+    \item Hear about specific situations
+    \item They should recall particular scenario
+    \item Step by step
+    \item What problems did they face tool?
+    \item How did they overcome them?
+}
+
+\begin{frame}[fragile]{Mentioned Debugging Techniques}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \pie[
+                cloud,
+                sum=10,
+                after number={\textvisiblespace times},
+                text=inside,
+                radius=3,
+                rotate=100,
+                color={pureminimalistic@text@red}
+            ]{
+                3/Debugger,
+                5/Trace Logs,
+                2/Add. Tool
+            }
+        \end{tikzpicture}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item Lengthy answers
+    \item Repeating themes
+    \item RP is awesome for composing complex data flows
+    \item That aside
+    \item Debugging is hard
+    \item Unsatisfying traditional debuggers
+    \item Most often manual trace logs
+    \item Or code extraction to to external tools
+}
+
+{
+\setbeamercolor{background canvas}{bg=pureminimalistic@text@white}
+\setbeamercolor{frametitle}{bg=pureminimalistic@text@white}
+\setbeamercolor{normal text}{fg=pureminimalistic@text@black}
+\begin{frame}[fragile, plain]{Key Take Aways}
+    \begin{figure}
+        \centering
+        \includegraphics[width=.7\textheight]{figures/rxjs-red-reveal-pattern.png}
+        \caption{\tiny{Source: Manuel Alabor}}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item What do we get from this?
+    \item Engineers use manual code modification
+    \item Either trace logs
+    \item Or extraction to external tools
+    \item But why?
+    \item Because traditional tools fail them
+    \item These tools cant see through RxJS abstractions
+}
+
+\begin{frame}[fragile, plain]{Key Take Aways}
+    \begin{figure}
+        \centering
+        \includegraphics[width=.7\textheight]{figures/rxjs-red-reveal-pattern-lens.png}
+        \caption{\tiny{Source: Manuel Alabor}}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item Hence, if given the correct filter
+    \item In other words, a tool that sees through RxJS
+    \item They could debug with less manual intervention
+    \item Improve efficiency
+    \item Improve quality because no code changes
+    \item If manual code modification is really common practice...
+}
+}
+
+\section{Study 2: Observational Study}
+
+\begin{frame}[plain, noframenumbering]{Study 2: Observational Study}
+    \centering
+    \vfill
+    {\fontsize{30}{40}\selectfont\color{white}{Controlled Experiment}}
+    \vfill
+\end{frame}
+
+\note[itemize] {
+    \item Is something we had to validate
+    \item In a Controlled Experiment
+    \item Obviously, we created hypothesis after previous findings...
+}
+
+\subsection{Study Design}
+
+\begin{frame}[fragile]{Hypothesis}
     \begin{vfilleditems}
-		\item Request via Twitter and direct emails
-		\item Responses from five engineers, including one RxJS core team member
+        \item[\Large{``}] If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
     \end{vfilleditems}
@@
 
-\begin{frame}{Results}
-
-\end{frame}
-
-\section{Observational Study}
-\begin{frame}{Hypothesis}
-    If software engineers must solve an RxJS-based problem, then they \textbf{will instrument the code manually} in order to understand its behavior.
-\end{frame}
+\note[itemize] {
+    \item Open to all manual modifications
+    \item trace logs
+    \item extracts
+}
 
@@
     \begin{vfilleditems}
-        \item Observational Study
-        \item Two Problems
-        \item 25 Minutes per Problem
-        \item 
+        \item \textbf{Observational Study}
+        \item \textbf{Briefing} upfront for every subject
+        \item \textbf{Uncontrolled} Environment
+        \item \textbf{Remote} moderation and execution
+        \item ``After Action'' \textbf{Survey}
     \end{vfilleditems}
@@
 
-\begin{frame}{Problem Example}
+\note[itemize] {
+    \item Remotely conducted observational study
+    \item Execution on subjects computer
+    \item Individual dev environment
+    \item Mimic "Daily dev routine"
+    \item Two problems, 25 minutes each
+    \item ~1h total
+    \item Unattended survey afterwards -> RxJS experience data
+}
 
+\begin{frame}{Problem 1: Example}
+    \begin{columns}[b]
+        \begin{column}{.5\linewidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=.9\linewidth]{figures/problem1-code.png}
+                \caption{RxJS Observable}
+            \end{figure}
+        \end{column}
+        \begin{column}{.5\linewidth}
+            \begin{figure}
+                \centering
+                \includegraphics[width=.9\linewidth]{figures/problem1-ui.png}
+                \caption{User Interface}
+            \end{figure}
+        \end{column}
+    \end{columns}
 \end{frame}
 
-\begin{frame}{Results}
+\note[itemize] {
+    \item A problem consisted of:
+    \item TypeScript \textbf{Code}, \textbf{Unit Tests}, runnable \textbf{UI}
+    \item Goal: "We want to know how you debug!"
+    \item Task: Identify and Resolve Bugs
+    \item Task: "Speak out loud"
+}
 
+\subsection{Study Execution}
+
+\begin{frame}{Subject Population}
+    \begin{vfilleditems}
+        \item \textbf{Four} professional software engineers
+        \item Two currently \textbf{work} with RxJS
+        \item Three have \textbf{two or more years} of \textbf{experience} with RxJS
+        \item One of them uses RxJS to develop \textbf{backend applications}
+    \end{vfilleditems}
 \end{frame}
 
-\section{Conclusion}
-\begin{frame}{}
+\note[itemize] {
+    \item Four engineers recruited from personal network
+    \item Experienced population; most >2y
+    \item All use RxJS for frontend development
+    \item One also for backend
+}
 
+\subsection{Study Results}
+
+\begin{frame}[fragile]{Typically Used vs. Observed Uses}
+    \begin{figure}
+        \centering
+        \begin{tikzpicture}
+            \begin{axis}[
+                    height=0.6\textheight,
+                    width=0.8\linewidth,
+                    xbar,
+                    xtick={0,1,2,3,4},
+                    symbolic y coords={Add. Tools,Debugger,Trace Logs},
+                    ytick=data,
+                    enlarge y limits=0.5,
+                    enlargelimits=0.15,
+                    nodes near coords, nodes near coords align={horizontal},
+                    legend pos=south east,
+                    legend style={fill=none,at={(0.5,-0.2)},anchor=north},
+                    legend cell align={left},
+                    reverse legend
+                ]
+
+                \addlegendentry{\# of subjects were observed using}
+                \addplot[
+                        fill=pureminimalistic@text@red,
+                        draw=pureminimalistic@text@red
+                    ]
+                    coordinates {
+                        (0,Add. Tools) (4,Trace Logs) (2,Debugger)
+                    };
+
+                \addlegendentry{\# of subjects typically use}
+                \addplot[
+                        pattern=crosshatch dots,
+                        pattern color=pureminimalistic@text@red,
+                        draw=pureminimalistic@text@red
+                    ]
+                    coordinates {
+                        (2,Add. Tools) (4,Trace Logs) (3,Debugger)
+                    };
+            \end{axis}
+        \end{tikzpicture}
+    \end{figure}
 \end{frame}
 
+\note[itemize] {
+    \item Based on survey and experiment
+    \item Able to compare typically vs observed
+    \item Trace logs: 100\% used and told to typically use
+    \item Debugger: Some small difference, not all use typically
+    \item Interpretation: Already gave up on debugger?
+    \item Largest discrepancy: Add. tools!
+    \item No one used them
+    \item \textbf{Even though survey told they usually do!}
+    \item Even if they had personal dev setup at home
+    \item Therefore, we reason...
+}
+
+\section{Conclusion}
+
+\begin{frame}[fragile]{Interpretation}
+    \begin{vfilleditems}
+        \item If Engineers \textbf{know} about RxJS debugging tools, they \textbf{do not use} them
+        \item \textbf{Manual code modification} is common practice
+        \item \textbf{Imperative} debuggers cannot handle RxJS abstractions
+    \end{vfilleditems}
+\end{frame}
+
+\note[itemize] {
+    \item Knowing about tools seems not enough
+    \item Is there a \textbf{boundary} that keeps engineers from using them?
+    \item Eventually, tendency to trace logs and code extraction
+    \item Repeated theme: Traditional debuggers cannot help in all cases
+    \item In summary, this builds basis for answer to RQ1
+}
+
+\begin{frame}[fragile]{Answer}
+    \begin{enumerate}
+        \item[\color{gray}{RQ1}] \color{gray}{What challenges do software engineers face when debugging RxJS-based applications?}
+        \vfill\item[\Large{``}] \color{white}{The most significant challenge software engineers face [\dots] is to know \textbf{when} they should apply \textbf{what tool} to resolve their current problem in the \textbf{most efficient} way.}
+    \end{enumerate}
+\end{frame}
+
+\note[itemize] {
+    \item Challenge: What tool, when and how
+    \item Often: The "next best" as we saw: Traditional tools
+    \item Often: Modify code
+    \item Less often: Specific tools
+    \item What can we do about it?
+    \item Answer question 2
+}
+
+\begin{frame}[fragile]{Answer}
+    \begin{enumerate}
+        \item[\color{gray}{RQ2}] \color{gray}{How can the experience of software engineers during the debugging process of RxJS-based applications be improved?}
+        \vfill\item[\Large{``}] \color{white}{[\dots] \textbf{improve} the \textbf{experience} of debugging [\dots] by providing RP specific debugging utilities where software engineers expect them the most: \textbf{Fully integrated} [\dots into] their IDE [\dots]}
+    \end{enumerate}
+\end{frame}
+
+\note[itemize] {
+    \item Improve the dev user experience
+    \item Lower the boundary to use the right tool
+    \item At the right time
+    \item Integrate the right tools where they are expected the most
+    \item Example: Async Stack Traces in Google Chrome
+    \item Dont look for it: Its there where it belongs.
+    \item Hence, outlook to future work...
+}
+
 \section{Future Work}
+
 \begin{frame}[fragile]{Future Work}
     \begin{vfilleditems}
-        \item Manuel Alabor \cite{10.1145/2501654.2501666}
-        \item Manuel Alabor
-        \item Manuel Alabor
+        \item \textbf{Propose} and \textbf{implement} solution to \textbf{prevent} manual code \textbf{modification}
+        \item \textbf{Evaluate} proposal on \textbdf{effectiveness} in follow up study
     \end{vfilleditems}
 \end{frame}
+
+\note[itemize] {
+    \item Work on a concrete prototype
+    \item Find solution for manual code modification scenarios
+    \item Without manual intervention required
+    \item Validate effectiveness in another study iteration
+    \item One more thing...
+}
+
+\begin{frame}[fragile]{Sneak Preview}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.6\textheight]{figures/tweet-rxjs-debugging-extension.png}
+        \caption{\tiny{Source: \url{https://twitter.com/swissmanu/status/1313412280072232960}}}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
+    \item Small sneak preview
+    \item Things are progressing
+    \item Thank you for your time and attention!
+}
 
@@
   \vfill
-  {\fontsize{40}{50}\selectfont Q \& A}
+  {\fontsize{40}{50}\selectfont\color{white}{Q \& A}}
   \vfill
@@
 
-\appendix
-\begin{frame}[allowframebreaks]{References}
-	\printbibliography
-\end{frame}
+\note[itemize] {
+    \item Happy for input
+    \item Happy to answer questions
+}
+
+% \appendix
+% \begin{frame}[allowframebreaks]{References}
+% 	\printbibliography
+% \end{frame}
 

### 2020-11-16 slides.tex
Practice Talk Feedback

@@
     \item I am Manuel
-    \item Master Student, OST, Switzerland
-    \item I want to tell you a short story...
-    \item State of debugging for RxJS
-    \item You might think, well:
-    \item ``It has JS in its name''
-    \item JavaScript and Engineering?
+    \item Frontend Engineer for 7 years
+    \item Web and native mobile
+    \item Now: Senior Frontend Engineer, Swiss Fintech
+    \item Now: Master Student, OST, Switzerland
+    \item In my talk about: State of debugging for RxJS
+    \item I want to tell a story
 }
@@
 \note[itemize] {
-    \item Thats frontend stuff!
+    \item JS, frontend, frontend engineer...
     \item Are ``Frontend Engineers'' Engineers?
@@
     \item Wait? Sales pitch?
-    \item I have a story of frustration to tell: Debugging
+    \item I have another story:
+    \item About: Debugging
 }
@@
     \item Helped sometimes (lambda functions, ...)
+}
+
+\begin{frame}[fragile]{Debugging: First Intuition}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=\textheight/2]{figures/stacktrace.png}
+        \caption{Stack Trace}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
     \item Often felt lost though
@@
     \item program step controls: Operator? No: Internals!
-    \item What else?
+    \item What else is possible?
 }
@@
     \item Sprinkle log statements
+}
+
+\begin{frame}[fragile]{Debugging: The Reality}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=\textheight/2]{figures/tracelogs.png}
+        \caption{Flood of Trace Logs}
+    \end{figure}
+\end{frame}
+
+\note[itemize] {
     \item Analyze flood of logs
@@
     \item Looked for existing solutions
-    \item Good concepts for other languages (Scala)
-    \item RxJS less stuff
+}
+
+\begin{frame}[fragile]{A New Hope?}
+    \begin{vfilleditems}
+        \item \textbf{Reactive Inspector} for REScala \cite{10.1145/2577080.2577083}, Salvaneschi et al. \cite{10.1145/2884781.2884815} \cite{10.1145/2889160.2893174}
+        \item \textbf{RxFiddle}, Banken et al. \cite{10.1145/3180155.3180156}
+        \item Better trace logs using \textbf{rxjs-spy} \cite{rxjsspy}
+    \end{vfilleditems}
+\end{frame}
+
+\note[itemize] {
+    \item Own research
+    \item Some interesting findings
+    \item Most elaborate: Reactive Inspector
+    \item Salvaneschi et al
+    \item REScala, Scala
+    \item Some, but not a lot for RxJS
+    \item rxfiddle, banken et al -> Visualizer Externally
+    \item rxjs-spy: Better logs, but still: trace log approach
+    \item MEH!
     \item Lets do this on our own!
@@
     \item Recruiting via Twitter and Email
-    \item Personal Network
     \item 10 participants
@@
 
-\begin{frame}[fragile]{Five One-to-One Interviews}
+\begin{frame}[fragile]{Scope}
+    \begin{columns}[T]
+        \begin{column}{.4\linewidth}
+
             \begin{vfilleditems}
-		\item How do you \textbf{use} Reactive Programming?
-		\item Which aspects do you \textbf{like} on Reactive Programming?
-		\item Which do you \textbf{dislike}?
+                \item Five \textbf{Interviews}\vspace{1em}
+                \item General \textbf{Experience} with Reactive Programming\vspace{0.5em}
+                \item What do you \textbf{like}?\vspace{0.5em}
+                \item What do you \textbf{dislike}?
         	\end{vfilleditems}
+        \end{column}
+        \begin{column}{.4\linewidth}
+            \begin{vfilleditems}
+                \item Five War Story \textbf{Reports}\vspace{1em}
+        		\item Recall a specific \textbf{challenging} debugging situation?\vspace{0.5em}
+        		\item How did you \textbf{solve} it?
+        	\end{vfilleditems}
+        \end{column}
+    \end{columns}
 \end{frame}
@@
 \note[itemize] {
-    \item Interviews had no focus on debugging yet
-    \item Interested in experience with RP
+    \item Experience with RP
     \item Pros
-    \item Cons
-    \item Pain Points
-}
-
-\begin{frame}[fragile]{Five Written War Story Reports}
-	\begin{vfilleditems}
-		\item When did you \textbf{last} have to debug an RxJS-based Application?
-		\item Do you remember a specific \textbf{challenging} debugging situation?
-		\item How did you \textbf{solve} it? (If at all?)
-	\end{vfilleditems}
-\end{frame}
-
-\note[itemize] {
-    \item This time: Focused on debugging
-    \item Hear about specific situations
+    \item Cons / Pain Points
     \item They should recall particular scenario
     \item Step by step
-    \item What problems did they face tool?
-    \item How did they overcome them?
+    \item What problems?
+    \item How did they solve them?
 }
@@
     \item Integrate the right tools where they are expected the most
-    \item Example: Async Stack Traces in Google Chrome
     \item Dont look for it: Its there where it belongs.
@@
 
-% \appendix
-% \begin{frame}[allowframebreaks]{References}
-% 	\printbibliography
-% \end{frame}
+\appendix
+\begin{frame}[allowframebreaks]{References}
+	\printbibliography
+\end{frame}
+
+\begin{frame}[plain, noframenumbering]{}
+    \centering
+    \vfill
+    {\fontsize{30}{40}\selectfont\color{white}{Backup Slides}}
+    \vfill
+\end{frame}
 

### 2020-11-16 slides.tex
Updates from Overleaf

@@
         \begin{column}{.4\linewidth}
-
-            \begin{vfilleditems}
+            \begin{itemize}
                 \item Five \textbf{Interviews}\vspace{1em}
-                \item General \textbf{Experience} with Reactive Programming\vspace{0.5em}
-                \item What do you \textbf{like}?\vspace{0.5em}
+                \begin{itemize}
+                    \item General \textbf{Experience} with Reactive Programming\vspace{0.7em}
+                    \item What do you \textbf{like}?\vspace{0.7em}
                     \item What do you \textbf{dislike}?
-        	\end{vfilleditems}
+                \end{itemize}
+        	\end{itemize}
         \end{column}
         \begin{column}{.4\linewidth}
-            \begin{vfilleditems}
+            \begin{itemize}
                 \item Five War Story \textbf{Reports}\vspace{1em}
-        		\item Recall a specific \textbf{challenging} debugging situation?\vspace{0.5em}
+                \begin{itemize}
+            		\item Recall a specific \textbf{challenging} debugging situation?\vspace{0.7em}
             		\item How did you \textbf{solve} it?
-        	\end{vfilleditems}
+        		\end{itemize}
+        	\end{itemize}
         \end{column}
@@
 \note[itemize] {
-    \item Lengthy answers
     \item Repeating themes
-    \item RP is awesome for composing complex data flows
-    \item That aside
     \item Debugging is hard
@@
     \item Validate effectiveness in another study iteration
-    \item One more thing...
-}
-
-\begin{frame}[fragile]{Sneak Preview}
-    \begin{figure}[H]
-        \centering
-        \includegraphics[height=0.6\textheight]{figures/tweet-rxjs-debugging-extension.png}
-        \caption{\tiny{Source: \url{https://twitter.com/swissmanu/status/1313412280072232960}}}
-    \end{figure}
-\end{frame}
-
-\note[itemize] {
-    \item Small sneak preview
-    \item Things are progressing
-    \item Thank you for your time and attention!
 }
@@
 
+\begin{frame}[fragile]{Future Work PoC: Add Probe}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.6\textheight]{figures/tweet-rxjs-debugging-extension.png}
+        \caption{\tiny{Source: \url{https://twitter.com/swissmanu/status/1313412280072232960}}}
+    \end{figure}
+\end{frame}
+
+\begin{frame}[fragile]{Future Work PoC: Monitor}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.7\textheight]{figures/backup-screenshots/step5-2.png}
+        \caption{Empty Monitor Pane (right)}
+    \end{figure}
+\end{frame}
+
+\begin{frame}[fragile]{Future Work PoC: Output}
+    \begin{figure}[H]
+        \centering
+        \includegraphics[height=0.7\textheight]{figures/backup-screenshots/step8.png}
+        \caption{Monitor Pane with Output (right)}
+    \end{figure}
+\end{frame}
+
 \end{document}
